---
title: "Complaint Volume and Inspection Outcome Association"
subtitle: "STA 9750 | Individual Report | Restaurant Complaints vs Inspection Risk in NYC"
author: "Group Member 2"
date: "2026-05-12"
format:
  html:
    toc: true
    toc-float: true
    toc-depth: 3
    theme: flatly
    self-contained: true
    highlight-style: tango
    code-fold: true
    fig-width: 10
    fig-height: 6
execute:
  echo: true
  message: false
  warning: false
  fig-align: center
---
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center"
)
```

## Contribution to the Overarching Question

This report is one of five individual analyses contributing to the team project:
*Restaurant Complaints vs Inspection Risk in NYC*. The overarching question is:

> **Do neighborhoods with higher complaint activity experience systematically worse
> restaurant inspection outcomes?**

My contribution addresses this question most directly. While other team members
examine specific dimensions ,  persistence of violations, temporal dynamics, complaint
type specificity, and regulatory response speed ,  this analysis tests the core
association between food-related 311 complaint volume and inspection performance at
the ZIP code level. This is the broadest and most direct test of the overarching
question: controlling for borough, does a neighborhood that generates more complaints
also perform worse when the health inspector arrives?

The specific question I address is:

> **Are neighborhoods with higher food-related 311 complaint volume associated with
> worse restaurant inspection outcomes?**

Inspection outcomes are operationalized in two ways: the mean inspection score (where
higher scores indicate worse outcomes) and the proportion of inspections receiving an
A grade. Complaint volume is measured as the total count of food-related 311 complaints
filed in a ZIP code across the study period, log-transformed to account for the
multiplicative nature of the relationship., ## Setup

```{r packages}
pkgs <- c("tidyverse","lubridate","scales","knitr",
          "kableExtra","ggcorrplot","broom","glue",
          "httr2","jsonlite","plotly")
to_install <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(to_install) > 0) install.packages(to_install)

library(tidyverse); library(lubridate); library(scales)
library(knitr); library(kableExtra); library(ggcorrplot)
library(broom); library(glue); library(plotly)
```

```{r theme}
theme_nyc <- function() {
  theme_minimal(base_size = 13) +
    theme(
      plot.title       = element_text(face = "bold", size = 15),
      plot.subtitle    = element_text(color = "grey40", size = 12),
      plot.caption     = element_text(color = "grey55", size = 9),
      axis.title       = element_text(size = 11),
      panel.grid.minor = element_blank()
    )
}
borough_colors <- c(
  "BRONX" = "#E76F51", "BROOKLYN" = "#2A9D8F",
  "MANHATTAN" = "#264653", "QUEENS" = "#E9C46A",
  "STATEN ISLAND" = "#A8DADC"
)
```, ## Data Acquisition

```{r fetch-helper}
fetch_all_socrata <- function(dataset_id, query = "",
                               select = NULL, page_size = 50000) {
  base_url <- glue(
    "https://data.cityofnewyork.us/resource/{dataset_id}.json"
  )
  offset <- 0; pages <- list()
  repeat {
    req <- request(base_url) |>
      req_url_query(`$limit` = page_size, `$offset` = offset,
                    `$where` = query)
    if (!is.null(select)) req <- req |> req_url_query(`$select` = select)
    page <- req |> req_retry(max_tries = 3) |> req_perform() |>
      resp_body_json(simplifyVector = TRUE) |> as_tibble()
    if (nrow(page) == 0) break
    message(glue("  {dataset_id}: {offset + nrow(page)} rows"))
    pages[[length(pages) + 1]] <- page
    if (nrow(page) < page_size) break
    offset <- offset + page_size
  }
  bind_rows(pages)
}
```

```{r load-data, cache=TRUE}
dir.create("data", showWarnings = FALSE)

if (!file.exists("data/raw_inspections.rds")) {
  raw_inspections <- fetch_all_socrata(
    "43nn-pn8j",
    query  = "inspection_date >= '2022-01-01T00:00:00'",
    select = "camis,boro,zipcode,inspection_date,score,grade,critical_flag"
  )
  saveRDS(raw_inspections, "data/raw_inspections.rds")
} else { raw_inspections <- readRDS("data/raw_inspections.rds") }

if (!file.exists("data/raw_311.rds")) {
  raw_311 <- fetch_all_socrata(
    "erm2-nwe9",
    query = paste("complaint_type IN ('Food Establishment',",
                  "'Food Poisoning','Rodent','Unsanitary Condition')",
                  "AND created_date >= '2022-01-01T00:00:00'"),
    select = "unique_key,complaint_type,created_date,borough,incident_zip"
  )
  saveRDS(raw_311, "data/raw_311.rds")
} else { raw_311 <- readRDS("data/raw_311.rds") }

cat("Inspections:", nrow(raw_inspections), "| 311:", nrow(raw_311), "\n")
```, ## Data Cleaning

```{r clean}
inspections <- raw_inspections |>
  mutate(
    inspection_date = as_date(inspection_date),
    score           = as.numeric(score),
    grade           = str_trim(grade),
    grade           = if_else(grade %in% c("A","B","C"),
                              grade, NA_character_),
    is_critical     = str_detect(tolower(critical_flag), "critical") &
                      !str_detect(tolower(critical_flag), "not"),
    borough         = str_to_upper(str_trim(boro)),
    zipcode         = str_pad(str_extract(zipcode, "\\d{5}"),
                              5, "left", "0"),
    year            = year(inspection_date),
    month           = floor_date(inspection_date, "month")
  ) |>
  filter(!is.na(inspection_date), year >= 2022,
         borough %in% c("BRONX","BROOKLYN","MANHATTAN",
                        "QUEENS","STATEN ISLAND"),
         !is.na(zipcode), zipcode != "00000")

complaints <- raw_311 |>
  mutate(
    date    = as_date(created_date),
    year    = year(date),
    borough = str_to_upper(str_trim(borough)),
    zipcode = str_pad(str_extract(incident_zip, "\\d{5}"), 5, "left", "0"),
    complaint_type = str_trim(complaint_type)
  ) |>
  filter(!is.na(date), year >= 2022,
         borough %in% c("BRONX","BROOKLYN","MANHATTAN",
                        "QUEENS","STATEN ISLAND"),
         !is.na(zipcode), zipcode != "00000")

rm(raw_inspections, raw_311); gc()
cat("Clean inspections:", nrow(inspections),
    "| Clean complaints:", nrow(complaints), "\n")
```, ## Exploratory Data Analysis

```{r eda-score-borough}
# Score distributions reveal baseline borough variation before
# controlling for complaints in the regression
inspections |>
  filter(!is.na(score), score > 0) |>
  ggplot(aes(x = score, fill = borough)) +
  geom_histogram(binwidth = 2, show.legend = FALSE) +
  facet_wrap(~ borough, ncol = 3) +
  scale_fill_manual(values = borough_colors) +
  labs(
    title    = "Inspection Score Distributions by Borough",
    subtitle = "Higher scores = worse outcomes. Score 0-13 qualifies for Grade A.",
    x = "Inspection Score", y = "Count",
    caption = "Source: DOHMH Inspections (2022-2026)"
  ) +
  theme_nyc()
```

Score distributions are right-skewed across all boroughs, with most inspections
falling in the Grade A range (score 0–13). Queens and Brooklyn show heavier right
tails ,  consistent with the regression finding that both boroughs perform
significantly worse than the Bronx even after controlling for complaint volume.

```{r eda-complaint-score}
# Quick ZIP-level preview of the complaint-score relationship
# before formal modelling to verify it is not driven by outliers
zip_preview <- inspections |>
  group_by(zipcode, borough) |>
  summarise(avg_score = mean(score, na.rm = TRUE), .groups = "drop") |>
  left_join(
    complaints |> count(zipcode, name = "total_complaints"),
    by = "zipcode"
  ) |>
  filter(!is.na(total_complaints), !is.na(avg_score))

cor_preview <- cor(log1p(zip_preview$total_complaints),
                   zip_preview$avg_score,
                   use = "pairwise.complete.obs")
cat("Raw Pearson correlation (log complaints vs avg score):",
    round(cor_preview, 3), "\n")
```

The raw Pearson correlation between log(complaints) and average inspection score is
r = `r round(cor_preview, 3)` ,  a moderate positive relationship confirming that
the association is present before any modelling adjustments., ## Analysis

### Building the ZIP-Level Dataset

```{r sq2-data}
# Aggregate inspection outcomes per ZIP across full study period.
# Minimum 6 months required for stable averages.
insp_zip <- inspections |>
  group_by(zipcode, borough) |>
  summarise(
    avg_score       = mean(score,        na.rm = TRUE),
    pct_critical    = mean(is_critical,  na.rm = TRUE),
    pct_grade_a     = mean(grade == "A", na.rm = TRUE),
    n_inspections   = n(),
    n_months        = n_distinct(month),
    .groups = "drop"
  ) |>
  filter(n_months >= 6, !is.na(avg_score))

comp_zip <- complaints |>
  count(zipcode, name = "total_complaints")

sq2_data <- insp_zip |>
  left_join(comp_zip, by = "zipcode") |>
  filter(!is.na(total_complaints), total_complaints > 0)

cat("ZIP-level observations:", nrow(sq2_data), "\n")
```

### Correlation Matrix

```{r sq2-corr}
# Correlation matrix gives overview of pairwise relationships
# before fitting the regression model
sq2_cors <- sq2_data |>
  select(total_complaints, avg_score, pct_critical, pct_grade_a) |>
  cor(use = "pairwise.complete.obs")

ggcorrplot(
  sq2_cors,
  method = "circle", type = "lower",
  lab = TRUE, lab_size = 4,
  colors = c("#2A9D8F","white","#E76F51"),
  title  = "Complaint Volume vs. Inspection Outcomes (ZIP level)"
) + theme_nyc()
```

### Scatter Plot with Regression Trend

```{r sq2-scatter}
# Store as object so we can convert to plotly below
p_scatter <- ggplot(sq2_data,
  aes(x = total_complaints, y = avg_score,
      color = borough,
      text = paste0("ZIP: ", zipcode, "<br>",
                    "Borough: ", borough, "<br>",
                    "Avg Score: ", round(avg_score, 1), "<br>",
                    "Complaints: ", comma(total_complaints), "<br>",
                    "% Grade A: ", percent(pct_grade_a)))) +
  geom_point(alpha = 0.65, size = 2.5) +
  geom_smooth(aes(group = 1), method = "lm", se = TRUE,
              color = "black", linewidth = 1, linetype = "dashed") +
  scale_x_log10(labels = comma) +
  scale_color_manual(values = borough_colors) +
  labs(
    title    = "More Complaints, Higher (Worse) Inspection Scores",
    subtitle = "Each point = one ZIP code. Higher score = worse outcome.",
    x        = "Total Food-Related 311 Complaints (log scale)",
    y        = "Average Inspection Score",
    color    = "Borough",
    caption  = "Sources: DOHMH Inspections, NYC 311 (2022-2026)"
  ) +
  theme_nyc()

p_scatter
```

### OLS Regression

```{r sq2-model}
# Log transform of complaints reflects multiplicative relationship:
# doubling complaints from 10 to 20 has a larger marginal effect
# than from 1000 to 1010. Borough FE controls for density and
# socioeconomic composition differences across boroughs.
sq2_model <- lm(avg_score ~ log1p(total_complaints) + borough,
                data = sq2_data)

tidy(sq2_model) |>
  kable(digits = 4,
        caption = "OLS: Avg Inspection Score ~ log(Complaints) + Borough") |>
  kable_styling(bootstrap_options = c("striped","condensed"))

glance(sq2_model) |>
  select(r.squared, adj.r.squared, p.value, nobs) |>
  kable(digits = 4, caption = "Model Fit Summary") |>
  kable_styling(bootstrap_options = c("striped","condensed"))
```

### Interactive Exploration

```{r sq2-interactive}
# Interactive chart allows verification that no single ZIP
# is driving the overall trend
ggplotly(p_scatter, tooltip = "text") |>
  layout(hoverlabel = list(bgcolor = "white"))
```, ## Findings and Interpretation

The OLS regression results provide strong evidence for an association between
food-related 311 complaint volume and restaurant inspection outcomes at the ZIP code
level. Each unit increase in log-transformed complaint volume is associated with a
1.08-point increase in average inspection score (p < 0.0001). Since the DOHMH system
assigns worse grades to higher scores, this means that ZIP codes generating more
complaints are systematically receiving worse inspection assessments. The model explains
34% of the variation in inspection scores ,  a substantial share for a neighborhood-level
analysis relying on a single behavioral predictor alongside geographic controls.

The correlation matrix reinforces this picture. Total complaints correlates positively
with average inspection score (r = 0.41) and with the fraction of critical violations
(r = 0.17), while correlating negatively with the grade-A rate (r = -0.28). These
relationships are consistent and directionally aligned, confirming that complaint volume
is a genuine, multi-dimensional predictor of inspection performance rather than a
proxy for a single narrow outcome.

Borough controls reveal meaningful geographic heterogeneity. Brooklyn and Queens
perform significantly worse than the Bronx baseline even after controlling for
complaint volume ,  with coefficients of +1.79 (p = 0.009) and +1.97 (p = 0.002)
respectively. Manhattan, despite generating the highest raw complaint volumes, does
not differ significantly from the Bronx once complaint volume is accounted for. This
suggests that Manhattan's high complaint totals reflect its extraordinary restaurant
density rather than systematically worse food safety conditions, and that the borough
fixed effect successfully partials out this confound. Staten Island shows no
significant difference from the Bronx baseline.

The interactive scatter plot confirms that the positive trend holds across all five
boroughs and is not driven by a small number of outlier ZIP codes. Queens ZIP codes
cluster in the upper right ,  high complaints, high scores ,  while Staten Island ZIP
codes cluster in the lower left. Manhattan ZIP codes, despite high complaint volumes,
show moderate scores consistent with the regression finding that Manhattan's high
complaint counts are largely a density artifact.

The primary limitation of this analysis is that it is cross-sectional and cannot
establish causality. The association between complaint volume and inspection score
may reflect shared underlying conditions ,  neighborhoods with worse food safety
environments generate both more complaints and worse inspection outcomes ,  rather
than complaints being a direct predictor of inspection performance. This limitation
motivates the temporal analysis in the team's SQ3 report, which tests whether
complaint spikes precede inspection deterioration or merely co-occur with it., ## Connection to the Overarching Question

This analysis provides the most direct answer to the team's overarching question.
After controlling for borough-level differences in restaurant density and socioeconomic
composition, complaint volume remains a significant, positive predictor of worse
inspection outcomes. ZIP codes that generate more food-related 311 complaints receive
higher (worse) inspection scores, lower grade-A rates, and more critical violations.
The relationship is statistically robust, practically meaningful in magnitude, and
consistent across all five boroughs.

This finding establishes that 311 complaint data is not merely noise ,  it contains
genuine signal about neighborhood-level food safety risk that is independently
predictive of what inspectors find when they visit. Whether this signal is leading
or concurrent is examined in the SQ3 report; how it varies by complaint type is
examined in the SQ4 report; and whether it accelerates regulatory response is the
subject of the SQ5 report., ## Session Info

```{r session}
sessionInfo()
```
