---
title: "Complaint Type Specificity and Critical Violation Rates"
subtitle: "STA 9750 | Individual Report | Restaurant Complaints vs Inspection Risk in NYC"
author: "Group Member 4"
date: "2026-05-12"
format:
  html:
    toc: true
    toc-float: true
    toc-depth: 3
    theme: flatly
    self-contained: true
    highlight-style: tango
    code-fold: true
    fig-width: 10
    fig-height: 6
execute:
  echo: true
  message: false
  warning: false
  fig-align: center
---
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center"
)
```

## Contribution to the Overarching Question

This report is one of five individual analyses contributing to the team project:
*Restaurant Complaints vs Inspection Risk in NYC*. The overarching question is:

> **Do neighborhoods with higher complaint activity experience systematically worse
> restaurant inspection outcomes?**

The team's SQ1 and SQ2 analyses treat all food-related 311 complaints as a single
aggregate signal. My contribution disaggregates that signal to ask a more precise
question: are all complaint types equally informative, or do specific kinds of
citizen concern carry stronger predictive power for food safety outcomes? This matters
for both the interpretation of the overarching question and for its practical application
,  if certain complaint types are stronger predictors of critical violations, regulators
could prioritize inspection responses to those specific complaint categories rather than
treating all food-related 311 activity as equivalent.

The specific question I address is:

> **Which types of food-related complaints are most strongly associated with
> higher neighborhood-level critical violation rates?**

The four complaint types available in the dataset are: Rodent, Food Establishment,
Food Poisoning, and Unsanitary Condition. My hypothesis entering the analysis is that
Rodent complaints will show the strongest association with critical violations, since
rodent presence is a direct sanitation failure and one of the most heavily weighted
violation categories in the DOHMH inspection rubric. I test this hypothesis and
examine what the results reveal about the nature of the complaint-inspection
relationship across complaint categories., ## Setup

```{r packages}
pkgs <- c("tidyverse","lubridate","scales","knitr",
          "kableExtra","broom","glue","httr2","jsonlite")
to_install <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(to_install) > 0) install.packages(to_install)
library(tidyverse); library(lubridate); library(scales)
library(knitr); library(kableExtra); library(broom); library(glue)
```

```{r theme}
theme_nyc <- function() {
  theme_minimal(base_size = 13) +
    theme(
      plot.title    = element_text(face = "bold", size = 15),
      plot.subtitle = element_text(color = "grey40", size = 12),
      plot.caption  = element_text(color = "grey55", size = 9),
      axis.title    = element_text(size = 11),
      panel.grid.minor = element_blank()
    )
}
```, ## Data Acquisition

```{r fetch-helper}
fetch_all_socrata <- function(dataset_id, query = "",
                               select = NULL, page_size = 50000) {
  base_url <- glue(
    "https://data.cityofnewyork.us/resource/{dataset_id}.json"
  )
  offset <- 0; pages <- list()
  repeat {
    req <- request(base_url) |>
      req_url_query(`$limit` = page_size, `$offset` = offset,
                    `$where` = query)
    if (!is.null(select)) req <- req |> req_url_query(`$select` = select)
    page <- req |> req_retry(max_tries = 3) |> req_perform() |>
      resp_body_json(simplifyVector = TRUE) |> as_tibble()
    if (nrow(page) == 0) break
    message(glue("  {dataset_id}: {offset + nrow(page)} rows"))
    pages[[length(pages) + 1]] <- page
    if (nrow(page) < page_size) break
    offset <- offset + page_size
  }
  bind_rows(pages)
}
```

```{r load-data, cache=TRUE}
dir.create("data", showWarnings = FALSE)
if (!file.exists("data/raw_inspections.rds")) {
  raw_inspections <- fetch_all_socrata(
    "43nn-pn8j",
    query = "inspection_date >= '2022-01-01T00:00:00'",
    select = "camis,boro,zipcode,inspection_date,score,grade,critical_flag"
  )
  saveRDS(raw_inspections, "data/raw_inspections.rds")
} else { raw_inspections <- readRDS("data/raw_inspections.rds") }

if (!file.exists("data/raw_311.rds")) {
  raw_311 <- fetch_all_socrata(
    "erm2-nwe9",
    query = paste("complaint_type IN ('Food Establishment',",
                  "'Food Poisoning','Rodent','Unsanitary Condition')",
                  "AND created_date >= '2022-01-01T00:00:00'"),
    select = "unique_key,complaint_type,created_date,borough,incident_zip"
  )
  saveRDS(raw_311, "data/raw_311.rds")
} else { raw_311 <- readRDS("data/raw_311.rds") }
```, ## Data Cleaning

```{r clean}
inspections <- raw_inspections |>
  mutate(
    inspection_date = as_date(inspection_date),
    score           = as.numeric(score),
    grade           = str_trim(grade),
    grade           = if_else(grade %in% c("A","B","C"),
                              grade, NA_character_),
    is_critical     = str_detect(tolower(critical_flag),"critical") &
                      !str_detect(tolower(critical_flag),"not"),
    borough         = str_to_upper(str_trim(boro)),
    zipcode         = str_pad(str_extract(zipcode,"\\d{5}"),5,"left","0"),
    year            = year(inspection_date)
  ) |>
  filter(!is.na(inspection_date), year >= 2022,
         borough %in% c("BRONX","BROOKLYN","MANHATTAN",
                        "QUEENS","STATEN ISLAND"),
         !is.na(zipcode), zipcode != "00000")

complaints <- raw_311 |>
  mutate(
    date    = as_date(created_date),
    year    = year(date),
    borough = str_to_upper(str_trim(borough)),
    zipcode = str_pad(str_extract(incident_zip,"\\d{5}"),5,"left","0"),
    complaint_type = str_trim(complaint_type)
  ) |>
  filter(!is.na(date), year >= 2022,
         borough %in% c("BRONX","BROOKLYN","MANHATTAN",
                        "QUEENS","STATEN ISLAND"),
         !is.na(zipcode), zipcode != "00000")

rm(raw_inspections, raw_311); gc()
```, ## Exploratory Data Analysis

```{r eda-complaint-volumes}
# Understand the relative volume of each complaint type before
# computing correlations ,  volume differences affect interpretability
complaints |>
  count(complaint_type, sort = TRUE) |>
  mutate(
    pct = n / sum(n),
    complaint_type = fct_reorder(complaint_type, n)
  ) |>
  ggplot(aes(x = n, y = complaint_type, fill = complaint_type)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(comma(n)," (",percent(pct,accuracy=0.1),")")),
            hjust = -0.05, size = 3.5) +
  scale_x_continuous(labels = comma,
                     expand = expansion(mult = c(0, 0.3))) +
  scale_fill_manual(values = c(
    "Rodent"               = "#E76F51",
    "Food Establishment"   = "#264653",
    "Food Poisoning"       = "#2A9D8F",
    "Unsanitary Condition" = "#E9C46A"
  )) +
  labs(
    title    = "Food-Related 311 Complaint Volumes by Type",
    subtitle = "Rodent complaints dominate, comprising over 70% of total volume",
    x = "Total Complaints (2022-2026)", y = NULL,
    caption = "Source: NYC 311 (2022-2026)"
  ) +
  theme_nyc()
```

Rodent complaints account for over 70% of all food-related 311 activity in the
dataset, making it the dominant signal by volume. Food Establishment complaints
represent roughly 22%, Food Poisoning about 6%, and Unsanitary Condition less than
1%. This volume imbalance is important context for interpreting the correlation
results: a complaint type with very few observations (Unsanitary Condition) will
have noisier correlation estimates than one with many thousands.

```{r eda-borough-type}
# How does complaint type composition vary across boroughs?
complaints |>
  count(borough, complaint_type) |>
  group_by(borough) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(x = borough, y = pct, fill = complaint_type)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = percent) +
  scale_fill_manual(values = c(
    "Rodent"               = "#E76F51",
    "Food Establishment"   = "#264653",
    "Food Poisoning"       = "#2A9D8F",
    "Unsanitary Condition" = "#E9C46A"
  )) +
  labs(
    title    = "Complaint Type Composition by Borough",
    subtitle = "Composition is broadly consistent across boroughs",
    x = NULL, y = "Share of Complaints", fill = "Complaint Type",
    caption = "Source: NYC 311 (2022-2026)"
  ) +
  theme_nyc()
```, ## Analysis

### Building the ZIP-Level Dataset

```{r sq4-data}
# ZIP-level complaint counts per type ,  wide format for correlation analysis
sq4_complaints <- complaints |>
  count(zipcode, complaint_type) |>
  pivot_wider(names_from = complaint_type, values_from = n,
              values_fill = 0)

# ZIP-level inspection outcomes ,  require >= 10 inspections for stable estimates
sq4_inspections <- inspections |>
  group_by(zipcode) |>
  summarise(
    n_inspections = n_distinct(paste(camis, inspection_date)),
    pct_critical  = mean(is_critical, na.rm = TRUE),
    mean_score    = mean(score,        na.rm = TRUE),
    pct_grade_a   = mean(grade == "A", na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(n_inspections >= 10)

sq4_data <- sq4_inspections |>
  left_join(sq4_complaints, by = "zipcode")

complaint_cols <- intersect(
  names(sq4_data),
  unique(complaints$complaint_type)
)

cat("ZIP observations:", nrow(sq4_data), "\n")
cat("Complaint types available:", paste(complaint_cols, collapse = ", "), "\n")
```

### Correlation Analysis

```{r sq4-cors}
# Compute Pearson correlation independently for each complaint type
# against three inspection outcome measures. Minimum 10 ZIPs with
# non-zero values required for a stable correlation estimate.
sq4_cors <- map_dfr(complaint_cols, function(ct) {
  vals <- sq4_data[[ct]]
  if (sum(!is.na(vals) & vals > 0) < 10) return(NULL)
  tibble(
    complaint_type   = ct,
    cor_pct_critical = cor(vals, sq4_data$pct_critical,
                           use = "pairwise.complete.obs"),
    cor_mean_score   = cor(vals, sq4_data$mean_score,
                           use = "pairwise.complete.obs"),
    cor_pct_grade_a  = cor(vals, sq4_data$pct_grade_a,
                           use = "pairwise.complete.obs"),
    n_zips           = sum(!is.na(vals) & vals > 0)
  )
}) |>
  filter(!is.na(cor_pct_critical)) |>
  arrange(desc(abs(cor_pct_critical)))

sq4_cors |>
  kable(
    digits    = 4,
    caption   = "Complaint Type Correlations with Inspection Outcomes (ZIP level)",
    col.names = c("Complaint Type","Cor w/ % Critical",
                  "Cor w/ Mean Score","Cor w/ % Grade A","ZIP Count")
  ) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"))
```

```{r sq4-plot}
sq4_cors |>
  mutate(complaint_type = fct_reorder(complaint_type, cor_pct_critical),
         direction = if_else(cor_pct_critical >= 0,"Positive","Negative")) |>
  ggplot(aes(x = cor_pct_critical, y = complaint_type, fill = direction)) +
  geom_col(show.legend = FALSE, width = 0.6) +
  geom_vline(xintercept = 0, linewidth = 0.9) +
  geom_text(aes(label = round(cor_pct_critical, 3),
                hjust = if_else(cor_pct_critical >= 0, -0.2, 1.2)),
            size = 4, fontface = "bold") +
  scale_fill_manual(values = c("Positive" = "#E76F51",
                               "Negative" = "#2A9D8F")) +
  scale_x_continuous(expand = expansion(mult = 0.25)) +
  labs(
    title    = "Which Complaint Types Best Predict Critical Violations?",
    subtitle = "Pearson correlation with % critical violations at ZIP level",
    x = "Correlation with % Critical Violations", y = NULL,
    caption = "Sources: DOHMH Inspections, NYC 311 (2022-2026)"
  ) +
  theme_nyc()
```

### Scatter Plots by Complaint Type

```{r sq4-scatter-grid}
# Visualize the raw ZIP-level scatter for each complaint type
# to verify correlations are not driven by outliers
sq4_long <- sq4_data |>
  select(zipcode, pct_critical, all_of(complaint_cols)) |>
  pivot_longer(all_of(complaint_cols),
               names_to = "complaint_type", values_to = "count") |>
  filter(count > 0)

ggplot(sq4_long, aes(x = count, y = pct_critical)) +
  geom_point(alpha = 0.4, color = "#264653", size = 1.8) +
  geom_smooth(method = "lm", se = TRUE, color = "#E76F51",
              linewidth = 1) +
  facet_wrap(~ complaint_type, scales = "free_x") +
  scale_x_continuous(labels = comma) +
  labs(
    title    = "ZIP-Level Scatter: Each Complaint Type vs % Critical Violations",
    subtitle = "All four types show positive trends; slopes differ by type",
    x = "Complaint Count", y = "% Critical Violations",
    caption = "Sources: DOHMH Inspections, NYC 311 (2022-2026)"
  ) +
  theme_nyc()
```, ## Findings and Interpretation

The correlation results produce a finding that challenges the initial hypothesis.
Food Establishment complaints ,  not Rodent complaints ,  show the strongest correlation
with critical violation rates (r = 0.157), followed by Rodent (r = 0.116), Food
Poisoning (r = 0.109), and Unsanitary Condition (r = 0.047). Crucially, all four
complaint types show positive correlations with both critical violation rates and mean
inspection scores, and negative correlations with grade-A rates. The directional
consistency across all outcome measures strengthens confidence that these are genuine
associations rather than statistical artifacts.

The stronger performance of Food Establishment complaints over Rodent complaints is
interpretable on reflection. The Food Establishment category is a broad catch-all that
captures a wide range of concerns ,  food handling practices, visible cleanliness,
storage conditions, pest evidence, and general restaurant hygiene ,  whereas Rodent
complaints are specific to one type of sanitation failure. This breadth may give Food
Establishment complaints greater predictive power because they capture a more
comprehensive signal about overall restaurant quality. Rodent complaints, while directly
tied to a major DOHMH violation category, reflect only one dimension of food safety
risk. Food Poisoning complaints perform similarly to Rodent (r = 0.109), suggesting
that health outcomes reported by consumers carry meaningful signal despite the
under-reporting inherent in attributing illness to a specific establishment. Unsanitary
Condition complaints show the weakest association, likely because this category
captures a heterogeneous mix of concerns ,  including street cleanliness and non-food
establishments ,  that dilutes its restaurant-specific signal.

The scatter plots by complaint type confirm that these correlations are not driven
by extreme outliers. The positive slope holds broadly across the range of complaint
counts for each type, though with considerable within-type variation that is
consistent with the modest correlation magnitudes.

One important limitation is that the Pearson correlations computed here are bivariate
and do not control for other factors. Higher Food Establishment complaint volume in a
ZIP may be correlated with restaurant density, which in turn affects inspection
frequency and outcome distributions. The team's SQ2 analysis addresses this confounding
more rigorously through OLS regression with borough fixed effects, but type-specific
controlled regressions remain a natural extension of this work., ## Connection to the Overarching Question

This analysis adds precision to the team's answer to the overarching question. Not
all complaints are equal: Food Establishment complaints carry the strongest signal
about neighborhood-level food safety risk, followed by Rodent, Food Poisoning, and
Unsanitary Condition. This finding has direct policy implications ,  inspection
prioritization algorithms that weight complaint types differentially could achieve
better targeting of enforcement resources than treating all food-related 311 activity
as equivalent. The finding also refines the theoretical interpretation: the
complaint-inspection outcome relationship is not simply about the volume of citizen
concern but about the specific nature of what citizens are reporting., ## Session Info

```{r session}
sessionInfo()
```
