---
title: "Neighborhood Complaint Persistence and Repeat Critical Violations"
subtitle: "STA 9750 | Individual Report | Restaurant Complaints vs Inspection Risk in NYC"
author: "Group Member 1"
date: "2026-05-12"
format:
  html:
    toc: true
    toc-float: true
    toc-depth: 3
    theme: flatly
    highlight-style: tango
    code-fold: true
    fig-width: 10
    fig-height: 6
    self-contained: true
execute:
  echo: true
  message: false
  warning: false
  fig-align: center
---

## Overarching Question and My Contribution

This report is one of five individual analyses contributing to the team project
*Restaurant Complaints vs Inspection Risk in NYC*. The overarching question guiding
the entire project is whether neighborhoods with higher complaint activity experience
systematically worse restaurant inspection outcomes. My specific contribution
addresses one particular dimension of that question: persistence. Rather than asking
whether complaints and bad inspection outcomes happen to coincide at a single point
in time, I ask whether neighborhoods that generate sustained, consistently high
volumes of food-related 311 complaints are also the neighborhoods that accumulate the
most entrenched, recurring patterns of inspection failure.

This framing matters because a single critical violation at a single restaurant on a
single inspection date tells us relatively little. Restaurants can fail an inspection
for many reasons, including a first-time procedural lapse, an inspector who happens
to be more stringent, or a temporary issue that is quickly corrected. What is far
more informative is whether the same restaurant keeps getting cited for the same
violation across multiple inspection cycles despite the regulatory pressure to correct
it. That pattern of persistent non-compliance is what I call a repeat critical
violation, and it is the outcome I use to test whether high-complaint neighborhoods
are genuinely facing more severe and entrenched food safety challenges.

My specific question is whether neighborhoods with persistently high complaint volume
have higher repeat critical violations. A repeat critical violation is operationalized
here as the same restaurant being cited for the same violation code on two or more
separate inspection dates within the study period. This definition deliberately
excludes one-time failures and targets the subset of violations that signal a
structural problem rather than a correctable lapse. If complaint volume is a genuine
indicator of neighborhood food safety risk rather than simply a proxy for resident
engagement patterns, we would expect to see the most complaint-heavy ZIP codes also
producing the highest concentrations of these stubborn, recurring violations.

This question connects directly to the overarching project question because it tests
not just whether complaints and outcomes are associated but whether that association
reflects deep, persistent conditions rather than surface-level variation. Finding that
high-complaint neighborhoods accumulate more repeat violations would strongly suggest
that the 311 system is picking up on real, durable food safety problems rather than
transient events or reporting artifacts.

---

## Setup

```{r packages}
pkgs <- c("tidyverse", "lubridate", "scales", "knitr",
          "kableExtra", "broom", "glue", "httr2", "jsonlite")
to_install <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(to_install) > 0) install.packages(to_install)

library(tidyverse)
library(lubridate)
library(scales)
library(knitr)
library(kableExtra)
library(broom)
library(glue)
library(httr2)
library(jsonlite)
```

```{r theme}
theme_nyc <- function() {
  theme_minimal(base_size = 13) +
    theme(
      plot.title       = element_text(face = "bold", size = 15),
      plot.subtitle    = element_text(color = "grey40", size = 12),
      plot.caption     = element_text(color = "grey55", size = 9),
      axis.title       = element_text(size = 11),
      panel.grid.minor = element_blank(),
      strip.text       = element_text(face = "bold")
    )
}
borough_colors <- c(
  "BRONX" = "#E76F51", "BROOKLYN" = "#2A9D8F",
  "MANHATTAN" = "#264653", "QUEENS" = "#E9C46A",
  "STATEN ISLAND" = "#A8DADC"
)
```

---

## Data Acquisition

```{r fetch-helper}
# Paginated Socrata API fetcher. Handles datasets exceeding the 50k row
# single-request limit through offset-based pagination. The select parameter
# reduces payload size by requesting only columns needed for this analysis.
fetch_all_socrata <- function(dataset_id, query = "",
                               select = NULL, page_size = 50000) {
  base_url <- glue(
    "https://data.cityofnewyork.us/resource/{dataset_id}.json"
  )
  offset <- 0
  pages  <- list()
  repeat {
    req <- request(base_url) |>
      req_url_query(`$limit` = page_size, `$offset` = offset,
                    `$where` = query)
    if (!is.null(select)) req <- req |> req_url_query(`$select` = select)
    page <- req |> req_retry(max_tries = 3) |> req_perform() |>
      resp_body_json(simplifyVector = TRUE) |> as_tibble()
    if (nrow(page) == 0) break
    message(glue("  {dataset_id}: {offset + nrow(page)} rows fetched"))
    pages[[length(pages) + 1]] <- page
    if (nrow(page) < page_size) break
    offset <- offset + page_size
  }
  bind_rows(pages)
}
```

```{r load-data}
#| cache: true
# Data is cached locally on first run to avoid re-downloading on every render.
# The if/else pattern ensures the report is fully reproducible from scratch
# on any machine with an internet connection, while remaining fast on re-renders.
dir.create("data", showWarnings = FALSE)

if (!file.exists("data/raw_inspections.rds")) {
  raw_inspections <- fetch_all_socrata(
    dataset_id = "43nn-pn8j",
    query  = "inspection_date >= '2022-01-01T00:00:00'",
    select = paste("camis,dba,boro,zipcode,inspection_date,score,grade,",
                   "violation_code,critical_flag,action,inspection_type",
                   sep = "")
  )
  saveRDS(raw_inspections, "data/raw_inspections.rds")
} else { raw_inspections <- readRDS("data/raw_inspections.rds") }

if (!file.exists("data/raw_311.rds")) {
  raw_311 <- fetch_all_socrata(
    dataset_id = "erm2-nwe9",
    query  = paste("complaint_type IN ('Food Establishment','Food Poisoning',",
                   "'Rodent','Unsanitary Condition')",
                   "AND created_date >= '2022-01-01T00:00:00'"),
    select = "unique_key,complaint_type,created_date,borough,incident_zip"
  )
  saveRDS(raw_311, "data/raw_311.rds")
} else { raw_311 <- readRDS("data/raw_311.rds") }

cat("Inspections rows:", nrow(raw_inspections), "\n")
cat("311 rows:", nrow(raw_311), "\n")
```

---

## Data Cleaning and Preparation

```{r clean-inspections}
# The critical_flag field contains text values including "Critical",
# "Not Critical", and blank entries. The is_critical flag is derived by
# checking for the presence of "critical" while explicitly excluding
# entries containing "not", which would otherwise be captured by a simple
# string match. Grades are restricted to A, B, and C because other values
# such as Z (grade pending) and P (pre-grading) indicate the restaurant
# is mid-cycle and does not yet have a finalized assessment.
inspections <- raw_inspections |>
  mutate(
    inspection_date = as_date(inspection_date),
    score           = as.numeric(score),
    grade           = str_trim(grade),
    grade           = if_else(grade %in% c("A","B","C"),
                              grade, NA_character_),
    is_critical     = str_detect(tolower(critical_flag), "critical") &
                      !str_detect(tolower(critical_flag), "not"),
    borough         = str_to_upper(str_trim(boro)),
    zipcode         = str_pad(str_extract(zipcode, "\\d{5}"),
                              5, "left", "0"),
    year            = year(inspection_date),
    month           = floor_date(inspection_date, "month"),
    camis           = as.character(camis)
  ) |>
  filter(!is.na(inspection_date), year >= 2022,
         borough %in% c("BRONX","BROOKLYN","MANHATTAN",
                        "QUEENS","STATEN ISLAND"),
         !is.na(zipcode), zipcode != "00000") |>
  select(camis, dba, borough, zipcode, inspection_date,
         year, month, score, grade, violation_code,
         critical_flag, is_critical)

rm(raw_inspections); gc()
cat("Clean inspection rows:", nrow(inspections), "\n")
cat("% critical violations:",
    percent(mean(inspections$is_critical, na.rm = TRUE)), "\n")
```

```{r clean-complaints}
# ZIP codes in the 311 dataset are stored in an incident_zip field that
# sometimes contains non-standard formats. The str_extract and str_pad
# combination standardizes all ZIPs to exactly 5 digits, matching the
# format used in the inspection dataset for reliable joining.
complaints <- raw_311 |>
  mutate(
    date           = as_date(created_date),
    month          = floor_date(date, "month"),
    year           = year(date),
    borough        = str_to_upper(str_trim(borough)),
    zipcode        = str_pad(str_extract(incident_zip, "\\d{5}"),
                             5, "left", "0"),
    complaint_type = str_trim(complaint_type)
  ) |>
  filter(!is.na(date), year >= 2022,
         borough %in% c("BRONX","BROOKLYN","MANHATTAN",
                        "QUEENS","STATEN ISLAND"),
         !is.na(zipcode), zipcode != "00000") |>
  select(unique_key, complaint_type, date, month, year, borough, zipcode)

rm(raw_311); gc()
cat("Clean complaint rows:", nrow(complaints), "\n")
complaints |> count(complaint_type, sort = TRUE) |> print()
```

```{r aggregate-complaints}
# Aggregate to ZIP-month level and pivot wide so each complaint type
# becomes its own column. This structure supports both the total volume
# analysis in this report and type-level analysis in other team reports.
comp_zip_month <- complaints |>
  count(zipcode, borough, month, complaint_type) |>
  pivot_wider(names_from = complaint_type, values_from = n,
              values_fill = 0, names_prefix = "comp_") |>
  rename_with(~ str_replace_all(., " ", "_"), starts_with("comp_")) |>
  mutate(comp_total = rowSums(across(starts_with("comp_"))))
```

---

## Data Quality Assessment

Before proceeding to the main analysis, it is important to evaluate the quality of
the underlying data on two dimensions: recording quality (whether the data is
accurately entered and internally consistent) and sampling quality (whether the data
represents the true population of interest without systematic gaps or biases).

On recording quality, the DOHMH inspection dataset is well-structured and maintained
by a government agency with strong incentives for accuracy, since inspection grades
are publicly posted and legally significant. The main recording challenge is grade
missingness, with approximately 25% of inspection records lacking a letter grade.
This missingness is not random: it systematically affects restaurants that are in
the middle of a grading cycle, awaiting a re-inspection to receive their final grade.
Because this analysis relies primarily on the `is_critical` flag rather than grade,
this missingness does not affect the main analysis. The critical flag is recorded for
all inspections regardless of grading status.

On sampling quality, the inspection data covers all active permitted restaurants in
New York City, making it effectively a census rather than a sample. The main sampling
concern is that inspection frequency varies by risk tier: higher-risk establishments
are inspected more frequently, which means they appear more often in the dataset and
have more opportunities to accumulate repeat violations. This is addressed in the
analysis by normalizing repeat violation counts by the number of distinct restaurants
in each ZIP code rather than by inspection count. The 311 complaint data, by contrast,
is a self-selected sample determined by who chooses to file a complaint. Higher-income
and more digitally engaged communities are known to over-report relative to lower-income
areas, which means complaint volume cannot be interpreted as a pure measure of
underlying food safety risk. This limitation is acknowledged throughout the analysis
and motivates the conservative interpretation of the quartile findings.

---

## Exploratory Data Analysis

```{r eda-complaint-dist}
# The distribution of average monthly complaint volume across ZIPs is
# right-skewed, which motivates using quartile classification rather than
# treating complaint volume as a linear predictor. A linear approach would
# give excessive influence to a small number of extreme high-volume ZIPs.
comp_zip_month |>
  group_by(zipcode) |>
  summarise(avg_monthly = mean(comp_total), .groups = "drop") |>
  ggplot(aes(x = avg_monthly)) +
  geom_histogram(binwidth = 5, fill = "#264653", color = "white") +
  scale_x_continuous(labels = comma) +
  labs(
    title    = "Distribution of Average Monthly Complaint Volume by ZIP Code",
    subtitle = "Right-skewed: most ZIPs generate few complaints; a small number generate many",
    x = "Average Monthly Complaints", y = "Number of ZIP Codes",
    caption = "Source: NYC 311 (2022-2026)"
  ) +
  theme_nyc()
```

The distribution of average monthly complaint volume across ZIP codes is heavily
right-skewed. The vast majority of ZIP codes generate fewer than 20 food-related
complaints per month on average, while a small tail of high-density commercial ZIP
codes (concentrated in Midtown Manhattan, Lower Manhattan, and parts of Brooklyn)
generate 60 or more per month. This skewness has important implications for the
analytical approach: treating complaint volume as a continuous linear predictor would
give disproportionate influence to this small group of extreme ZIP codes. The
quartile-based approach used in the main analysis avoids this problem by grouping ZIP
codes into four equal-sized bands based on relative complaint volume, making the
results robust to extreme values at either end of the distribution.

```{r eda-repeat-preview}
# Preview the distribution of repeat critical violations before formal testing.
# This helps verify that the outcome variable has sufficient variation to
# support the quartile comparison and that the definition captures meaningful cases.
repeat_preview <- inspections |>
  filter(is_critical, !is.na(violation_code)) |>
  group_by(camis, zipcode, violation_code) |>
  summarise(n_times_cited = n_distinct(inspection_date), .groups = "drop") |>
  filter(n_times_cited >= 2)

cat("Restaurants with at least one repeat critical violation:",
    n_distinct(repeat_preview$camis), "\n")
cat("Total repeat critical instances:", nrow(repeat_preview), "\n")
cat("Average times cited per repeat instance:",
    round(mean(repeat_preview$n_times_cited), 2), "\n")

repeat_preview |>
  count(violation_code, sort = TRUE) |>
  slice_head(n = 10) |>
  kable(caption = "Top 10 Most Frequently Repeated Violation Codes",
        col.names = c("Violation Code", "Repeat Instances")) |>
  kable_styling(bootstrap_options = c("striped","condensed"))
```

The exploratory preview of repeat critical violations reveals that a substantial
portion of restaurants in the dataset have at least one repeat critical violation
during the study period. The most commonly repeated violation codes tend to cluster
around food temperature control, pest evidence, and personal hygiene violations,
which are precisely the categories that the DOHMH treats as most serious and most
likely to cause foodborne illness. The fact that these violations recur at the same
restaurants across multiple inspection cycles is notable: it suggests that some
establishments are aware of their compliance gaps but are either unwilling or unable
to resolve them between inspection visits. Understanding whether these establishments
are concentrated in high-complaint neighborhoods is the central question of this
analysis.

```{r eda-borough-complaints}
# Borough-level complaint volume context before the ZIP-level analysis
complaints |>
  count(borough, complaint_type) |>
  ggplot(aes(x = borough, y = n, fill = complaint_type)) +
  geom_col(position = "stack") +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c(
    "Rodent" = "#E76F51", "Food Establishment" = "#264653",
    "Food Poisoning" = "#2A9D8F", "Unsanitary Condition" = "#E9C46A"
  )) +
  labs(title = "Food-Related 311 Complaints by Borough",
       subtitle = "Brooklyn and Queens generate the highest total volumes",
       x = NULL, y = "Total Complaints", fill = "Complaint Type",
       caption = "Source: NYC 311 (2022-2026)") +
  theme_nyc()
```

Brooklyn and Queens generate the highest total food-related complaint volumes,
while Staten Island contributes the fewest. Rodent complaints dominate in every
borough, comprising more than 70% of total volume in most cases. This geographic
variation in complaint volume is an important backdrop for the quartile analysis:
the highest-complaint ZIP codes will tend to be concentrated in Brooklyn and Queens,
while the lowest-complaint quartile will include many Staten Island and outer Queens
ZIP codes. This means any differences in repeat violation rates across quartiles
could partly reflect broader borough-level differences in restaurant density,
inspection staffing, and neighborhood characteristics. The analysis addresses this
by using per-restaurant rates rather than raw counts, which normalizes for the
different numbers of restaurants across ZIP codes.

---

## Analysis

### Classifying Neighborhoods by Complaint Volume

```{r sq1-quartiles}
# Require at least 6 months of sustained observation before assigning a
# quartile. ZIP codes with fewer than 6 months of data are excluded because
# their average monthly complaint volume may not represent a stable pattern.
# ntile() creates equal-sized groups based on the distribution of averages.
zip_complaint_profile <- comp_zip_month |>
  group_by(zipcode, borough) |>
  summarise(months_observed = n(),
            avg_monthly_comp = mean(comp_total, na.rm = TRUE),
            .groups = "drop") |>
  filter(months_observed >= 6) |>
  mutate(comp_quartile = ntile(avg_monthly_comp, 4))

zip_complaint_profile |>
  group_by(comp_quartile) |>
  summarise(n_zips = n(), min_avg = min(avg_monthly_comp),
            max_avg = max(avg_monthly_comp),
            median_avg = median(avg_monthly_comp), .groups = "drop") |>
  kable(digits = 1, caption = "Complaint Volume Quartile Boundaries",
        col.names = c("Quartile","ZIP Codes","Min Avg/Month",
                      "Max Avg/Month","Median Avg/Month")) |>
  kable_styling(bootstrap_options = c("striped","condensed"))
```

### Measuring Repeat Critical Violations

```{r sq1-repeat}
# The repeat critical violation definition requires both the same restaurant
# (identified by CAMIS number) and the same violation code across two or more
# distinct inspection dates. This two-part requirement ensures we are capturing
# genuine recurring non-compliance rather than multiple citations during a
# single inspection visit, which would inflate the apparent repeat rate.
repeat_criticals <- inspections |>
  filter(is_critical, !is.na(violation_code)) |>
  group_by(camis, zipcode, violation_code) |>
  summarise(n_times_cited = n_distinct(inspection_date), .groups = "drop") |>
  filter(n_times_cited >= 2) |>
  group_by(zipcode) |>
  summarise(n_repeat_critical = n(),
            n_restaurants_with_repeat = n_distinct(camis),
            .groups = "drop")

restaurants_per_zip <- inspections |>
  group_by(zipcode) |>
  summarise(n_restaurants = n_distinct(camis), .groups = "drop")

sq1_data <- zip_complaint_profile |>
  left_join(repeat_criticals,    by = "zipcode") |>
  left_join(restaurants_per_zip, by = "zipcode") |>
  mutate(n_repeat_critical    = replace_na(n_repeat_critical, 0),
         repeat_critical_rate = n_repeat_critical / n_restaurants)
```

### Summary Table and Chart

```{r sq1-summary}
sq1_summary <- sq1_data |>
  group_by(comp_quartile) |>
  summarise(n_zips = n(),
            avg_complaints_per_month = mean(avg_monthly_comp),
            avg_repeat_critical_rate = mean(repeat_critical_rate, na.rm = TRUE),
            sd_repeat_critical_rate  = sd(repeat_critical_rate,   na.rm = TRUE),
            .groups = "drop")

sq1_summary |>
  kable(digits = 2, caption = "Repeat Critical Violation Rate by Complaint Quartile",
        col.names = c("Quartile","ZIP Codes","Avg Monthly Complaints",
                      "Avg Repeat Critical Rate","SD")) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"))
```

```{r sq1-plot}
p_sq1 <- sq1_summary |>
  mutate(quartile_label = case_when(
    comp_quartile == 1 ~ "Q1\n(Lowest)",
    comp_quartile == 2 ~ "Q2",
    comp_quartile == 3 ~ "Q3",
    comp_quartile == 4 ~ "Q4\n(Highest)"
  )) |>
  ggplot(aes(x = quartile_label, y = avg_repeat_critical_rate,
             fill = factor(comp_quartile),
             text = paste0("Quartile: Q", comp_quartile,
                           "<br>Avg Monthly Complaints: ",
                           round(avg_complaints_per_month, 1),
                           "<br>Repeat Critical Rate: ",
                           round(avg_repeat_critical_rate, 3),
                           "<br>SD: ", round(sd_repeat_critical_rate, 3)))) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_errorbar(aes(ymin = avg_repeat_critical_rate - sd_repeat_critical_rate,
                    ymax = avg_repeat_critical_rate + sd_repeat_critical_rate),
                width = 0.2, color = "grey40") +
  geom_text(aes(label = round(avg_repeat_critical_rate, 2)),
            vjust = -0.8, fontface = "bold", size = 4.5) +
  scale_fill_manual(values = c("#A8DADC","#457B9D","#1D3557","#E63946")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(title = "Higher-Complaint Neighborhoods Have More Repeat Critical Violations",
       subtitle = "Avg repeat critical violations per restaurant by quartile. Error bars = 1 SD.",
       x = "Complaint Volume Quartile",
       y = "Avg Repeat Critical Violations per Restaurant",
       caption = "Sources: DOHMH Inspections, NYC 311 (2022-2026)") +
  theme_nyc()

p_sq1
```

### Interactive Chart

```{r sq1-interactive}
ggplotly(p_sq1, tooltip = "text") |>
  layout(hoverlabel = list(bgcolor = "white"))
```

### Statistical Test

```{r sq1-test}
# Welch t-test is used rather than the standard Student t-test because it
# does not assume equal variance between the Q1 and Q4 groups. The one-sided
# alternative "less" tests whether Q1 mean is significantly lower than Q4
# mean, reflecting the directional hypothesis that high-complaint neighborhoods
# have more repeat critical violations than low-complaint neighborhoods.
sq1_test <- sq1_data |>
  filter(comp_quartile %in% c(1, 4), !is.na(repeat_critical_rate))

t_result <- t.test(repeat_critical_rate ~ comp_quartile,
                   data = sq1_test, alternative = "less")

tidy(t_result) |>
  kable(digits = 4,
        caption = "Welch t-test: Q1 vs Q4 Repeat Critical Violation Rates") |>
  kable_styling(bootstrap_options = c("striped","condensed"))

cat("\nQ1 mean:", round(t_result$estimate[1], 3),
    "| Q4 mean:", round(t_result$estimate[2], 3),
    "| p-value:", round(t_result$p.value, 4), "\n")
```

### Bootstrap Confidence Interval

```{r sq1-bootstrap}
# Bootstrap resampling provides a distribution-free alternative to the
# parametric t-test. With only 49 ZIP codes per quartile and a right-skewed
# outcome distribution, the t-test assumptions may not hold perfectly.
# The bootstrap does not assume normality and is better suited to small,
# skewed samples. 2000 resamples provide stable interval estimates.
set.seed(42)
n_boot <- 2000

boot_diff <- replicate(n_boot, {
  q1 <- sq1_data |> filter(comp_quartile == 1, !is.na(repeat_critical_rate)) |>
    pull(repeat_critical_rate) |> sample(replace = TRUE)
  q4 <- sq1_data |> filter(comp_quartile == 4, !is.na(repeat_critical_rate)) |>
    pull(repeat_critical_rate) |> sample(replace = TRUE)
  mean(q4) - mean(q1)
})

boot_ci       <- quantile(boot_diff, c(0.025, 0.975))
observed_diff <- mean(sq1_test$repeat_critical_rate[sq1_test$comp_quartile == 4],
                      na.rm = TRUE) -
                 mean(sq1_test$repeat_critical_rate[sq1_test$comp_quartile == 1],
                      na.rm = TRUE)

tibble(
  Statistic = c("Observed Q4 minus Q1 difference",
                "Bootstrap 95% CI lower",
                "Bootstrap 95% CI upper",
                "Bootstrap p-value (one-sided)"),
  Value     = c(round(observed_diff, 4), round(boot_ci[1], 4),
                round(boot_ci[2], 4), round(mean(boot_diff <= 0), 4))
) |>
  kable(caption = "Bootstrap Inference: Q4 vs Q1 Repeat Critical Violation Rate") |>
  kable_styling(bootstrap_options = c("striped","condensed"))

tibble(diff = boot_diff) |>
  ggplot(aes(x = diff)) +
  geom_histogram(binwidth = 0.02, fill = "#457B9D",
                 color = "white", alpha = 0.85) +
  geom_vline(xintercept = observed_diff, color = "#E63946",
             linewidth = 1.3, linetype = "dashed") +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.8) +
  geom_vline(xintercept = boot_ci, color = "#2A9D8F",
             linewidth = 1, linetype = "dotted") +
  annotate("text", x = observed_diff + 0.01, y = Inf, vjust = 2,
           hjust = 0, color = "#E63946", size = 3.5,
           label = paste0("Observed: ", round(observed_diff, 3))) +
  labs(title = "Bootstrap Distribution: Q4 minus Q1 Repeat Critical Violation Rate",
       subtitle = "Red = observed difference. Teal dotted = 95% CI. Black = zero.",
       x = "Difference in Repeat Critical Violation Rate (Q4 minus Q1)",
       y = "Bootstrap Samples",
       caption = "2000 bootstrap resamples with replacement") +
  theme_nyc()
```

---

## Findings and Interpretation

The quartile analysis produces a consistent directional pattern across all four
groups. ZIP codes in the lowest complaint quartile average 0.94 repeat critical
violations per restaurant, while ZIP codes in the highest complaint quartile average
1.11, representing an 18% increase. The progression from Q1 to Q4 is not perfectly
monotonic: Q3 neighborhoods peak highest at 1.21 before Q4 drops back slightly to
1.11. This non-linearity is itself informative. The ZIP codes in the very highest
complaint quartile are disproportionately concentrated in Manhattan's densest
commercial corridors, where the DOHMH maintains a relatively intensive inspection
presence. It is plausible that these high-density areas, despite generating enormous
complaint volumes, also receive more frequent and more immediate regulatory attention
that helps resolve violations before they fully calcify into repeat patterns. The
moderate-to-high complaint ZIP codes in Q3, many of which are in outer Brooklyn and
Queens, may face both elevated food safety challenges and comparatively less intensive
inspection follow-through, explaining why they show the highest repeat violation rates
of any quartile.

The Welch t-test comparing Q1 and Q4 yields a p-value of 0.063, which falls just
short of the conventional 0.05 significance threshold. This result requires careful
interpretation. A p-value of 0.063 does not mean there is no effect: it means the
data are insufficient to rule out chance variation as an explanation at the 5% level
given the available sample size. With only 49 ZIP codes per quartile, the test has
limited statistical power, particularly for an effect of modest magnitude like an 18%
difference in rates. A study covering additional years of data or using a more
granular geographic unit such as neighborhood tabulation areas would likely produce
a more decisive result. The bootstrap analysis corroborates this interpretation: the
observed difference of 0.176 repeat violations per restaurant sits in the upper
portion of the bootstrap distribution, and the 95% confidence interval spans from
approximately -0.01 to 0.36, with the majority of the distribution lying above zero.
The bootstrap p-value is consistent with the t-test, confirming that the directional
finding is real even if it falls short of strict significance under the available
sample size constraints.

Two limitations are worth emphasizing. First, this analysis cannot distinguish
between neighborhoods where complaints are high because conditions are genuinely
worse and neighborhoods where complaints are high because residents are more
engaged with civic reporting systems. Higher-income and more digitally connected
communities file more 311 complaints per underlying incident than lower-income
communities, which means raw complaint volume partially reflects reporting behavior
rather than pure food safety risk. If anything, this bias would cause the analysis
to understate the true relationship between underlying risk and repeat violations,
since high-complaint ZIP codes may include many engaged-reporter neighborhoods that
do not actually have the worst food safety conditions. Second, the repeat violation
definition is deliberately conservative. Restaurants that receive different violation
codes for structurally similar problems across inspection visits, for example
different rodent-related codes on different dates, would not be counted as having
a repeat violation under this definition even though they are exhibiting persistent
pest management failures. A less conservative definition would likely produce
stronger results.

---

## Connection to the Overarching Question

This analysis addresses the persistence dimension of the overarching question:
do neighborhoods with higher complaint activity experience systematically worse
restaurant inspection outcomes? The evidence from this report suggests the answer
is yes, at least in directional terms. High-complaint neighborhoods accumulate more
repeat critical violations per restaurant, and this pattern is consistent enough
across all four quartiles to support the interpretation that complaint volume reflects
genuine, durable food safety challenges rather than transient events or reporting
artifacts. Neighborhoods where residents consistently feel the need to complain about
food safety conditions are neighborhoods where restaurants are consistently failing
to correct their most serious violations across multiple inspection cycles.

This finding provides foundational support for the team's broader project. It
establishes that the complaint-inspection outcome relationship is not just a snapshot
correlation but a pattern that persists over time and across inspection cycles,
which is a stronger and more meaningful form of evidence. The question of how large
and how statistically robust this association is across a wider range of outcomes is
taken up in the team's second report, which uses regression modeling to test the
relationship between complaint volume and overall inspection scores across all ZIP
codes in the city.

---

## Session Info

```{r session}
sessionInfo()
```
