---
title: "Temporal Dynamics of Complaints and Inspection Outcomes"
subtitle: "STA 9750 | Individual Report | Restaurant Complaints vs Inspection Risk in NYC"
author: "Group Member 3"
date: "2026-05-12"
format:
  html:
    toc: true
    toc-float: true
    toc-depth: 3
    theme: flatly
    self-contained: true
    highlight-style: tango
    code-fold: true
    fig-width: 10
    fig-height: 6
execute:
  echo: true
  message: false
  warning: false
  fig-align: center
---
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center"
)
```

## Contribution to the Overarching Question

This report is one of five individual analyses contributing to the team project:
*Restaurant Complaints vs Inspection Risk in NYC*. The overarching question is:

> **Do neighborhoods with higher complaint activity experience systematically worse
> restaurant inspection outcomes?**

The team's SQ2 report established that complaint volume and inspection outcomes are
positively correlated at the ZIP code level. My contribution investigates the *timing*
of this relationship. Knowing that complaints and violations co-occur does not tell us
whether complaints are a leading indicator that could be used to predict and prevent
future violations, or whether both simply reflect the same persistent neighborhood
conditions at the same time. This distinction has critical policy implications: if
complaints lead violations, they can serve as an early warning system; if they are
contemporaneous, they are a concurrent diagnostic but not a predictive tool.

The specific question I address is:

> **Do increases in neighborhood-level food-related complaints precede worse
> inspection outcomes?**

I test this by examining whether complaint volume in month T predicts inspection
outcomes in months T+1, T+2, and T+3, using both lagged correlations and a panel
regression with ZIP fixed effects that controls for all stable neighborhood
characteristics., ## Setup

```{r packages}
pkgs <- c("tidyverse","lubridate","scales","knitr",
          "kableExtra","broom","glue","httr2","jsonlite")
to_install <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(to_install) > 0) install.packages(to_install)
library(tidyverse); library(lubridate); library(scales)
library(knitr); library(kableExtra); library(broom); library(glue)
```

```{r theme}
theme_nyc <- function() {
  theme_minimal(base_size = 13) +
    theme(
      plot.title    = element_text(face = "bold", size = 15),
      plot.subtitle = element_text(color = "grey40", size = 12),
      plot.caption  = element_text(color = "grey55", size = 9),
      axis.title    = element_text(size = 11),
      panel.grid.minor = element_blank()
    )
}
```, ## Data Acquisition

```{r fetch-helper}
fetch_all_socrata <- function(dataset_id, query = "",
                               select = NULL, page_size = 50000) {
  base_url <- glue(
    "https://data.cityofnewyork.us/resource/{dataset_id}.json"
  )
  offset <- 0; pages <- list()
  repeat {
    req <- request(base_url) |>
      req_url_query(`$limit` = page_size, `$offset` = offset,
                    `$where` = query)
    if (!is.null(select)) req <- req |> req_url_query(`$select` = select)
    page <- req |> req_retry(max_tries = 3) |> req_perform() |>
      resp_body_json(simplifyVector = TRUE) |> as_tibble()
    if (nrow(page) == 0) break
    message(glue("  {dataset_id}: {offset + nrow(page)} rows"))
    pages[[length(pages) + 1]] <- page
    if (nrow(page) < page_size) break
    offset <- offset + page_size
  }
  bind_rows(pages)
}
```

```{r load-data, cache=TRUE}
dir.create("data", showWarnings = FALSE)
if (!file.exists("data/raw_inspections.rds")) {
  raw_inspections <- fetch_all_socrata(
    "43nn-pn8j",
    query = "inspection_date >= '2022-01-01T00:00:00'",
    select = "camis,boro,zipcode,inspection_date,score,critical_flag"
  )
  saveRDS(raw_inspections, "data/raw_inspections.rds")
} else { raw_inspections <- readRDS("data/raw_inspections.rds") }

if (!file.exists("data/raw_311.rds")) {
  raw_311 <- fetch_all_socrata(
    "erm2-nwe9",
    query = paste("complaint_type IN ('Food Establishment',",
                  "'Food Poisoning','Rodent','Unsanitary Condition')",
                  "AND created_date >= '2022-01-01T00:00:00'"),
    select = "unique_key,complaint_type,created_date,borough,incident_zip"
  )
  saveRDS(raw_311, "data/raw_311.rds")
} else { raw_311 <- readRDS("data/raw_311.rds") }
```, ## Data Cleaning and Aggregation

```{r clean}
inspections <- raw_inspections |>
  mutate(
    inspection_date = as_date(inspection_date),
    score           = as.numeric(score),
    is_critical     = str_detect(tolower(critical_flag), "critical") &
                      !str_detect(tolower(critical_flag), "not"),
    borough         = str_to_upper(str_trim(boro)),
    zipcode         = str_pad(str_extract(zipcode, "\\d{5}"),
                              5, "left", "0"),
    year            = year(inspection_date),
    month           = floor_date(inspection_date, "month")
  ) |>
  filter(!is.na(inspection_date), year >= 2022,
         borough %in% c("BRONX","BROOKLYN","MANHATTAN",
                        "QUEENS","STATEN ISLAND"),
         !is.na(zipcode), zipcode != "00000")

complaints <- raw_311 |>
  mutate(
    date    = as_date(created_date),
    year    = year(date),
    month   = floor_date(date, "month"),
    borough = str_to_upper(str_trim(borough)),
    zipcode = str_pad(str_extract(incident_zip, "\\d{5}"),
                     5, "left", "0")
  ) |>
  filter(!is.na(date), year >= 2022,
         borough %in% c("BRONX","BROOKLYN","MANHATTAN",
                        "QUEENS","STATEN ISLAND"),
         !is.na(zipcode), zipcode != "00000")

rm(raw_inspections, raw_311); gc()

# Monthly time series: inspection outcomes per ZIP per month
zip_insp_monthly <- inspections |>
  group_by(zipcode, month) |>
  summarise(
    pct_critical = mean(is_critical, na.rm = TRUE),
    mean_score   = mean(score,       na.rm = TRUE),
    .groups = "drop"
  )

# Monthly time series: complaint totals per ZIP per month
zip_comp_monthly <- complaints |>
  count(zipcode, month, name = "comp_total")

cat("ZIP-month inspection rows:", nrow(zip_insp_monthly), "\n")
cat("ZIP-month complaint rows:",  nrow(zip_comp_monthly),  "\n")
```, ## Exploratory Data Analysis

```{r eda-monthly-trends}
# Examine city-wide monthly trends in both complaints and
# inspection outcomes to understand the temporal context
city_monthly <- zip_insp_monthly |>
  group_by(month) |>
  summarise(avg_pct_critical = mean(pct_critical, na.rm = TRUE),
            .groups = "drop") |>
  left_join(
    zip_comp_monthly |>
      group_by(month) |>
      summarise(total_comp = sum(comp_total), .groups = "drop"),
    by = "month"
  )

# Dual-axis plot: complaints (bars) vs critical violation rate (line)
ggplot(city_monthly, aes(x = month)) +
  geom_col(aes(y = total_comp / 5000), fill = "#A8DADC", alpha = 0.7) +
  geom_line(aes(y = avg_pct_critical), color = "#E63946",
            linewidth = 1.2) +
  geom_point(aes(y = avg_pct_critical), color = "#E63946", size = 2) +
  scale_y_continuous(
    name     = "Avg % Critical Violations",
    sec.axis = sec_axis(~ . * 5000, name = "Total Complaints",
                        labels = comma)
  ) +
  labs(
    title    = "City-Wide Monthly Trends: Complaints vs Critical Violations",
    subtitle = "Bars = total complaints (right axis); Line = avg % critical violations (left axis)",
    x        = NULL,
    caption  = "Sources: DOHMH Inspections, NYC 311 (2022-2026)"
  ) +
  theme_nyc()
```

The city-wide trend plot reveals that complaints and critical violation rates do not
move together in an obviously predictive fashion ,  complaint spikes (visible in summer
months due to rodent activity) do not appear to be followed by corresponding increases
in critical violation rates in subsequent months. This visual pattern anticipates the
null finding from the formal lagged analysis., ## Analysis

### Building the Lagged Panel Dataset

```{r sq3-lag}
# Construct lagged predictors by shifting complaint month forward.
# A complaint in month T appears as a predictor for outcomes in T+N,
# allowing us to test whether prior complaint activity forecasts
# future inspection deterioration.
make_lag <- function(df, lag_months) {
  df |>
    mutate(month = month + months(lag_months)) |>
    rename_with(~ paste0(.x, "_lag", lag_months), comp_total)
}

sq3_data <- zip_insp_monthly |>
  left_join(make_lag(zip_comp_monthly, 1), by = c("zipcode","month")) |>
  left_join(make_lag(zip_comp_monthly, 2), by = c("zipcode","month")) |>
  left_join(make_lag(zip_comp_monthly, 3), by = c("zipcode","month")) |>
  filter(!is.na(pct_critical))

cat("Panel rows for lagged analysis:", nrow(sq3_data), "\n")
cat("Distinct ZIPs:", n_distinct(sq3_data$zipcode), "\n")
cat("Months covered:", n_distinct(sq3_data$month), "\n")
```

### Lagged Correlations

```{r sq3-cors}
# Pearson correlations between lagged complaint volume and
# future inspection outcomes across all ZIP-month observations
lag_cors <- tibble(
  lag = c("Lag 1 month","Lag 2 months","Lag 3 months"),
  cor_pct_critical = c(
    cor(sq3_data$comp_total_lag1, sq3_data$pct_critical,
        use = "pairwise.complete.obs"),
    cor(sq3_data$comp_total_lag2, sq3_data$pct_critical,
        use = "pairwise.complete.obs"),
    cor(sq3_data$comp_total_lag3, sq3_data$pct_critical,
        use = "pairwise.complete.obs")
  ),
  cor_mean_score = c(
    cor(sq3_data$comp_total_lag1, sq3_data$mean_score,
        use = "pairwise.complete.obs"),
    cor(sq3_data$comp_total_lag2, sq3_data$mean_score,
        use = "pairwise.complete.obs"),
    cor(sq3_data$comp_total_lag3, sq3_data$mean_score,
        use = "pairwise.complete.obs")
  )
)

lag_cors |>
  kable(digits = 4,
        caption = "Lagged Correlations ,  Complaints in T vs Outcomes in T+N",
        col.names = c("Lag Window",
                      "Cor w/ % Critical","Cor w/ Mean Score")) |>
  kable_styling(bootstrap_options = c("striped","condensed"))
```

```{r sq3-plot}
lag_cors |>
  pivot_longer(starts_with("cor_"),
               names_to = "outcome", values_to = "correlation") |>
  mutate(outcome = recode(outcome,
    "cor_pct_critical" = "% Critical Violations",
    "cor_mean_score"   = "Mean Inspection Score")) |>
  ggplot(aes(x = lag, y = correlation, fill = outcome)) +
  geom_col(position = "dodge", width = 0.5) +
  geom_hline(yintercept = 0, linewidth = 0.8) +
  scale_fill_manual(values = c("#E76F51","#264653")) +
  labs(
    title    = "Do Complaint Spikes Predict Future Inspection Outcomes?",
    subtitle = "Lagged correlations: complaints in month T vs outcomes in T+N",
    x = "Complaint Lag Window", y = "Pearson Correlation", fill = "Outcome",
    caption = "Sources: DOHMH Inspections, NYC 311 (2022-2026)"
  ) +
  theme_nyc()
```

### Panel Regression with ZIP Fixed Effects

```{r sq3-regression}
# ZIP fixed effects remove all time-invariant neighborhood
# characteristics, isolating whether within-ZIP changes in complaint
# volume predict within-ZIP changes in inspection outcomes.
# This is a much stricter test than the lagged correlations above.
sq3_model <- lm(
  pct_critical ~ comp_total_lag1 + comp_total_lag2 +
                 comp_total_lag3 + factor(zipcode),
  data = sq3_data
)

# Show only lag coefficients ,  suppress ZIP fixed effects
tidy(sq3_model) |>
  filter(str_detect(term, "comp_total_lag")) |>
  kable(digits = 5,
        caption = "Panel Regression (ZIP FE): Lagged Complaints -> % Critical") |>
  kable_styling(bootstrap_options = c("striped","condensed"))
```, ## Findings and Interpretation

The lagged correlation analysis produces consistently positive but small correlations
between prior complaint volume and subsequent inspection outcomes ,  ranging from 0.028
to 0.076 across all lag windows and outcomes. While positive, these magnitudes are
substantially smaller than the contemporaneous associations reported in the team's
SQ2 analysis (r = 0.41 between complaint volume and inspection score at the ZIP level),
indicating that most of the complaint-outcome association operates contemporaneously
rather than through a temporal leading relationship.

The panel regression with ZIP fixed effects provides the most stringent test. After
removing stable neighborhood characteristics through ZIP fixed effects, none of the
three lagged complaint predictors reaches statistical significance (all p > 0.20).
The coefficient on lag-3 complaints is even slightly negative, though not significantly
so. Taken together, these results constitute an important and honest null finding:
complaints do not reliably predict future inspection deterioration within the 1–3 month
window tested here.

This does not mean that complaints are uninformative ,  the team's SQ2 analysis
demonstrates a robust cross-sectional association. Rather, the null temporal finding
reveals the mechanism: complaints and violations co-occur because they are both
responses to the same persistent underlying neighborhood conditions, rather than
complaints causing or predicting future violations. A neighborhood with chronic food
safety challenges generates both complaints and violations in the same period.

There are several reasons why a genuine predictive relationship might be difficult to
detect even if one exists. The monthly aggregation introduces noise because inspection
cycles are irregular and do not align with calendar months. The three-month lag window
may also be too narrow ,  complaint-triggered inspections (demonstrated in the team's
SQ5 analysis) happen within days, while the regulatory effects on violation rates may
take longer to materialize. Finally, the within-ZIP variation available for identification
after removing fixed effects is limited, reducing statistical power. These structural
limitations mean the null finding should be interpreted cautiously rather than as
definitive evidence against a lagged relationship., ## Connection to the Overarching Question

This analysis refines the team's answer to the overarching question. The relationship
between complaint activity and inspection outcomes is real (established by SQ1 and SQ2)
but contemporaneous rather than predictive. Both complaints and violations are
simultaneous manifestations of the same underlying neighborhood food safety environment.
This finding is important for policy: it suggests that complaint data is most useful as
a concurrent diagnostic of neighborhood risk rather than as an early warning system.
The question of whether complaints at least trigger faster inspections ,  even if they
do not predict future violation rates ,  is taken up in the team's SQ5 report., ## Session Info

```{r session}
sessionInfo()
```
