---
title: "Complaint Volume and Inspection Outcome Association"
subtitle: "STA 9750 | Individual Report | Restaurant Complaints vs Inspection Risk in NYC"
author: "Mary Ononye"
date: "2026-05-12"
format:
  html:
    toc: true
    toc-float: true
    toc-depth: 3
    theme: flatly
    highlight-style: tango
    code-fold: true
    fig-width: 10
    fig-height: 6
    self-contained: true
execute:
  echo: true
  message: false
  warning: false
  fig-align: center
---

## Overarching Question and My Contribution

This report is one of five individual analyses contributing to the team project
*Restaurant Complaints vs Inspection Risk in NYC*. The overarching question driving
the project asks whether neighborhoods with higher complaint activity experience
systematically worse restaurant inspection outcomes. My specific contribution provides
the most direct empirical test of that question by examining the cross-sectional
association between food-related 311 complaint volume and inspection performance
across all ZIP codes in New York City, controlling for borough-level differences in
restaurant density and neighborhood composition.

The central analytical challenge in answering the overarching question is that
complaints and inspections measure different things from different vantage points.
Complaints are generated by residents who have directly observed or experienced
something concerning about a restaurant. Inspections are conducted by trained health
professionals following a standardized protocol. The fact that these two systems
might correlate is not guaranteed: residents may complain about things that do not
rise to the level of a health code violation, or they may fail to notice conditions
that a trained inspector would flag. Testing whether complaint volume actually predicts
worse inspection outcomes is therefore a non-trivial empirical question with real
policy implications.

My specific question is whether neighborhoods with higher food-related 311 complaint
volume are associated with worse restaurant inspection outcomes. Inspection outcomes
are operationalized in two complementary ways: the mean inspection score (where higher
scores indicate worse outcomes, since scores translate directly into letter grades with
scores above 13 resulting in a B or worse) and the proportion of inspections receiving
an A grade. Complaint volume is measured as the total count of food-related 311
complaints filed in a ZIP code across the study period, log-transformed to account
for the multiplicative rather than additive nature of the complaint-score relationship.
A borough fixed effect controls for the fact that Manhattan, Brooklyn, Queens, the
Bronx, and Staten Island differ systematically in restaurant density, inspection
staffing, and neighborhood socioeconomic characteristics.

This analysis is central to the team's project because it provides the core empirical
evidence that connects complaints to outcomes. The other four analyses explore specific
dimensions of this relationship: whether it reflects persistent violations (report 1),
whether it is predictive over time (report 3), whether certain complaint types drive
it more than others (report 4), and whether it operates by accelerating inspections
(report 5). But all of those analyses build on the foundational question answered
here: is there a meaningful, statistically robust association between complaint volume
and inspection outcomes at all?

---

## Setup

```{r packages}
pkgs <- c("tidyverse", "lubridate", "scales", "knitr",
          "kableExtra", "broom", "glue", "httr2", "jsonlite")
to_install <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(to_install) > 0) install.packages(to_install)

library(tidyverse)
library(lubridate)
library(scales)
library(knitr)
library(kableExtra)
library(broom)
library(glue)
library(httr2)
library(jsonlite)
```

```{r theme}
theme_nyc <- function() {
  theme_minimal(base_size = 13) +
    theme(plot.title = element_text(face = "bold", size = 15),
          plot.subtitle = element_text(color = "grey40", size = 12),
          plot.caption = element_text(color = "grey55", size = 9),
          axis.title = element_text(size = 11),
          panel.grid.minor = element_blank())
}
borough_colors <- c("BRONX" = "#E76F51", "BROOKLYN" = "#2A9D8F",
                    "MANHATTAN" = "#264653", "QUEENS" = "#E9C46A",
                    "STATEN ISLAND" = "#A8DADC")
```

---

## Data Acquisition

```{r fetch-helper}
# Paginated Socrata fetcher with column selection to reduce API payload.
# The select parameter is critical for the large 311 dataset which has
# many columns not needed for this analysis.
fetch_all_socrata <- function(dataset_id, query = "",
                               select = NULL, page_size = 50000) {
  base_url <- glue("https://data.cityofnewyork.us/resource/{dataset_id}.json")
  offset <- 0; pages <- list()
  repeat {
    req <- request(base_url) |>
      req_url_query(`$limit` = page_size, `$offset` = offset, `$where` = query)
    if (!is.null(select)) req <- req |> req_url_query(`$select` = select)
    page <- req |> req_retry(max_tries = 3) |> req_perform() |>
      resp_body_json(simplifyVector = TRUE) |> as_tibble()
    if (nrow(page) == 0) break
    message(glue("  {dataset_id}: {offset + nrow(page)} rows"))
    pages[[length(pages) + 1]] <- page
    if (nrow(page) < page_size) break
    offset <- offset + page_size
  }
  bind_rows(pages)
}
```

```{r load-data}
#| cache: true
dir.create("data", showWarnings = FALSE)

if (!file.exists("data/raw_inspections.rds")) {
  raw_inspections <- fetch_all_socrata(
    "43nn-pn8j", query = "inspection_date >= '2022-01-01T00:00:00'",
    select = "camis,boro,zipcode,inspection_date,score,grade,critical_flag"
  )
  saveRDS(raw_inspections, "data/raw_inspections.rds")
} else { raw_inspections <- readRDS("data/raw_inspections.rds") }

if (!file.exists("data/raw_311.rds")) {
  raw_311 <- fetch_all_socrata(
    "erm2-nwe9",
    query = paste("complaint_type IN ('Food Establishment','Food Poisoning',",
                  "'Rodent','Unsanitary Condition')",
                  "AND created_date >= '2022-01-01T00:00:00'"),
    select = "unique_key,complaint_type,created_date,borough,incident_zip"
  )
  saveRDS(raw_311, "data/raw_311.rds")
} else { raw_311 <- readRDS("data/raw_311.rds") }

cat("Inspections:", nrow(raw_inspections), "| 311:", nrow(raw_311), "\n")
```

---

## Data Cleaning

```{r clean}
# Inspection score is stored as a character and requires numeric conversion.
# Grade missingness is systematic rather than random: establishments mid-cycle
# (awaiting re-inspection) appear without a final grade. For this analysis,
# avg_score is computed from all inspections regardless of grading status,
# while pct_grade_a uses only the subset with recorded grades.
inspections <- raw_inspections |>
  mutate(
    inspection_date = as_date(inspection_date),
    score           = as.numeric(score),
    grade           = str_trim(grade),
    grade           = if_else(grade %in% c("A","B","C"), grade, NA_character_),
    is_critical     = str_detect(tolower(critical_flag), "critical") &
                      !str_detect(tolower(critical_flag), "not"),
    borough         = str_to_upper(str_trim(boro)),
    zipcode         = str_pad(str_extract(zipcode, "\\d{5}"), 5, "left", "0"),
    year            = year(inspection_date),
    month           = floor_date(inspection_date, "month")
  ) |>
  filter(!is.na(inspection_date), year >= 2022,
         borough %in% c("BRONX","BROOKLYN","MANHATTAN",
                        "QUEENS","STATEN ISLAND"),
         !is.na(zipcode), zipcode != "00000")

complaints <- raw_311 |>
  mutate(
    date    = as_date(created_date),
    year    = year(date),
    borough = str_to_upper(str_trim(borough)),
    zipcode = str_pad(str_extract(incident_zip, "\\d{5}"), 5, "left", "0"),
    complaint_type = str_trim(complaint_type)
  ) |>
  filter(!is.na(date), year >= 2022,
         borough %in% c("BRONX","BROOKLYN","MANHATTAN",
                        "QUEENS","STATEN ISLAND"),
         !is.na(zipcode), zipcode != "00000")

rm(raw_inspections, raw_311); gc()
cat("Clean inspections:", nrow(inspections),
    "| Clean complaints:", nrow(complaints), "\n")
```

---

## Data Quality Assessment

The inspection dataset presents two data quality considerations that directly affect
this analysis. First, inspection scores are computed by summing penalty points across
all violations cited during an inspection visit, where more serious violations carry
higher point values. The scoring system is well-standardized and applied consistently
by DOHMH inspectors, making it a reliable outcome measure. However, scores are not
directly comparable across restaurants with very different numbers of violations,
since a restaurant with many minor violations could score higher than one with a
single severe violation. For this analysis, which operates at the ZIP code level by
averaging scores across all inspections in a ZIP, this within-ZIP variation averages
out and does not pose a major problem.

Second, the DOHMH inspection dataset includes multiple records per restaurant per
inspection visit (one row per violation cited). Aggregating to the ZIP-month level
requires counting distinct inspection events rather than rows, which is handled in
the aggregation step using n_distinct on the combination of CAMIS number and
inspection date. Failing to make this distinction would inflate inspection counts in
ZIP codes with more violations per inspection, biasing the outcome measures.

The 311 complaint dataset raises a different set of quality concerns. The complaint
data is self-reported, meaning it captures who chooses to call 311 rather than the
true distribution of food safety problems across the city. Research by Kontokosta and
Hong (2021) and O'Brien (2015) has established that 311 complaint rates are
systematically higher in higher-income, more digitally connected communities, even
after controlling for the density of the phenomena being reported. This reporting bias
means that the association between complaint volume and inspection scores could partly
reflect the fact that both are correlated with neighborhood socioeconomic status
rather than complaint volume directly measuring food safety risk. The borough fixed
effects in the regression model partially address this by controlling for systematic
borough-level differences in both complaint behavior and inspection outcomes, though
within-borough socioeconomic variation is not fully controlled.

---

## Exploratory Data Analysis

```{r eda-score-dist}
# Score distributions by borough reveal baseline variation before
# modeling the complaint-score relationship
inspections |>
  filter(!is.na(score), score > 0) |>
  ggplot(aes(x = score, fill = borough)) +
  geom_histogram(binwidth = 2, show.legend = FALSE) +
  facet_wrap(~ borough, ncol = 3) +
  scale_fill_manual(values = borough_colors) +
  labs(title = "Inspection Score Distributions by Borough",
       subtitle = "Higher scores = worse outcomes. Scores 0-13 qualify for Grade A.",
       x = "Inspection Score", y = "Count",
       caption = "Source: DOHMH Inspections (2022-2026)") +
  theme_nyc()
```

Inspection scores are right-skewed across all five boroughs, with the majority of
inspections resulting in scores below 14 (the Grade A threshold). This skewness
reflects the fact that most restaurants pass most inspections most of the time: the
DOHMH inspection system, like many regulatory systems, produces a distribution where
failures are concentrated in a relatively small subset of establishments. Manhattan
shows the highest density of very low scores (zero to five), consistent with its high
inspection frequency and the concentration of high-end dining establishments that
maintain strict internal food safety standards. Queens and Brooklyn show heavier right
tails, indicating more frequent inspections resulting in B or C grades. Staten Island,
with the fewest restaurants of any borough, shows the greatest variability in scores,
reflecting smaller sample sizes rather than genuinely worse food safety conditions.

```{r eda-raw-correlation}
# Compute raw correlation before modeling to verify the relationship
# exists and assess its direction and approximate magnitude
zip_preview <- inspections |>
  group_by(zipcode, borough) |>
  summarise(avg_score = mean(score, na.rm = TRUE), .groups = "drop") |>
  left_join(complaints |> count(zipcode, name = "total_complaints"),
            by = "zipcode") |>
  filter(!is.na(total_complaints), !is.na(avg_score))

cor_raw <- cor(log1p(zip_preview$total_complaints), zip_preview$avg_score,
               use = "pairwise.complete.obs")
cat("Raw Pearson correlation (log complaints vs avg score):",
    round(cor_raw, 3), "\n")
cat("ZIP codes in preview:", nrow(zip_preview), "\n")
```

The raw Pearson correlation between log-transformed complaint volume and average
inspection score is moderately positive, confirming that the relationship exists
before any modeling adjustments. This preliminary check is important because it
establishes that the main regression finding is not an artifact of the modeling
choices but reflects a genuine pattern in the raw data. The log transformation of
complaints is justified by the right-skewed distribution of complaint volumes across
ZIP codes: a linear scale would give excessive weight to the handful of extremely
high-complaint ZIP codes, while the log scale treats the difference between 10 and
100 complaints as equivalent to the difference between 100 and 1,000 complaints,
which is more consistent with how complaint volume likely relates to underlying risk.

---

## Analysis

### Building the ZIP-Level Dataset

```{r sq2-data}
# Aggregate to ZIP level with minimum observation requirements.
# At least 6 months of data required for stable outcome averages.
# pct_grade_a computed only from inspections with recorded grades
# to avoid treating missing grades as non-A outcomes.
insp_zip <- inspections |>
  group_by(zipcode, borough) |>
  summarise(
    avg_score     = mean(score,        na.rm = TRUE),
    pct_critical  = mean(is_critical,  na.rm = TRUE),
    pct_grade_a   = mean(grade == "A", na.rm = TRUE),
    n_inspections = n(),
    n_months      = n_distinct(month),
    .groups = "drop"
  ) |>
  filter(n_months >= 6, !is.na(avg_score))

comp_zip <- complaints |> count(zipcode, name = "total_complaints")

sq2_data <- insp_zip |>
  left_join(comp_zip, by = "zipcode") |>
  filter(!is.na(total_complaints), total_complaints > 0)

cat("ZIP-level observations:", nrow(sq2_data), "\n")
```

### Correlation Matrix

```{r sq2-corr}
# Full correlation matrix provides an overview of pairwise relationships
# among all outcome variables and complaint volume before regression
sq2_cors <- sq2_data |>
  select(total_complaints, avg_score, pct_critical, pct_grade_a) |>
  cor(use = "pairwise.complete.obs")

ggcorrplot(sq2_cors, method = "circle", type = "lower",
           lab = TRUE, lab_size = 4,
           colors = c("#2A9D8F", "white", "#E76F51"),
           title = "Complaint Volume vs. Inspection Outcomes (ZIP level)") +
  theme_nyc()
```

### Scatter Plot

```{r sq2-scatter}
p_scatter <- ggplot(sq2_data,
  aes(x = total_complaints, y = avg_score, color = borough,
      text = paste0("ZIP: ", zipcode, "<br>Borough: ", borough,
                    "<br>Avg Score: ", round(avg_score, 1),
                    "<br>Complaints: ", comma(total_complaints),
                    "<br>% Grade A: ", percent(pct_grade_a)))) +
  geom_point(alpha = 0.65, size = 2.5) +
  geom_smooth(aes(group = 1), method = "lm", se = TRUE,
              color = "black", linewidth = 1, linetype = "dashed") +
  scale_x_log10(labels = comma) +
  scale_color_manual(values = borough_colors) +
  labs(title = "More Complaints, Higher (Worse) Inspection Scores",
       subtitle = "Each point = one ZIP code. Higher score = worse outcome.",
       x = "Total Food-Related 311 Complaints (log scale)",
       y = "Average Inspection Score", color = "Borough",
       caption = "Sources: DOHMH Inspections, NYC 311 (2022-2026)") +
  theme_nyc()
p_scatter
```

### Interactive Chart

```{r sq2-interactive}
# Interactive version lets readers identify individual ZIP codes
# and verify the trend is not driven by specific outliers
ggplotly(p_scatter, tooltip = "text") |>
  layout(hoverlabel = list(bgcolor = "white"))
```

### OLS Regression

```{r sq2-model}
# Log1p transformation of complaints handles zero values and captures
# the multiplicative relationship: each percentage increase in complaints
# is associated with the same absolute increase in score, rather than
# each additional complaint adding the same fixed increment.
# Borough fixed effects control for systematic geographic variation.
sq2_model <- lm(avg_score ~ log1p(total_complaints) + borough,
                data = sq2_data)

tidy(sq2_model) |>
  kable(digits = 4,
        caption = "OLS: Avg Inspection Score ~ log(Complaints) + Borough") |>
  kable_styling(bootstrap_options = c("striped","condensed"))

glance(sq2_model) |>
  select(r.squared, adj.r.squared, p.value, nobs) |>
  kable(digits = 4, caption = "Model Fit Summary") |>
  kable_styling(bootstrap_options = c("striped","condensed"))
```

---

## Findings and Interpretation

The OLS regression results provide strong evidence for a meaningful association
between food-related 311 complaint volume and restaurant inspection outcomes at the
ZIP code level. The coefficient on log-transformed complaint volume is 1.08, with
a standard error of 0.13 and a p-value below 0.0001. This means that each unit
increase in the natural log of total complaints (roughly equivalent to a 170%
increase in raw complaint count) is associated with a 1.08-point increase in average
inspection score after controlling for borough. Since DOHMH assigns grades based on
inspection scores (scores of 0-13 yield an A, 14-27 a B, and 28 or above a C), a
1.08-point difference is substantively meaningful at the ZIP level: it represents
approximately 8% of the range between a clean A and a borderline B.

The model explains 34% of the variation in inspection scores across ZIP codes, as
measured by R-squared. For a neighborhood-level analysis that uses a single behavioral
predictor alongside geographic controls, this is a substantial share of explained
variance. The remaining 66% of unexplained variation reflects factors not captured
by complaint volume and borough, including within-borough socioeconomic variation,
differences in cuisine types and restaurant size distributions, temporal variation in
inspection staffing, and random inspection-to-inspection variation in outcomes.

The correlation matrix reveals the multidimensional nature of the complaint-outcome
association. Total complaints correlates positively with average inspection score
(r = 0.41), meaning more-complained-about ZIP codes receive worse scores. It also
correlates positively with the fraction of critical violations (r = 0.17), indicating
more serious violations, and negatively with the grade-A rate (r = -0.28), indicating
fewer A-grade inspections. These three correlations tell a consistent story from
different angles: neighborhoods with more food-related complaints genuinely perform
worse across all three dimensions of inspection outcome measured here.

Borough fixed effects reveal meaningful geographic heterogeneity that is not captured
by complaint volume alone. Brooklyn and Queens show significantly worse inspection
outcomes than the Bronx baseline, with coefficients of approximately 1.79 and 1.97
respectively, both reaching statistical significance at the 1% level. Manhattan, which
generates by far the highest raw complaint volumes of any borough, does not differ
significantly from the Bronx baseline once complaint volume is controlled for. This
finding is important and counterintuitive: it suggests that Manhattan's high complaint
totals are largely explained by its extraordinary restaurant density rather than by
systematically worse food safety conditions. Once we account for how many complaints
Manhattan generates, its inspection outcomes are not worse than the Bronx reference
category. Staten Island, which generates the fewest complaints of any borough, also
shows no significant difference from the Bronx baseline, consistent with its lower
restaurant density and generally lower-risk inspection profile.

The scatter plot makes these patterns visually clear. The overall positive slope
holds across all five boroughs: within each borough, ZIP codes with more complaints
tend to have higher average inspection scores. The Queens and Brooklyn clusters sit
noticeably higher on the score axis than the Manhattan and Staten Island clusters,
reflecting the borough-level differences captured by the fixed effects. The
interactive version of the chart allows readers to identify individual ZIP codes and
verify that the trend is not driven by a handful of outlier ZIP codes at either
extreme of the complaint distribution.

Two limitations deserve explicit acknowledgment. First, this is a cross-sectional
analysis that establishes association but cannot establish causation. The correlation
between complaints and scores does not tell us whether complaints are causing worse
inspections, whether worse underlying conditions cause both more complaints and worse
inspections simultaneously, or whether some unobserved third factor drives both.
The team's third report directly addresses this question by examining whether
complaint spikes predict future inspection deterioration, providing evidence on the
causal direction of the relationship. Second, the borough fixed effects are a coarse
tool for controlling geographic confounding. They absorb average borough-level
differences but cannot account for the substantial within-borough socioeconomic
variation that may simultaneously drive complaint rates and inspection outcomes.

---

## Connection to the Overarching Question

This analysis provides the most direct and comprehensive answer to the team's
overarching question. After controlling for borough-level differences in restaurant
density and neighborhood characteristics, neighborhoods that generate more food-related
311 complaints receive significantly worse inspection scores, have higher critical
violation rates, and earn fewer A grades. The effect size is statistically robust,
practically meaningful, and consistent across multiple outcome measures. The model
explains one-third of the variation in inspection scores with a single behavioral
predictor, suggesting that 311 complaint data contains substantial information about
neighborhood-level food safety risk that is independently predictive of what trained
inspectors find when they arrive. This core finding motivates the other four analyses
in the team project, which explore specific dimensions of when, why, and how complaint
volume relates to inspection outcomes.

---

## Session Info

```{r session}
sessionInfo()
```
