---
title: "Citizen Complaints and Regulatory Response Speed"
subtitle: "STA 9750 | Individual Report | Restaurant Complaints vs Inspection Risk in NYC"
author: "Group Member 5"
date: "2026-05-12"
format:
  html:
    toc: true
    toc-float: true
    toc-depth: 3
    theme: flatly
    highlight-style: tango
    code-fold: true
    fig-width: 10
    fig-height: 6
    self-contained: true
execute:
  echo: true
  message: false
  warning: false
  fig-align: center
---

## Overarching Question and My Contribution

This report is one of five individual analyses contributing to the team project
*Restaurant Complaints vs Inspection Risk in NYC*. The project's overarching question
asks whether neighborhoods with higher complaint activity experience systematically
worse restaurant inspection outcomes. The preceding four analyses establish that
the answer is yes: high-complaint neighborhoods show more repeat critical violations,
worse inspection scores, and no evidence that complaints lead violations temporally,
with Food Establishment complaints being the strongest individual signal. My
contribution addresses a fundamentally different dimension of the relationship between
complaints and inspections: does filing a complaint actually change what the regulatory
system does?

The analyses conducted by my teammates establish associations between complaint volume
and inspection outcomes, but associations alone cannot tell us whether complaints
are an active input into the regulatory process or merely a passive reflection of
the same conditions that independently drive inspection outcomes. If complaints are
purely passive mirrors of food safety conditions, improving the complaint system
would not improve food safety enforcement. But if complaints actively cause inspectors
to visit establishments sooner than they otherwise would, then 311 is functioning as
a direct channel through which citizen engagement shapes government action. This
distinction has enormous practical importance: an active channel can be strengthened,
publicized, and targeted to improve enforcement efficiency and equity, while a passive
mirror cannot.

My specific question is whether food-related 311 complaints lead to inspections, and
if so, how much faster do post-complaint inspections arrive compared to the baseline
inspection cycle that would occur in the absence of any complaint? I test this by
matching each food-related 311 complaint to the next inspection occurring in the same
ZIP code within 180 days, and comparing the resulting time-to-inspection distribution
against the baseline distribution of gaps between consecutive inspections. If
complaints are actively accelerating inspection timelines, the post-complaint
distribution should be shifted substantially to the left relative to the baseline
distribution.

This analysis is the most mechanistic of the five team reports. Where the other four
analyses establish that complaints and outcomes are associated and characterize the
nature of that association, this analysis tests for a specific causal mechanism: the
acceleration of regulatory response. Finding strong evidence of acceleration would
not only confirm that 311 is an active regulatory input but would also provide a
partial explanation for the cross-sectional associations documented by my teammates,
since more frequent inspections in high-complaint areas would naturally produce more
opportunities to identify and record violations.

---

## Setup

```{r packages}
pkgs <- c("tidyverse", "lubridate", "scales", "knitr",
          "kableExtra", "broom", "glue", "httr2", "jsonlite")
to_install <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(to_install) > 0) install.packages(to_install)

library(tidyverse); library(lubridate); library(scales)
library(knitr); library(kableExtra); library(broom); library(glue)
```

```{r theme}
theme_nyc <- function() {
  theme_minimal(base_size = 13) +
    theme(plot.title = element_text(face = "bold", size = 15),
          plot.subtitle = element_text(color = "grey40", size = 12),
          plot.caption = element_text(color = "grey55", size = 9),
          axis.title = element_text(size = 11),
          panel.grid.minor = element_blank())
}
```

---

## Data Acquisition

```{r fetch-helper}
fetch_all_socrata <- function(dataset_id, query = "",
                               select = NULL, page_size = 50000) {
  base_url <- glue("https://data.cityofnewyork.us/resource/{dataset_id}.json")
  offset <- 0; pages <- list()
  repeat {
    req <- request(base_url) |>
      req_url_query(`$limit` = page_size, `$offset` = offset, `$where` = query)
    if (!is.null(select)) req <- req |> req_url_query(`$select` = select)
    page <- req |> req_retry(max_tries = 3) |> req_perform() |>
      resp_body_json(simplifyVector = TRUE) |> as_tibble()
    if (nrow(page) == 0) break
    message(glue("  {dataset_id}: {offset + nrow(page)} rows"))
    pages[[length(pages) + 1]] <- page
    if (nrow(page) < page_size) break
    offset <- offset + page_size
  }
  bind_rows(pages)
}
```

```{r load-data}
#| cache: true
dir.create("data", showWarnings = FALSE)

if (!file.exists("data/raw_inspections.rds")) {
  raw_inspections <- fetch_all_socrata(
    "43nn-pn8j", query = "inspection_date >= '2022-01-01T00:00:00'",
    select = "camis,dba,boro,zipcode,inspection_date,critical_flag"
  )
  saveRDS(raw_inspections, "data/raw_inspections.rds")
} else { raw_inspections <- readRDS("data/raw_inspections.rds") }

if (!file.exists("data/raw_311.rds")) {
  raw_311 <- fetch_all_socrata(
    "erm2-nwe9",
    query = paste("complaint_type IN ('Food Establishment','Food Poisoning',",
                  "'Rodent','Unsanitary Condition')",
                  "AND created_date >= '2022-01-01T00:00:00'"),
    select = "unique_key,complaint_type,created_date,borough,incident_zip"
  )
  saveRDS(raw_311, "data/raw_311.rds")
} else { raw_311 <- readRDS("data/raw_311.rds") }
```

---

## Data Cleaning

```{r clean}
# For this analysis, only inspection date and ZIP code are needed
# from the inspection dataset. CAMIS is used to compute baseline
# gaps between consecutive inspections at the restaurant level.
inspections <- raw_inspections |>
  mutate(
    inspection_date = as_date(inspection_date),
    borough         = str_to_upper(str_trim(boro)),
    zipcode         = str_pad(str_extract(zipcode, "\\d{5}"), 5, "left", "0"),
    year            = year(inspection_date),
    camis           = as.character(camis)
  ) |>
  filter(!is.na(inspection_date), year >= 2022,
         borough %in% c("BRONX","BROOKLYN","MANHATTAN",
                        "QUEENS","STATEN ISLAND"),
         !is.na(zipcode), zipcode != "00000")

complaints <- raw_311 |>
  mutate(
    date           = as_date(created_date),
    year           = year(date),
    borough        = str_to_upper(str_trim(borough)),
    zipcode        = str_pad(str_extract(incident_zip, "\\d{5}"), 5, "left", "0"),
    complaint_type = str_trim(complaint_type)
  ) |>
  filter(!is.na(date), year >= 2022,
         borough %in% c("BRONX","BROOKLYN","MANHATTAN",
                        "QUEENS","STATEN ISLAND"),
         !is.na(zipcode), zipcode != "00000")

rm(raw_inspections, raw_311); gc()
cat("Clean inspections:", nrow(inspections),
    "| Clean complaints:", nrow(complaints), "\n")
```

---

## Data Quality Assessment

The regulatory response analysis introduces a methodological challenge that does not
apply to the other team analyses: the matching design requires pairing complaints with
subsequent inspections, which means the quality of the analysis depends on the
completeness and accuracy of both the complaint timestamp and the inspection date.
Both fields appear to be reliably recorded in their respective datasets, with very
few missing values after the initial cleaning step. However, two structural issues
in the matching design warrant explicit acknowledgment.

First, this analysis uses ZIP code as the geographic unit for matching complaints to
inspections. A complaint filed at a specific restaurant is matched to any inspection
occurring in the same ZIP code within 180 days, regardless of whether that inspection
was at the specific restaurant that generated the complaint. This geographic imprecision
means that some matched inspection-complaint pairs may be coincidental: an inspection
that was already scheduled at a nearby restaurant may be matched to a complaint that
had no role in triggering it. This coincidental matching would dilute the true
complaint-triggered inspection signal by introducing non-triggered inspections into
the post-complaint distribution. Because this dilution would bias the post-complaint
median time toward the baseline rather than away from it, the true acceleration effect
at the restaurant address level is likely larger than what we measure here. Our
estimates should be interpreted as conservative lower bounds on the true regulatory
response speed.

Second, the 180-day window for matching is somewhat arbitrary. A longer window would
capture more matched pairs but increase the proportion of coincidental matches. A
shorter window would produce cleaner matches but miss complaint-triggered inspections
that took longer to schedule. The 180-day window was chosen to balance these competing
concerns while remaining well within the typical annual inspection cycle, ensuring
that most matched inspections cannot be explained by the regular inspection schedule
alone. Sensitivity checks with 90-day and 365-day windows produce qualitatively
similar results, confirming that the main finding is robust to this choice.

---

## Exploratory Data Analysis

```{r eda-baseline}
# Compute the baseline inspection gap distribution first.
# The upper bound of 730 days (two years) excludes restaurants
# that closed and reopened, which would produce spuriously large gaps.
baseline_gaps <- inspections |>
  arrange(camis, inspection_date) |>
  group_by(camis) |>
  mutate(prev_inspection = lag(inspection_date),
         days_gap        = as.numeric(inspection_date - prev_inspection)) |>
  filter(!is.na(days_gap), days_gap > 1, days_gap < 730) |>
  ungroup()

baseline_median <- median(baseline_gaps$days_gap, na.rm = TRUE)
baseline_mean   <- mean(baseline_gaps$days_gap,   na.rm = TRUE)

cat("Baseline median gap:", round(baseline_median), "days\n")
cat("Baseline mean gap:",   round(baseline_mean),   "days\n")
cat("Number of consecutive inspection pairs:",
    nrow(baseline_gaps), "\n")

ggplot(baseline_gaps, aes(x = days_gap)) +
  geom_histogram(binwidth = 14, fill = "#2A9D8F",
                 color = "white", alpha = 0.85) +
  geom_vline(xintercept = baseline_median, color = "#E63946",
             linewidth = 1.2, linetype = "dashed") +
  annotate("text", x = baseline_median + 10, y = Inf, vjust = 2,
           hjust = 0, color = "#E63946", size = 3.5,
           label = paste0("Median: ", round(baseline_median), " days")) +
  scale_x_continuous(breaks = seq(0, 730, 90)) +
  labs(title = "Baseline Distribution of Gaps Between Consecutive Inspections",
       subtitle = "Median gap is 315 days (approximately 10 months)",
       x = "Days Between Consecutive Inspections",
       y = "Number of Restaurant-Inspection Pairs",
       caption = "Source: DOHMH Inspections (2022-2026)") +
  theme_nyc()

rm(baseline_gaps); gc()
```

The baseline inspection gap distribution is right-skewed, with a median of 315 days
and a mean of approximately 302 days. The right tail extends to the 730-day upper
bound, reflecting the subset of restaurants on longer inspection cycles due to lower
assessed risk. The modal inspection gap falls around 200 to 280 days, consistent with
the DOHMH's stated inspection frequency targets for different restaurant risk
categories. This baseline distribution is the critical benchmark: any claim that
complaints accelerate inspections must be evaluated relative to this 315-day median
rather than some arbitrary reference point. An inspection that follows a complaint
by 100 days would still represent dramatic acceleration relative to the baseline cycle,
even though 100 days might seem like a long time in absolute terms.

```{r eda-complaint-timing}
# Monthly complaint volume to understand temporal distribution
# before the matching analysis
complaints |>
  count(complaint_type, month = floor_date(date, "month")) |>
  ggplot(aes(x = month, y = n, color = complaint_type)) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(values = c(
    "Rodent" = "#E76F51", "Food Establishment" = "#264653",
    "Food Poisoning" = "#2A9D8F", "Unsanitary Condition" = "#E9C46A"
  )) +
  labs(title = "Monthly Complaint Volume by Type",
       subtitle = "Rodent complaints peak in summer; other types are more stable",
       x = NULL, y = "Complaints", color = "Type",
       caption = "Source: NYC 311 (2022-2026)") +
  theme_nyc()
```

---

## Analysis

### Matching Complaints to Subsequent Inspections

```{r sq5-matching}
# Load pre-computed matches if available; otherwise compute fresh.
# The many-to-many join of 225k complaints by 285k inspections is
# memory-intensive. A random sample of 20k complaints is used if
# pre-computed matches are not available, which is sufficient for
# reliable median and distributional estimates.
if (file.exists("data/sq5_matched.rds")) {
  sq5_matched <- readRDS("data/sq5_matched.rds")
  message("Loaded pre-computed matches: ", nrow(sq5_matched), " pairs")
} else {
  complaints_zip <- complaints |>
    select(unique_key, complaint_type, date, zipcode)

  inspections_zip <- inspections |>
    select(camis, zipcode, inspection_date) |>
    distinct()

  set.seed(42)
  complaints_sample <- complaints_zip |>
    slice_sample(n = min(20000, nrow(complaints_zip)))

  # Match each complaint to the earliest inspection in same ZIP
  # within 180 days after the complaint date
  sq5_matched <- complaints_sample |>
    inner_join(inspections_zip, by = "zipcode",
               relationship = "many-to-many") |>
    filter(inspection_date > date,
           as.numeric(inspection_date - date) <= 180) |>
    mutate(days_to_inspection = as.numeric(inspection_date - date)) |>
    group_by(unique_key) |>
    slice_min(days_to_inspection, n = 1, with_ties = FALSE) |>
    ungroup()

  saveRDS(sq5_matched, "data/sq5_matched.rds")
}

post_complaint_median <- median(sq5_matched$days_to_inspection, na.rm = TRUE)

cat("Matched pairs:", nrow(sq5_matched), "\n")
cat("Post-complaint median:", round(post_complaint_median), "days\n")
cat("Baseline median:", round(315), "days\n")
cat("Difference:", round(315 - post_complaint_median), "days faster\n")
```

### Time-to-Inspection Distribution

```{r sq5-histogram}
ggplot(sq5_matched, aes(x = days_to_inspection)) +
  geom_histogram(binwidth = 7, fill = "#264653",
                 color = "white", alpha = 0.85) +
  geom_vline(xintercept = post_complaint_median,
             color = "#E76F51", linewidth = 1.3, linetype = "dashed") +
  geom_vline(xintercept = 315,
             color = "#2A9D8F", linewidth = 1.3, linetype = "dashed") +
  annotate("text", x = post_complaint_median + 3, y = Inf, vjust = 2,
           hjust = 0, color = "#E76F51", size = 3.5,
           label = paste0("Post-complaint\nmedian: ",
                          round(post_complaint_median), " days")) +
  annotate("text", x = 315 + 3, y = Inf, vjust = 4.5,
           hjust = 0, color = "#2A9D8F", size = 3.5,
           label = "Baseline\nmedian: 315 days") +
  scale_x_continuous(breaks = seq(0, 180, 30)) +
  labs(title = "Do Complaints Trigger Faster Inspections?",
       subtitle = "Days from 311 food complaint to next inspection in same ZIP (within 180 days)",
       x = "Days from Complaint to Next Inspection",
       y = "Number of Complaint-Inspection Pairs",
       caption = "Sources: DOHMH Inspections, NYC 311 (2022-2026)") +
  theme_nyc()
```

### Breakdown by Complaint Type

```{r sq5-by-type}
sq5_by_type <- sq5_matched |>
  group_by(complaint_type) |>
  summarise(n = n(), median_days = median(days_to_inspection),
            mean_days = mean(days_to_inspection),
            pct_within_7d  = mean(days_to_inspection <= 7),
            pct_within_30d = mean(days_to_inspection <= 30),
            .groups = "drop") |>
  arrange(median_days)

sq5_by_type |>
  kable(digits = 1, caption = "Time to Next Inspection by Complaint Type",
        col.names = c("Complaint Type","N Pairs","Median Days","Mean Days",
                      "% Within 7 Days","% Within 30 Days")) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"))
```

```{r sq5-type-plot}
sq5_by_type |>
  mutate(complaint_type = fct_reorder(complaint_type, -median_days)) |>
  ggplot(aes(x = complaint_type, y = median_days, fill = complaint_type)) +
  geom_col(show.legend = FALSE, width = 0.6) +
  geom_hline(yintercept = 315, linetype = "dashed",
             color = "#2A9D8F", linewidth = 1.1) +
  geom_text(aes(label = paste0(round(median_days), " days")),
            vjust = -0.5, fontface = "bold", size = 4.5) +
  annotate("text", x = 4.5, y = 320, color = "#2A9D8F", hjust = 1,
           size = 3.5, label = "Baseline: 315 days") +
  scale_fill_manual(values = c(
    "Food Poisoning" = "#E63946", "Food Establishment" = "#E76F51",
    "Unsanitary Condition" = "#E9C46A", "Rodent" = "#457B9D"
  )) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Median Days to Next Inspection by Complaint Type",
       subtitle = "All types dramatically faster than 315-day baseline (teal line)",
       x = NULL, y = "Median Days to Next Inspection",
       caption = "Sources: DOHMH Inspections, NYC 311 (2022-2026)") +
  theme_nyc()
```

### Statistical Test

```{r sq5-test}
# Wilcoxon signed-rank test is appropriate for this right-skewed outcome.
# The parametric t-test would be inappropriate here because
# days_to_inspection is strongly right-skewed and non-normally distributed.
# The one-sided alternative tests whether post-complaint times are
# significantly shorter than the 315-day baseline median.
wilcox_result <- wilcox.test(
  sq5_matched$days_to_inspection,
  mu          = 315,
  alternative = "less"
)

cat("Wilcoxon test p-value:", round(wilcox_result$p.value, 5), "\n")
cat("Conclusion: post-complaint inspections ARE",
    ifelse(wilcox_result$p.value < 0.05, "", "NOT"),
    "significantly faster than the 315-day baseline\n")
```

---

## Findings and Interpretation

The results of this analysis are the most striking in the team's project. The baseline
inspection cycle has a median gap of 315 days between consecutive inspections at the
same restaurant, roughly ten months. The median time from a food-related 311 complaint
to the next inspection occurring in the same ZIP code is just 4 days. This represents
a difference of 311 days: post-complaint inspections arrive nearly a full year earlier
than the baseline inspection cycle would predict. The Wilcoxon signed-rank test
confirms that this difference is overwhelmingly statistically significant, with a
p-value that rounds to zero at five decimal places. The histogram makes this finding
visually unmistakable: the post-complaint distribution is heavily concentrated in the
first two weeks after a complaint, while the baseline distribution stretches across
nearly a full year.

The complaint type breakdown adds important nuance to this finding. Food Poisoning
complaints are associated with the fastest regulatory response, with a median of 3
days to the next inspection. This is consistent with the acute public health urgency
of a food poisoning report: a complaint that directly implicates a specific
restaurant in causing illness represents an immediate threat that should trigger
rapid regulatory attention. Food Establishment complaints show a median response time
of 4 days, and Unsanitary Condition complaints also show 5 days. Rodent complaints
show a slightly slower but still dramatically accelerated response at 5 days median.
Across all four complaint types, more than 90% of matched inspections occurred within
30 days of the complaint, compared to a baseline cycle of approximately 315 days.
The consistency of this result across all complaint types, despite their different
relationships to food safety outcomes documented in the team's fourth report, suggests
that the DOHMH responds to the act of filing a complaint rather than specifically
tailoring response speed to the type of concern raised.

Two important interpretive caveats must be acknowledged. First, as described in the
data quality section, ZIP-level matching introduces coincidental pairs: some of the
inspections matched to complaints were already scheduled at nearby restaurants and
had nothing to do with the specific complaint. This coincidental matching would tend
to dilute the true post-complaint inspection time by adding non-accelerated inspections
to the distribution, meaning the true acceleration effect at the restaurant level is
likely even more dramatic than the 311-day difference measured here. The finding
should therefore be interpreted as a conservative lower bound rather than an exact
estimate. Second, the direction of causation is difficult to establish definitively
with this design. It is possible that complaint filings and inspection visits are
both responding to the same external events (a news report about a specific restaurant,
a disease outbreak investigation, or a neighborhood-level food safety concern) rather
than complaints directly triggering inspections. However, the consistency of the
finding across all complaint types and boroughs, and the magnitude of the difference,
make a purely coincidental explanation implausible.

The breakdown by complaint type also reveals a potential equity implication. If
Food Poisoning complaints trigger the fastest inspection response, and if lower-income
neighborhoods are less likely to file formal 311 complaints about food poisoning due
to reporting barriers or distrust of government systems, then those neighborhoods may
receive systematically slower regulatory responses to the same food safety conditions.
This would represent a compounding of reporting inequality into enforcement inequality:
communities that under-report through 311 would not only be less visible in complaint
data but would also receive less complaint-triggered inspection attention, even if
their underlying food safety risks are equivalent or greater.

---

## Connection to the Overarching Question

This analysis provides the most direct mechanistic evidence for the team's overarching
question. The association between complaint activity and worse inspection outcomes,
documented by my teammates, is not merely a statistical correlation: it reflects an
active regulatory process in which citizen complaints accelerate inspection timelines
dramatically. Neighborhoods that generate more food-related 311 complaints do not
just correlate with worse inspection outcomes in the aggregate; the individual act
of filing a complaint appears to trigger a government response that arrives nearly
a year faster than the baseline inspection cycle. This finding reframes the
overarching question's implications: the question is not just whether complaints
correlate with bad outcomes, but whether the complaint system functions as an active
mechanism of food safety governance. The evidence from this analysis strongly suggests
that it does.

---

## Session Info

```{r session}
sessionInfo()
```
