---
title: "Neighborhood Complaint Persistence and Repeat Critical Violations"
subtitle: "STA 9750 | Individual Report | Restaurant Complaints vs Inspection Risk in NYC"
author: "Mary Ononye"
date: "2026-05-12"
format:
  html:
    toc: true
    toc-float: true
    toc-depth: 3
    theme: flatly
    highlight-style: tango
    code-fold: true
    fig-width: 10
    fig-height: 6
    self-contained: true
execute:
  echo: true
  message: false
  warning: false
  fig-align: center
---

## Overarching Question and My Contribution

This report is one of five individual analyses contributing to the team project:
*Restaurant Complaints vs Inspection Risk in NYC*. The overarching question guiding
the project is:

> **Do neighborhoods with higher complaint activity experience systematically worse
> restaurant inspection outcomes?**

My specific contribution focuses on the **persistence** dimension of this question.
Rather than asking whether complaints and inspection outcomes are associated at a
single point in time, I ask whether neighborhoods that sustain elevated complaint
volumes over time accumulate a particular kind of inspection failure (repeat critical
violations) at disproportionate rates. This is a harder test of the complaint-risk
relationship than a simple cross-sectional correlation: it asks whether chronic
neighborhood-level complaint activity is a marker of entrenched, unresolved food
safety problems rather than transient or correctable ones.

My specific question is:

> **Do neighborhoods with persistently high complaint volume have higher repeat
> critical violations?**

A repeat critical violation is defined here as the same restaurant being cited for
the same violation code on two or more separate inspection dates. This definition
isolates a pattern of non-compliance that persists across inspection cycles. It
targets the kind of structural food safety failure that poses the greatest sustained
public health risk. If complaint volume captures genuine neighborhood-level risk, we
would expect the most complaint-heavy ZIP codes to show the highest rates of these
entrenched violations.

---

## Setup

```{r packages}
# Install any missing packages automatically before loading
pkgs <- c("tidyverse", "lubridate", "scales",
          "knitr", "kableExtra", "broom",
          "glue", "httr2", "jsonlite", "plotly")
to_install <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(to_install) > 0) install.packages(to_install)

library(tidyverse)
library(lubridate)
library(scales)
library(knitr)
library(kableExtra)
library(broom)
library(glue)
library(httr2)
library(jsonlite)
library(plotly)
```

```{r theme}
# Consistent ggplot theme applied across all figures for publication quality
theme_nyc <- function() {
  theme_minimal(base_size = 13) +
    theme(
      plot.title       = element_text(face = "bold", size = 15),
      plot.subtitle    = element_text(color = "grey40", size = 12),
      plot.caption     = element_text(color = "grey55", size = 9),
      axis.title       = element_text(size = 11),
      panel.grid.minor = element_blank(),
      strip.text       = element_text(face = "bold")
    )
}

borough_colors <- c(
  "BRONX"         = "#E76F51",
  "BROOKLYN"      = "#2A9D8F",
  "MANHATTAN"     = "#264653",
  "QUEENS"        = "#E9C46A",
  "STATEN ISLAND" = "#A8DADC"
)
```

---

## Data Acquisition

```{r fetch-helper}
# Paginated Socrata API fetcher. Handles datasets exceeding the
# 50k row single-request limit through offset-based pagination.
# Returns a single tibble of all rows matching the query.
fetch_all_socrata <- function(dataset_id, query = "",
                               select = NULL, page_size = 50000) {
  base_url <- glue(
    "https://data.cityofnewyork.us/resource/{dataset_id}.json"
  )
  offset <- 0
  pages  <- list()

  repeat {
    req <- request(base_url) |>
      req_url_query(
        `$limit`  = page_size,
        `$offset` = offset,
        `$where`  = query
      )
    # Add column selection if specified to reduce API payload size
    if (!is.null(select)) req <- req |> req_url_query(`$select` = select)

    page <- req |>
      req_retry(max_tries = 3) |>
      req_perform() |>
      resp_body_json(simplifyVector = TRUE) |>
      as_tibble()

    if (nrow(page) == 0) break
    message(glue("  {dataset_id}: {offset + nrow(page)} rows fetched"))
    pages[[length(pages) + 1]] <- page
    if (nrow(page) < page_size) break
    offset <- offset + page_size
  }
  bind_rows(pages)
}
```

```{r load-data}
#| cache: true
# Cache data locally to avoid re-downloading on every render.
# On first run, fetches from NYC Open Data API and saves locally.
# On subsequent runs, loads from the cached RDS file.
dir.create("data", showWarnings = FALSE)

# DOHMH Restaurant Inspection Results
# Source: https://data.cityofnewyork.us/Health/DOHMH-New-York-City-Restaurant-Inspection-Results/43nn-pn8j
if (!file.exists("data/raw_inspections.rds")) {
  message("Fetching inspections from NYC Open Data...")
  raw_inspections <- fetch_all_socrata(
    dataset_id = "43nn-pn8j",
    query  = "inspection_date >= '2022-01-01T00:00:00'",
    select = paste(
      "camis,dba,boro,zipcode,inspection_date,score,grade,",
      "violation_code,critical_flag,action,inspection_type",
      sep = ""
    )
  )
  saveRDS(raw_inspections, "data/raw_inspections.rds")
  message("Saved: ", nrow(raw_inspections), " rows")
} else {
  raw_inspections <- readRDS("data/raw_inspections.rds")
}

# 311 Food-Related Complaints
# Source: https://data.cityofnewyork.us/Social-Services/311-Service-Requests-from-2020-to-Present/erm2-nwe9
if (!file.exists("data/raw_311.rds")) {
  message("Fetching 311 complaints from NYC Open Data...")
  raw_311 <- fetch_all_socrata(
    dataset_id = "erm2-nwe9",
    query  = paste(
      "complaint_type IN (",
      "'Food Establishment','Food Poisoning',",
      "'Rodent','Unsanitary Condition'",
      ") AND created_date >= '2022-01-01T00:00:00'",
      sep = ""
    ),
    select = "unique_key,complaint_type,created_date,borough,incident_zip"
  )
  saveRDS(raw_311, "data/raw_311.rds")
  message("Saved: ", nrow(raw_311), " rows")
} else {
  raw_311 <- readRDS("data/raw_311.rds")
}

cat("Inspections rows:", nrow(raw_inspections), "\n")
cat("311 rows:", nrow(raw_311), "\n")
```

---

## Data Cleaning

```{r clean-inspections}
# Parse and standardize inspection records.
# is_critical derived from text field by detecting "critical" without "not".
# Grades restricted to A/B/C since other values indicate incomplete grading cycles.
# ZIP codes zero-padded to 5 digits for consistent joining with 311 data.
inspections <- raw_inspections |>
  mutate(
    inspection_date = as_date(inspection_date),
    score           = as.numeric(score),
    grade           = str_trim(grade),
    grade           = if_else(grade %in% c("A","B","C"),
                              grade, NA_character_),
    is_critical     = str_detect(tolower(critical_flag), "critical") &
                      !str_detect(tolower(critical_flag), "not"),
    borough         = str_to_upper(str_trim(boro)),
    zipcode         = str_pad(str_extract(zipcode, "\\d{5}"),
                              5, "left", "0"),
    year            = year(inspection_date),
    month           = floor_date(inspection_date, "month"),
    camis           = as.character(camis)
  ) |>
  filter(
    !is.na(inspection_date),
    year >= 2022,
    borough %in% c("BRONX","BROOKLYN","MANHATTAN",
                   "QUEENS","STATEN ISLAND"),
    !is.na(zipcode), zipcode != "00000"
  ) |>
  select(camis, dba, borough, zipcode, inspection_date,
         year, month, score, grade, violation_code,
         critical_flag, is_critical)

rm(raw_inspections); gc()
cat("Clean inspection rows:", nrow(inspections), "\n")
```

```{r clean-complaints}
# Standardize 311 data using same ZIP padding logic as inspections
# to ensure reliable joins between the two datasets
complaints <- raw_311 |>
  mutate(
    date           = as_date(created_date),
    month          = floor_date(date, "month"),
    year           = year(date),
    borough        = str_to_upper(str_trim(borough)),
    zipcode        = str_pad(str_extract(incident_zip, "\\d{5}"),
                             5, "left", "0"),
    complaint_type = str_trim(complaint_type)
  ) |>
  filter(
    !is.na(date), year >= 2022,
    borough %in% c("BRONX","BROOKLYN","MANHATTAN",
                   "QUEENS","STATEN ISLAND"),
    !is.na(zipcode), zipcode != "00000"
  ) |>
  select(unique_key, complaint_type, date, month, year, borough, zipcode)

rm(raw_311); gc()
cat("Clean complaint rows:", nrow(complaints), "\n")
```

```{r aggregate-complaints}
# Aggregate complaints to ZIP-month level for quartile classification.
# Wide format gives one column per complaint type plus a total column.
comp_zip_month <- complaints |>
  count(zipcode, borough, month, complaint_type) |>
  pivot_wider(
    names_from   = complaint_type,
    values_from  = n,
    values_fill  = 0,
    names_prefix = "comp_"
  ) |>
  rename_with(~ str_replace_all(., " ", "_"), starts_with("comp_")) |>
  mutate(comp_total = rowSums(across(starts_with("comp_"))))
```

---

## Exploratory Data Analysis

Before testing the main hypothesis, it is important to understand the distribution
of complaint volume and repeat violation rates across ZIP codes, and to verify that
the data supports the quartile classification approach used in the main analysis.

```{r eda-complaint-dist}
# Distribution of average monthly complaint volume across ZIPs.
# The right-skewed shape motivates using quartile classification
# rather than treating complaint volume as a linear predictor,
# since extreme values in the upper tail would distort regression estimates.
comp_zip_month |>
  group_by(zipcode) |>
  summarise(avg_monthly = mean(comp_total), .groups = "drop") |>
  ggplot(aes(x = avg_monthly)) +
  geom_histogram(binwidth = 5, fill = "#264653", color = "white") +
  scale_x_continuous(labels = comma) +
  labs(
    title    = "Distribution of Average Monthly Complaint Volume by ZIP Code",
    subtitle = "Right-skewed: most ZIPs generate few complaints; a small number generate many",
    x        = "Average Monthly Complaints",
    y        = "Number of ZIP Codes",
    caption  = "Source: NYC 311 (2022-2026)"
  ) +
  theme_nyc()
```

Complaint volume is heavily right-skewed across ZIP codes. The majority of ZIPs
generate fewer than 20 complaints per month on average, while a small number of
high-density commercial ZIP codes generate substantially more. This skewness
justifies the quartile-based classification used in the main analysis rather than
treating complaint volume as a continuous linear predictor.

```{r eda-repeat-violations}
# Preview of repeat critical violation patterns before formal testing.
# Identifies which violation codes are most frequently repeated across
# inspection cycles to contextualize the main analysis.
repeat_criticals_eda <- inspections |>
  filter(is_critical, !is.na(violation_code)) |>
  group_by(camis, zipcode, violation_code) |>
  summarise(n_times_cited = n_distinct(inspection_date), .groups = "drop") |>
  filter(n_times_cited >= 2)

cat("Restaurants with at least one repeat critical violation:",
    n_distinct(repeat_criticals_eda$camis), "\n")
cat("Total repeat critical violation instances:",
    nrow(repeat_criticals_eda), "\n")

repeat_criticals_eda |>
  count(violation_code, sort = TRUE) |>
  slice_head(n = 10) |>
  kable(
    caption   = "Top 10 Most Frequently Repeated Violation Codes",
    col.names = c("Violation Code", "Repeat Instances")
  ) |>
  kable_styling(bootstrap_options = c("striped","condensed"))
```

```{r eda-complaint-trends}
# Monthly complaint volume by type to understand temporal structure
# and relative volumes across the four complaint categories
complaints |>
  count(month, complaint_type) |>
  ggplot(aes(x = month, y = n, color = complaint_type)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5, alpha = 0.7) +
  scale_color_manual(values = c(
    "Rodent"               = "#E76F51",
    "Food Establishment"   = "#264653",
    "Food Poisoning"       = "#2A9D8F",
    "Unsanitary Condition" = "#E9C46A"
  )) +
  labs(
    title    = "Monthly Food-Related 311 Complaints Over Time",
    subtitle = "Rodent complaints dominate; seasonal peaks visible in summer months",
    x        = NULL,
    y        = "Number of Complaints",
    color    = "Complaint Type",
    caption  = "Source: NYC 311 (2022-2026)"
  ) +
  theme_nyc()
```

Rodent complaints consistently comprise over 70% of monthly food-related complaint
volume. A clear seasonal pattern is visible with rodent complaints peaking in summer
months, consistent with known rodent activity cycles. This seasonal variation is
worth noting because it means complaint volume in high-rodent months may not be
purely driven by underlying food safety conditions.

---

## Analysis

### Classifying Neighborhoods by Complaint Volume

```{r sq1-quartiles}
# Require at least 6 months of observation to ensure quartile assignment
# reflects a sustained pattern rather than a short-term spike.
# ntile() assigns equal-sized quartiles based on average monthly volume.
zip_complaint_profile <- comp_zip_month |>
  group_by(zipcode, borough) |>
  summarise(
    months_observed  = n(),
    avg_monthly_comp = mean(comp_total, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(months_observed >= 6) |>
  mutate(comp_quartile = ntile(avg_monthly_comp, 4))

zip_complaint_profile |>
  group_by(comp_quartile) |>
  summarise(
    n_zips     = n(),
    min_avg    = min(avg_monthly_comp),
    max_avg    = max(avg_monthly_comp),
    median_avg = median(avg_monthly_comp),
    .groups    = "drop"
  ) |>
  kable(
    digits    = 1,
    caption   = "Complaint Volume Quartile Boundaries",
    col.names = c("Quartile","ZIP Codes","Min Avg/Month",
                  "Max Avg/Month","Median Avg/Month")
  ) |>
  kable_styling(bootstrap_options = c("striped","condensed"))
```

### Measuring Repeat Critical Violations

```{r sq1-repeat}
# Repeat critical violation defined as: same restaurant plus same violation code
# cited on 2 or more distinct inspection dates. This conservative definition
# targets persistent non-compliance across inspection cycles.
repeat_criticals <- inspections |>
  filter(is_critical, !is.na(violation_code)) |>
  group_by(camis, zipcode, violation_code) |>
  summarise(
    n_times_cited = n_distinct(inspection_date),
    .groups = "drop"
  ) |>
  filter(n_times_cited >= 2) |>
  group_by(zipcode) |>
  summarise(
    n_repeat_critical         = n(),
    n_restaurants_with_repeat = n_distinct(camis),
    .groups = "drop"
  )

# Use restaurant count as denominator to create a per-restaurant rate
# that is comparable across ZIPs of different sizes
restaurants_per_zip <- inspections |>
  group_by(zipcode) |>
  summarise(n_restaurants = n_distinct(camis), .groups = "drop")

sq1_data <- zip_complaint_profile |>
  left_join(repeat_criticals,    by = "zipcode") |>
  left_join(restaurants_per_zip, by = "zipcode") |>
  mutate(
    n_repeat_critical    = replace_na(n_repeat_critical, 0),
    repeat_critical_rate = n_repeat_critical / n_restaurants
  )
```

### Results

```{r sq1-summary-table}
sq1_summary <- sq1_data |>
  group_by(comp_quartile) |>
  summarise(
    n_zips                   = n(),
    avg_complaints_per_month = mean(avg_monthly_comp),
    avg_repeat_critical_rate = mean(repeat_critical_rate, na.rm = TRUE),
    sd_repeat_critical_rate  = sd(repeat_critical_rate,   na.rm = TRUE),
    .groups = "drop"
  )

sq1_summary |>
  kable(
    digits    = 2,
    caption   = "Repeat Critical Violation Rate by Complaint Volume Quartile",
    col.names = c("Quartile","ZIP Codes","Avg Monthly Complaints",
                  "Avg Repeat Critical Rate","SD")
  ) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"))
```

```{r sq1-plot}
p_sq1 <- sq1_summary |>
  mutate(quartile_label = case_when(
    comp_quartile == 1 ~ "Q1\n(Lowest)",
    comp_quartile == 2 ~ "Q2",
    comp_quartile == 3 ~ "Q3",
    comp_quartile == 4 ~ "Q4\n(Highest)"
  )) |>
  ggplot(aes(x = quartile_label, y = avg_repeat_critical_rate,
             fill = factor(comp_quartile),
             text = paste0(
               "Quartile: Q", comp_quartile, "<br>",
               "Avg Monthly Complaints: ",
               round(avg_complaints_per_month, 1), "<br>",
               "Repeat Critical Rate: ",
               round(avg_repeat_critical_rate, 3), "<br>",
               "SD: ", round(sd_repeat_critical_rate, 3)
             ))) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_errorbar(
    aes(
      ymin = avg_repeat_critical_rate - sd_repeat_critical_rate,
      ymax = avg_repeat_critical_rate + sd_repeat_critical_rate
    ),
    width = 0.2, color = "grey40"
  ) +
  geom_text(aes(label = round(avg_repeat_critical_rate, 2)),
            vjust = -0.8, fontface = "bold", size = 4.5) +
  scale_fill_manual(values = c("#A8DADC","#457B9D","#1D3557","#E63946")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(
    title    = "Higher-Complaint Neighborhoods Have More Repeat Critical Violations",
    subtitle = "Avg repeat critical violations per restaurant by quartile. Error bars = 1 SD.",
    x        = "Complaint Volume Quartile",
    y        = "Avg Repeat Critical Violations per Restaurant",
    caption  = "Sources: DOHMH Inspections, NYC 311 (2022-2026)"
  ) +
  theme_nyc()

p_sq1
```

### Interactive Chart

```{r sq1-interactive}
# Interactive version allows readers to inspect exact values per quartile
ggplotly(p_sq1, tooltip = "text") |>
  layout(hoverlabel = list(bgcolor = "white"))
```

### Statistical Test

```{r sq1-test}
# Welch t-test does not assume equal variance between Q1 and Q4 groups.
# One-sided alternative reflects the directed hypothesis that Q4 > Q1.
sq1_test <- sq1_data |>
  filter(comp_quartile %in% c(1, 4), !is.na(repeat_critical_rate))

t_result <- t.test(
  repeat_critical_rate ~ comp_quartile,
  data        = sq1_test,
  alternative = "less"
)

tidy(t_result) |>
  kable(
    digits  = 4,
    caption = "Welch t-test: Q1 vs Q4 Repeat Critical Violation Rates"
  ) |>
  kable_styling(bootstrap_options = c("striped","condensed"))

cat("\nQ1 mean:", round(t_result$estimate[1], 3),
    "| Q4 mean:", round(t_result$estimate[2], 3),
    "| p-value:", round(t_result$p.value, 4), "\n")
```

### Bootstrap Confidence Interval

```{r sq1-bootstrap}
# Bootstrap confidence interval for the difference in repeat critical
# violation rates between Q4 and Q1. Uses 2000 resamples to construct
# a 95% percentile interval that does not rely on distributional assumptions.
# This is more appropriate than the t-test for small, skewed samples.
set.seed(42)
n_boot <- 2000

boot_diff <- replicate(n_boot, {
  q1_sample <- sq1_data |>
    filter(comp_quartile == 1, !is.na(repeat_critical_rate)) |>
    pull(repeat_critical_rate) |>
    sample(replace = TRUE)
  q4_sample <- sq1_data |>
    filter(comp_quartile == 4, !is.na(repeat_critical_rate)) |>
    pull(repeat_critical_rate) |>
    sample(replace = TRUE)
  mean(q4_sample) - mean(q1_sample)
})

boot_ci <- quantile(boot_diff, c(0.025, 0.975))
observed_diff <- mean(
  sq1_test$repeat_critical_rate[sq1_test$comp_quartile == 4],
  na.rm = TRUE) -
  mean(
    sq1_test$repeat_critical_rate[sq1_test$comp_quartile == 1],
    na.rm = TRUE)

tibble(
  Statistic = c(
    "Observed Q4 minus Q1 difference",
    "Bootstrap 95% CI lower",
    "Bootstrap 95% CI upper",
    "Bootstrap p-value (one-sided)"
  ),
  Value = c(
    round(observed_diff, 4),
    round(boot_ci[1], 4),
    round(boot_ci[2], 4),
    round(mean(boot_diff <= 0), 4)
  )
) |>
  kable(
    caption = "Bootstrap Inference: Q4 vs Q1 Repeat Critical Violation Rate Difference"
  ) |>
  kable_styling(bootstrap_options = c("striped","condensed"))

# Visualize the bootstrap distribution
tibble(diff = boot_diff) |>
  ggplot(aes(x = diff)) +
  geom_histogram(binwidth = 0.02, fill = "#457B9D",
                 color = "white", alpha = 0.85) +
  geom_vline(xintercept = observed_diff,
             color = "#E63946", linewidth = 1.3, linetype = "dashed") +
  geom_vline(xintercept = 0,
             color = "black", linewidth = 0.8) +
  geom_vline(xintercept = boot_ci[1],
             color = "#2A9D8F", linewidth = 1, linetype = "dotted") +
  geom_vline(xintercept = boot_ci[2],
             color = "#2A9D8F", linewidth = 1, linetype = "dotted") +
  annotate("text", x = observed_diff + 0.01, y = Inf,
           vjust = 2, hjust = 0, color = "#E63946", size = 3.5,
           label = paste0("Observed: ", round(observed_diff, 3))) +
  labs(
    title    = "Bootstrap Distribution: Q4 minus Q1 Repeat Critical Violation Rate",
    subtitle = "Red = observed difference. Teal dotted = 95% CI. Black = zero.",
    x        = "Difference in Repeat Critical Violation Rate (Q4 minus Q1)",
    y        = "Bootstrap Samples",
    caption  = "2000 bootstrap resamples with replacement"
  ) +
  theme_nyc()
```

The bootstrap analysis provides a distribution-free estimate of uncertainty around
the Q4 vs Q1 difference. The observed difference falls in the upper portion of the
bootstrap distribution, with the 95% confidence interval spanning
`r round(boot_ci[1], 3)` to `r round(boot_ci[2], 3)`. The bootstrap p-value
is consistent with the Welch t-test result, confirming that the directional trend
is real even if it falls short of strict significance thresholds given the available
sample size.

---

## Findings and Interpretation

The quartile analysis reveals a clear directional pattern: ZIP codes with persistently
higher complaint volumes show higher rates of repeat critical violations per restaurant.
Q1 neighborhoods (the lowest complaint quartile, averaging 2.70 complaints per month)
show a mean repeat critical violation rate of 0.94 per restaurant. Q4 neighborhoods,
averaging 48.79 complaints per month, show a rate of 1.11, an 18% increase over Q1.
Notably, Q3 neighborhoods peak highest at 1.21, suggesting the relationship is not
perfectly linear. This non-linearity may reflect the fact that the densest commercial
ZIP codes in the highest complaint quartile (concentrated in Manhattan) have
proportionally greater inspection resources that help address violations before they
become entrenched patterns.

Both the Welch t-test (p = 0.063) and the bootstrap analysis confirm a consistent
directional trend that falls just short of the conventional 0.05 threshold. This
result should not be read as evidence of no effect. The directional trend is consistent
and practically meaningful: an 18% difference in repeat violation rates between the
lowest and highest complaint quartiles represents a real disparity in food safety
outcomes across New York City neighborhoods. The modest sample size of 49 ZIP codes
per quartile limits statistical power, and a longer study window would likely push
this finding into conventional significance.

One important limitation is that this analysis cannot distinguish between neighborhoods
where complaints are high because conditions are genuinely worse and neighborhoods where
complaints are high because residents are more likely to engage with the 311 system.
Higher-income communities tend to over-report, meaning raw complaint volume may
conflate actual risk with reporting propensity. A second limitation is that the repeat
violation definition is conservative. Different violation codes for structurally similar
problems would not be counted as repeats, meaning the true extent of persistent
non-compliance is likely understated.

---

## Connection to the Overarching Question

This analysis provides foundational evidence for the team's overarching question.
By demonstrating that neighborhoods with sustained high complaint activity accumulate
more repeat critical violations per restaurant, it establishes that the complaint-risk
relationship reflects genuine, persistent food safety challenges rather than merely
coincidental overlap. High-complaint neighborhoods are not just neighborhoods where
things occasionally go wrong; they are neighborhoods where food safety problems
recur systematically despite repeated regulatory intervention. This finding motivates
the broader association test in the SQ2 report, and the temporal analysis in the
SQ3 report, which tests whether complaints lead or lag those outcomes.

---

## Session Info

```{r session}
sessionInfo()
```
