---
title: "Complaint Type Specificity and Critical Violation Rates"
subtitle: "STA 9750 | Individual Report | Restaurant Complaints vs Inspection Risk in NYC"
author: "Group Member 4"
date: "2026-05-12"
format:
  html:
    toc: true
    toc-float: true
    toc-depth: 3
    theme: flatly
    highlight-style: tango
    code-fold: true
    fig-width: 10
    fig-height: 6
    self-contained: true
execute:
  echo: true
  message: false
  warning: false
  fig-align: center
---

## Overarching Question and My Contribution

This report is one of five individual analyses contributing to the team project
*Restaurant Complaints vs Inspection Risk in NYC*. The project's overarching question
asks whether neighborhoods with higher complaint activity experience systematically
worse restaurant inspection outcomes. My teammates have established that the answer
is yes in both cross-sectional and persistence terms, and that the relationship is
contemporaneous rather than predictive. My contribution refines this finding by
asking a more precise question: are all types of food-related 311 complaints equally
informative about food safety risk, or do certain complaint categories carry
substantially stronger signals about underlying violations?

This question matters for both scientific and practical reasons. On the scientific
side, understanding which complaint types most strongly predict critical violations
tells us something about the mechanisms connecting citizen observations to regulatory
outcomes. If rodent complaints are the strongest predictor, it suggests that visible
pest activity (something residents can directly observe from outside an establishment)
is the most reliable signal of internal food safety failure. If food establishment
complaints are strongest, it suggests that broader, more diffuse citizen concerns
about restaurant conditions are more informative than specific observable events.
On the practical side, if certain complaint types carry stronger signals, regulators
could develop a weighted complaint scoring system that prioritizes inspection
responses to the most predictive complaint categories, improving the efficiency of
enforcement resource allocation without requiring any new data collection.

The 311 dataset used in this project contains four food-related complaint categories:
Rodent, Food Establishment, Food Poisoning, and Unsanitary Condition. These categories
differ substantially in their volume, specificity, and relationship to the regulatory
framework. Rodent complaints, the most numerous at 160,737 over the study period,
reflect visible pest activity that residents can observe directly. Food Establishment
complaints (50,604) are a broad catch-all category that encompasses general concerns
about restaurant conditions. Food Poisoning complaints (12,586) reflect the most
direct kind of harm: a resident who believes they became ill after eating at a
restaurant. Unsanitary Condition complaints (1,263) are the rarest and most
heterogeneous category, covering cleanliness concerns that may or may not be
restaurant-specific.

My specific question is which types of food-related complaints are most strongly
associated with higher neighborhood-level critical violation rates. My hypothesis
entering the analysis is that Rodent complaints will show the strongest correlation
with critical violations, for two reasons. First, rodent presence is itself a
critical violation category under DOHMH's inspection rubric, meaning that a
confirmed rodent infestation that generates citizen complaints will likely also
generate critical violation citations when an inspector visits. Second, rodent
complaints are the most objectively verifiable type: a resident who reports seeing a
rat or mouse near a restaurant is reporting something specific and observable rather
than a general subjective concern about the establishment's cleanliness.

---

## Setup

```{r packages}
pkgs <- c("tidyverse", "lubridate", "scales", "knitr",
          "kableExtra", "broom", "glue", "httr2", "jsonlite")
to_install <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(to_install) > 0) install.packages(to_install)

library(tidyverse)
library(lubridate)
library(scales)
library(knitr)
library(kableExtra)
library(broom)
library(glue)
library(httr2)
library(jsonlite)
```

```{r theme}
theme_nyc <- function() {
  theme_minimal(base_size = 13) +
    theme(plot.title = element_text(face = "bold", size = 15),
          plot.subtitle = element_text(color = "grey40", size = 12),
          plot.caption = element_text(color = "grey55", size = 9),
          axis.title = element_text(size = 11),
          panel.grid.minor = element_blank())
}
```

---

## Data Acquisition

```{r fetch-helper}
fetch_all_socrata <- function(dataset_id, query = "",
                               select = NULL, page_size = 50000) {
  base_url <- glue("https://data.cityofnewyork.us/resource/{dataset_id}.json")
  offset <- 0; pages <- list()
  repeat {
    req <- request(base_url) |>
      req_url_query(`$limit` = page_size, `$offset` = offset, `$where` = query)
    if (!is.null(select)) req <- req |> req_url_query(`$select` = select)
    page <- req |> req_retry(max_tries = 3) |> req_perform() |>
      resp_body_json(simplifyVector = TRUE) |> as_tibble()
    if (nrow(page) == 0) break
    message(glue("  {dataset_id}: {offset + nrow(page)} rows"))
    pages[[length(pages) + 1]] <- page
    if (nrow(page) < page_size) break
    offset <- offset + page_size
  }
  bind_rows(pages)
}
```

```{r load-data}
#| cache: true
dir.create("data", showWarnings = FALSE)

if (!file.exists("data/raw_inspections.rds")) {
  raw_inspections <- fetch_all_socrata(
    "43nn-pn8j", query = "inspection_date >= '2022-01-01T00:00:00'",
    select = "camis,boro,zipcode,inspection_date,score,grade,critical_flag"
  )
  saveRDS(raw_inspections, "data/raw_inspections.rds")
} else { raw_inspections <- readRDS("data/raw_inspections.rds") }

if (!file.exists("data/raw_311.rds")) {
  raw_311 <- fetch_all_socrata(
    "erm2-nwe9",
    query = paste("complaint_type IN ('Food Establishment','Food Poisoning',",
                  "'Rodent','Unsanitary Condition')",
                  "AND created_date >= '2022-01-01T00:00:00'"),
    select = "unique_key,complaint_type,created_date,borough,incident_zip"
  )
  saveRDS(raw_311, "data/raw_311.rds")
} else { raw_311 <- readRDS("data/raw_311.rds") }
```

---

## Data Cleaning

```{r clean}
inspections <- raw_inspections |>
  mutate(
    inspection_date = as_date(inspection_date),
    score           = as.numeric(score),
    grade           = str_trim(grade),
    grade           = if_else(grade %in% c("A","B","C"), grade, NA_character_),
    is_critical     = str_detect(tolower(critical_flag), "critical") &
                      !str_detect(tolower(critical_flag), "not"),
    borough         = str_to_upper(str_trim(boro)),
    zipcode         = str_pad(str_extract(zipcode, "\\d{5}"), 5, "left", "0"),
    year            = year(inspection_date)
  ) |>
  filter(!is.na(inspection_date), year >= 2022,
         borough %in% c("BRONX","BROOKLYN","MANHATTAN",
                        "QUEENS","STATEN ISLAND"),
         !is.na(zipcode), zipcode != "00000")

complaints <- raw_311 |>
  mutate(
    date           = as_date(created_date),
    year           = year(date),
    borough        = str_to_upper(str_trim(borough)),
    zipcode        = str_pad(str_extract(incident_zip, "\\d{5}"), 5, "left", "0"),
    complaint_type = str_trim(complaint_type)
  ) |>
  filter(!is.na(date), year >= 2022,
         borough %in% c("BRONX","BROOKLYN","MANHATTAN",
                        "QUEENS","STATEN ISLAND"),
         !is.na(zipcode), zipcode != "00000")

rm(raw_inspections, raw_311); gc()
```

---

## Data Quality Assessment

The complaint type analysis faces a specific data quality challenge that does not
affect the aggregate analyses: the unequal sample sizes across complaint categories
create different levels of statistical reliability for each correlation estimate.
With 160,737 Rodent complaints distributed across 183 ZIP codes, the Rodent
correlation estimate is based on a large, stable sample. With only 1,263 Unsanitary
Condition complaints across 146 ZIP codes, the Unsanitary Condition correlation
estimate is much noisier and more sensitive to individual observations. This unequal
reliability should be kept in mind when interpreting the relative magnitudes of the
four correlation coefficients.

A second quality consideration is the specificity of each complaint category relative
to restaurants specifically. All four complaint types were filtered from the 311
dataset based on their category label, but not all complaints within each category
necessarily relate to a restaurant. Rodent complaints, for example, may refer to
rodent activity in a residential building, a street, or a park rather than a
restaurant. The 311 data does not contain a field that reliably identifies whether a
complaint was filed about a food service establishment, though the Food Establishment
and Food Poisoning categories are more likely to be restaurant-specific than Rodent
and Unsanitary Condition. This differential specificity creates a potential source
of attenuation bias: if Rodent complaints are less restaurant-specific than Food
Establishment complaints, their correlation with restaurant inspection outcomes would
be attenuated relative to the true underlying relationship. This may explain why
rodent complaints, despite being directly tied to a major critical violation category,
show a lower correlation with inspection outcomes than Food Establishment complaints.

---

## Exploratory Data Analysis

```{r eda-complaint-volumes}
# Visualize relative volumes to contextualize the correlation analysis.
# Large volume differences between types affect the reliability of
# correlation estimates and should be disclosed upfront.
complaints |>
  count(complaint_type, sort = TRUE) |>
  mutate(
    pct = n / sum(n),
    complaint_type = fct_reorder(complaint_type, n)
  ) |>
  ggplot(aes(x = n, y = complaint_type, fill = complaint_type)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(comma(n), " (", percent(pct, accuracy = 0.1), ")")),
            hjust = -0.05, size = 3.5) +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0, 0.35))) +
  scale_fill_manual(values = c(
    "Rodent" = "#E76F51", "Food Establishment" = "#264653",
    "Food Poisoning" = "#2A9D8F", "Unsanitary Condition" = "#E9C46A"
  )) +
  labs(title = "Food-Related 311 Complaint Volumes by Type",
       subtitle = "Rodent complaints dominate, comprising over 70% of total volume",
       x = "Total Complaints (2022-2026)", y = NULL,
       caption = "Source: NYC 311 (2022-2026)") +
  theme_nyc()
```

Rodent complaints account for 71% of all food-related 311 activity, making it the
dominant signal by volume. Food Establishment complaints represent 22%, Food Poisoning
about 6%, and Unsanitary Condition less than 1%. This volume imbalance has two
implications for the analysis. First, the ZIP-level correlation estimates will be
most reliable for Rodent and Food Establishment complaints, which have sufficient
volume to produce stable ZIP-level counts, and least reliable for Unsanitary
Condition, which averages fewer than 10 complaints per ZIP over the entire study
period. Second, any aggregate complaint-outcome relationship documented by the team's
cross-sectional analysts will be disproportionately driven by Rodent complaint volume
since it constitutes such a large fraction of total complaints. The type-specific
analysis here tests whether Rodent complaints are actually the type carrying the most
signal, or whether the smaller-volume types punch above their weight.

```{r eda-seasonal}
# Examine seasonal patterns by complaint type to understand whether
# volume differences are stable year-round or seasonal artifacts
complaints |>
  mutate(month_of_year = month(date, label = TRUE)) |>
  count(month_of_year, complaint_type) |>
  ggplot(aes(x = month_of_year, y = n, fill = complaint_type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(
    "Rodent" = "#E76F51", "Food Establishment" = "#264653",
    "Food Poisoning" = "#2A9D8F", "Unsanitary Condition" = "#E9C46A"
  )) +
  scale_y_continuous(labels = comma) +
  labs(title = "Seasonal Patterns in Food-Related 311 Complaints",
       subtitle = "Rodent complaints peak strongly in summer; other types are more stable",
       x = NULL, y = "Monthly Complaint Count", fill = "Complaint Type",
       caption = "Source: NYC 311 (2022-2026)") +
  theme_nyc()
```

The seasonal pattern chart reveals important differences in temporal behavior across
complaint types. Rodent complaints show a pronounced summer peak, with volume in July
and August roughly double that of December and January. This seasonal pattern reflects
the well-documented seasonal cycle of rodent activity, which increases in warm months
as outdoor dining, garbage disposal, and foraging behavior all intensify. The other
three complaint types show much more stable month-to-month patterns, with Food
Poisoning complaints showing a modest elevation in summer months consistent with
the increased risk of foodborne illness in warm weather. This seasonal heterogeneity
is relevant because it means the correlation between Rodent complaint volume and
inspection outcomes may partly capture seasonal patterns rather than pure food safety
signal, while the correlations for more temporally stable complaint types may be
cleaner measures of the underlying relationship.

---

## Analysis

### Building the ZIP-Level Dataset

```{r sq4-data}
# Wide-format complaint counts per ZIP enable simultaneous comparison
# of all four complaint types in the correlation analysis.
# Minimum 10 inspections required per ZIP for stable outcome estimates.
sq4_complaints <- complaints |>
  count(zipcode, complaint_type) |>
  pivot_wider(names_from = complaint_type, values_from = n,
              values_fill = 0)

sq4_inspections <- inspections |>
  group_by(zipcode) |>
  summarise(
    n_inspections = n_distinct(paste(camis, inspection_date)),
    pct_critical  = mean(is_critical, na.rm = TRUE),
    mean_score    = mean(score,        na.rm = TRUE),
    pct_grade_a   = mean(grade == "A", na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(n_inspections >= 10)

sq4_data <- sq4_inspections |>
  left_join(sq4_complaints, by = "zipcode")

complaint_cols <- intersect(names(sq4_data), unique(complaints$complaint_type))
cat("ZIP observations:", nrow(sq4_data), "\n")
cat("Complaint types:", paste(complaint_cols, collapse = ", "), "\n")
```

### Correlation Analysis

```{r sq4-cors}
# Pearson correlation computed independently for each complaint type
# against three inspection outcome measures. The minimum of 10 ZIP
# observations with non-zero complaint counts is enforced to avoid
# correlations based on a handful of data points.
sq4_cors <- map_dfr(complaint_cols, function(ct) {
  vals <- sq4_data[[ct]]
  if (sum(!is.na(vals) & vals > 0) < 10) return(NULL)
  tibble(
    complaint_type   = ct,
    cor_pct_critical = cor(vals, sq4_data$pct_critical,
                           use = "pairwise.complete.obs"),
    cor_mean_score   = cor(vals, sq4_data$mean_score,
                           use = "pairwise.complete.obs"),
    cor_pct_grade_a  = cor(vals, sq4_data$pct_grade_a,
                           use = "pairwise.complete.obs"),
    n_zips           = sum(!is.na(vals) & vals > 0)
  )
}) |>
  filter(!is.na(cor_pct_critical)) |>
  arrange(desc(abs(cor_pct_critical)))

sq4_cors |>
  kable(digits = 4, caption = "Complaint Type Correlations with Inspection Outcomes",
        col.names = c("Complaint Type","Cor w/ % Critical","Cor w/ Mean Score",
                      "Cor w/ % Grade A","ZIP Count")) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"))
```

```{r sq4-plot}
sq4_cors |>
  mutate(complaint_type = fct_reorder(complaint_type, cor_pct_critical),
         direction = if_else(cor_pct_critical >= 0, "Positive","Negative")) |>
  ggplot(aes(x = cor_pct_critical, y = complaint_type, fill = direction)) +
  geom_col(show.legend = FALSE, width = 0.6) +
  geom_vline(xintercept = 0, linewidth = 0.9) +
  geom_text(aes(label = round(cor_pct_critical, 3),
                hjust = if_else(cor_pct_critical >= 0, -0.2, 1.2)),
            size = 4, fontface = "bold") +
  scale_fill_manual(values = c("Positive" = "#E76F51","Negative" = "#2A9D8F")) +
  scale_x_continuous(expand = expansion(mult = 0.25)) +
  labs(title = "Which Complaint Types Best Predict Critical Violations?",
       subtitle = "Pearson correlation with % critical violations at ZIP level",
       x = "Correlation with % Critical Violations", y = NULL,
       caption = "Sources: DOHMH Inspections, NYC 311 (2022-2026)") +
  theme_nyc()
```

### Scatter Grid by Complaint Type

```{r sq4-scatter-grid}
# Individual scatter plots verify that correlations reflect genuine
# trends rather than being driven by outliers in any single complaint type
sq4_long <- sq4_data |>
  select(zipcode, pct_critical, all_of(complaint_cols)) |>
  pivot_longer(all_of(complaint_cols), names_to = "complaint_type",
               values_to = "count") |>
  filter(count > 0)

ggplot(sq4_long, aes(x = count, y = pct_critical)) +
  geom_point(alpha = 0.4, color = "#264653", size = 1.8) +
  geom_smooth(method = "lm", se = TRUE, color = "#E76F51", linewidth = 1) +
  facet_wrap(~ complaint_type, scales = "free_x") +
  scale_x_continuous(labels = comma) +
  labs(title = "ZIP-Level Scatter: Each Complaint Type vs % Critical Violations",
       subtitle = "All four types show positive trends; slopes differ by type",
       x = "Complaint Count", y = "% Critical Violations",
       caption = "Sources: DOHMH Inspections, NYC 311 (2022-2026)") +
  theme_nyc()
```

---

## Findings and Interpretation

The correlation results produce a finding that directly contradicts the initial
hypothesis. Food Establishment complaints, not Rodent complaints, show the strongest
correlation with critical violation rates at the ZIP level (r = 0.157). Rodent
complaints come second at r = 0.116, followed by Food Poisoning at r = 0.109, and
Unsanitary Condition at r = 0.047. All four complaint types show positive correlations
with both the critical violation rate and the mean inspection score, and negative
correlations with the grade-A rate, confirming that all four types point in the same
direction even if their magnitudes differ. The directional consistency across all
three outcome measures for all four complaint types is itself an important finding:
it rules out the possibility that the positive correlations are artifacts of a
particular outcome measure and suggests a genuine underlying relationship between
complaint activity and food safety risk regardless of how that risk is measured.

The stronger performance of Food Establishment complaints relative to Rodent
complaints warrants careful explanation. The Food Establishment category captures a
wide and heterogeneous range of citizen concerns about restaurant conditions, from
visible cleanliness and food handling practices to concerns about ingredient
freshness, employee hygiene, and general sanitation. This breadth may actually be
an advantage: because the category encompasses such a wide range of concerns, a
high Food Establishment complaint rate in a ZIP code signals a broadly elevated
level of concern about restaurant conditions across multiple dimensions, rather than
a single specific problem. Rodent complaints, by contrast, are specific to one type
of sanitation failure, and while rodent presence is directly cited as a critical
violation, it is a narrower signal than the broad-based concern captured by Food
Establishment complaints.

The relatively strong performance of Food Poisoning complaints (r = 0.109) is
notable given the small volume of this complaint type. Food Poisoning complaints
reflect the most direct possible evidence of food safety failure: a consumer who
became ill after eating at a restaurant. The fact that this small-volume complaint
type produces correlations nearly as large as the much higher-volume Rodent
category suggests that Food Poisoning complaints may carry a very high per-complaint
signal-to-noise ratio. Each Food Poisoning complaint represents a confirmed health
outcome rather than a precautionary observation, making it a direct rather than
indirect indicator of the kind of food safety failures that inspectors are trained
to identify.

Unsanitary Condition complaints show the weakest association (r = 0.047), which is
consistent with the heterogeneity of this category. Unsanitary Condition complaints
include concerns about street cleanliness, garbage disposal, and conditions in
non-food establishments, meaning that many complaints in this category may have no
relationship to restaurant food safety outcomes. The attenuation caused by this
category-level imprecision would naturally produce weaker correlations with
restaurant inspection outcomes.

The scatter grid confirms that all four correlations reflect genuine distributed
trends rather than being driven by outliers. The positive relationship between
complaint count and critical violation rate holds broadly across the range of ZIP
code complaint counts for each type, with no obvious structural breaks or outlier
clusters driving the results.

---

## Connection to the Overarching Question

This analysis adds specificity to the team's answer to the overarching question.
The overarching question asks whether complaint activity is associated with worse
inspection outcomes, and the cross-sectional analyses confirm that it is. My
contribution reveals that this association is not uniform across complaint types:
Food Establishment complaints carry the strongest signal despite not being the most
common complaint type, while Unsanitary Condition complaints carry relatively little
signal despite covering food safety concerns. This finding refines the policy
implications of the project. Rather than treating all food-related 311 complaints
as equivalent signals, regulators could develop type-weighted prioritization systems
that give greater weight to Food Establishment complaints, which appear to be the
most reliable indicators of underlying critical violation rates. This type-weighted
approach could improve inspection targeting efficiency without requiring any new
data collection infrastructure, since the 311 system already disaggregates complaints
by type.

---

## Session Info

```{r session}
sessionInfo()
```
