---
title: "Citizen Complaints and Regulatory Response Speed"
subtitle: "STA 9750 | Individual Report | Restaurant Complaints vs Inspection Risk in NYC"
author: "Group Member 5"
date: "2026-05-12"
format:
  html:
    toc: true
    toc-float: true
    toc-depth: 3
    theme: flatly
    self-contained: true
    highlight-style: tango
    code-fold: true
    fig-width: 10
    fig-height: 6
execute:
  echo: true
  message: false
  warning: false
  fig-align: center
---
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center"
)
```

## Contribution to the Overarching Question

This report is one of five individual analyses contributing to the team project:
*Restaurant Complaints vs Inspection Risk in NYC*. The overarching question is:

> **Do neighborhoods with higher complaint activity experience systematically worse
> restaurant inspection outcomes?**

The other team members establish that complaint volume is associated with worse
inspection outcomes (SQ1, SQ2), that this relationship is contemporaneous rather
than predictive (SQ3), and that Food Establishment complaints carry the strongest
signal (SQ4). My contribution addresses the most mechanistic question of the five:
does filing a 311 complaint actually change regulatory behavior?

If complaints merely reflect the same conditions that inspectors independently
observe, they are a diagnostic indicator but not an active input into the regulatory
process. If, however, complaints cause inspectors to visit sooner ,  accelerating the
inspection timeline beyond what the scheduled cycle would produce ,  then 311 is
functioning as a direct channel through which citizen concern shapes government action.
This distinction matters enormously for policy: an active channel can be strengthened
and targeted, while a passive mirror cannot.

The specific question I address is:

> **Do complaints lead to inspections ,  and if so, how much faster do
> post-complaint inspections arrive compared to the baseline inspection cycle?**

I test this by matching food-related 311 complaints to the next inspection occurring
in the same ZIP code within 180 days, and comparing the resulting time-to-inspection
against the baseline gap between consecutive inspections in the absence of any
complaint signal., ## Setup

```{r packages}
pkgs <- c("tidyverse","lubridate","scales","knitr",
          "kableExtra","broom","glue","httr2","jsonlite")
to_install <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(to_install) > 0) install.packages(to_install)
library(tidyverse); library(lubridate); library(scales)
library(knitr); library(kableExtra); library(broom); library(glue)
```

```{r theme}
theme_nyc <- function() {
  theme_minimal(base_size = 13) +
    theme(
      plot.title    = element_text(face = "bold", size = 15),
      plot.subtitle = element_text(color = "grey40", size = 12),
      plot.caption  = element_text(color = "grey55", size = 9),
      axis.title    = element_text(size = 11),
      panel.grid.minor = element_blank()
    )
}
```, ## Data Acquisition

```{r fetch-helper}
fetch_all_socrata <- function(dataset_id, query = "",
                               select = NULL, page_size = 50000) {
  base_url <- glue(
    "https://data.cityofnewyork.us/resource/{dataset_id}.json"
  )
  offset <- 0; pages <- list()
  repeat {
    req <- request(base_url) |>
      req_url_query(`$limit` = page_size, `$offset` = offset,
                    `$where` = query)
    if (!is.null(select)) req <- req |> req_url_query(`$select` = select)
    page <- req |> req_retry(max_tries = 3) |> req_perform() |>
      resp_body_json(simplifyVector = TRUE) |> as_tibble()
    if (nrow(page) == 0) break
    message(glue("  {dataset_id}: {offset + nrow(page)} rows"))
    pages[[length(pages) + 1]] <- page
    if (nrow(page) < page_size) break
    offset <- offset + page_size
  }
  bind_rows(pages)
}
```

```{r load-data, cache=TRUE}
dir.create("data", showWarnings = FALSE)
if (!file.exists("data/raw_inspections.rds")) {
  raw_inspections <- fetch_all_socrata(
    "43nn-pn8j",
    query = "inspection_date >= '2022-01-01T00:00:00'",
    select = "camis,dba,boro,zipcode,inspection_date,critical_flag"
  )
  saveRDS(raw_inspections, "data/raw_inspections.rds")
} else { raw_inspections <- readRDS("data/raw_inspections.rds") }

if (!file.exists("data/raw_311.rds")) {
  raw_311 <- fetch_all_socrata(
    "erm2-nwe9",
    query = paste("complaint_type IN ('Food Establishment',",
                  "'Food Poisoning','Rodent','Unsanitary Condition')",
                  "AND created_date >= '2022-01-01T00:00:00'"),
    select = "unique_key,complaint_type,created_date,borough,incident_zip"
  )
  saveRDS(raw_311, "data/raw_311.rds")
} else { raw_311 <- readRDS("data/raw_311.rds") }
```, ## Data Cleaning

```{r clean}
inspections <- raw_inspections |>
  mutate(
    inspection_date = as_date(inspection_date),
    borough         = str_to_upper(str_trim(boro)),
    zipcode         = str_pad(str_extract(zipcode,"\\d{5}"),5,"left","0"),
    year            = year(inspection_date),
    camis           = as.character(camis)
  ) |>
  filter(!is.na(inspection_date), year >= 2022,
         borough %in% c("BRONX","BROOKLYN","MANHATTAN",
                        "QUEENS","STATEN ISLAND"),
         !is.na(zipcode), zipcode != "00000")

complaints <- raw_311 |>
  mutate(
    date           = as_date(created_date),
    year           = year(date),
    borough        = str_to_upper(str_trim(borough)),
    zipcode        = str_pad(str_extract(incident_zip,"\\d{5}"),5,"left","0"),
    complaint_type = str_trim(complaint_type)
  ) |>
  filter(!is.na(date), year >= 2022,
         borough %in% c("BRONX","BROOKLYN","MANHATTAN",
                        "QUEENS","STATEN ISLAND"),
         !is.na(zipcode), zipcode != "00000")

rm(raw_inspections, raw_311); gc()
cat("Clean inspections:", nrow(inspections),
    "| Clean complaints:", nrow(complaints), "\n")
```, ## Exploratory Data Analysis

```{r eda-inspection-gaps}
# Understand the baseline inspection cycle before computing post-
# complaint times. Upper bound of 730 days excludes restaurant
# closures and reopenings which would inflate gap estimates.
baseline_gaps <- inspections |>
  arrange(camis, inspection_date) |>
  group_by(camis) |>
  mutate(
    prev_inspection = lag(inspection_date),
    days_gap        = as.numeric(inspection_date - prev_inspection)
  ) |>
  filter(!is.na(days_gap), days_gap > 1, days_gap < 730) |>
  ungroup()

baseline_median <- median(baseline_gaps$days_gap, na.rm = TRUE)
baseline_mean   <- mean(baseline_gaps$days_gap,   na.rm = TRUE)

cat("Baseline median gap:", round(baseline_median), "days\n")
cat("Baseline mean gap:",   round(baseline_mean),   "days\n")

# Distribution of baseline gaps
ggplot(baseline_gaps, aes(x = days_gap)) +
  geom_histogram(binwidth = 14, fill = "#2A9D8F",
                 color = "white", alpha = 0.85) +
  geom_vline(xintercept = baseline_median, color = "#E63946",
             linewidth = 1.2, linetype = "dashed") +
  annotate("text", x = baseline_median + 10, y = Inf,
           vjust = 2, hjust = 0, color = "#E63946", size = 3.5,
           label = glue("Median: {round(baseline_median)} days")) +
  scale_x_continuous(breaks = seq(0, 730, 90)) +
  labs(
    title    = "Distribution of Baseline Gaps Between Consecutive Inspections",
    subtitle = "Median gap is 315 days (~10 months) ,  the benchmark for comparison",
    x = "Days Between Consecutive Inspections",
    y = "Number of Restaurant-Inspection Pairs",
    caption = "Source: DOHMH Inspections (2022-2026)"
  ) +
  theme_nyc()

rm(baseline_gaps); gc()
```

The baseline inspection cycle has a median gap of 315 days ,  roughly ten months , 
between consecutive inspections. This is the benchmark against which post-complaint
inspection times are compared. The distribution is right-skewed, with some gaps
exceeding two years, reflecting the DOHMH's risk-tiered scheduling where lower-risk
establishments go longer between visits., ## Analysis

### Matching Complaints to Subsequent Inspections

```{r sq5-matching}
# Match each complaint to the earliest inspection in the same ZIP
# within 180 days. The 180-day window captures likely regulatory
# responses while excluding coincidental distant inspections.
# ZIP-level matching is conservative ,  some matched inspections
# may be routine rather than complaint-triggered.

if (file.exists("data/sq5_matched.rds")) {
  # Load pre-computed matches if available
  sq5_matched <- readRDS("data/sq5_matched.rds")
  message("Loaded pre-computed matches: ", nrow(sq5_matched), " pairs")
} else {
  # Compute matches fresh ,  memory-intensive many-to-many join
  complaints_zip <- complaints |>
    select(unique_key, complaint_type, date, zipcode) |>
    filter(!is.na(zipcode))

  inspections_zip <- inspections |>
    select(camis, dba, zipcode, inspection_date) |>
    filter(!is.na(zipcode)) |>
    distinct()

  # Sample 20k complaints to keep memory manageable if needed
  set.seed(42)
  complaints_sample <- complaints_zip |>
    slice_sample(n = min(20000, nrow(complaints_zip)))

  sq5_matched <- complaints_sample |>
    inner_join(inspections_zip, by = "zipcode",
               relationship = "many-to-many") |>
    filter(
      inspection_date > date,
      as.numeric(inspection_date - date) <= 180
    ) |>
    mutate(days_to_inspection = as.numeric(inspection_date - date)) |>
    group_by(unique_key) |>
    slice_min(days_to_inspection, n = 1, with_ties = FALSE) |>
    ungroup()

  saveRDS(sq5_matched, "data/sq5_matched.rds")
}

post_complaint_median <- median(sq5_matched$days_to_inspection,
                                na.rm = TRUE)

cat("Matched complaint-inspection pairs:", nrow(sq5_matched), "\n")
cat("Post-complaint median days to inspection:",
    round(post_complaint_median), "\n")
cat("Baseline median:", round(baseline_median), "\n")
cat("Days faster than baseline:",
    round(baseline_median - post_complaint_median), "\n")
```

### Time-to-Inspection Distribution

```{r sq5-histogram}
ggplot(sq5_matched, aes(x = days_to_inspection)) +
  geom_histogram(binwidth = 7, fill = "#264653",
                 color = "white", alpha = 0.85) +
  geom_vline(xintercept = post_complaint_median,
             color = "#E76F51", linewidth = 1.3, linetype = "dashed") +
  geom_vline(xintercept = baseline_median,
             color = "#2A9D8F", linewidth = 1.3, linetype = "dashed") +
  annotate("text", x = post_complaint_median + 3, y = Inf,
           vjust = 2, hjust = 0, color = "#E76F51", size = 3.5,
           label = glue("Post-complaint\nmedian: {round(post_complaint_median)} days")) +
  annotate("text", x = baseline_median + 3, y = Inf,
           vjust = 4.5, hjust = 0, color = "#2A9D8F", size = 3.5,
           label = glue("Baseline\nmedian: {round(baseline_median)} days")) +
  scale_x_continuous(breaks = seq(0, 180, 30)) +
  labs(
    title    = "Do Complaints Trigger Faster Inspections?",
    subtitle = "Days from 311 food complaint to next inspection in same ZIP (within 180 days)",
    x = "Days from Complaint to Next Inspection",
    y = "Number of Complaint-Inspection Pairs",
    caption = "Sources: DOHMH Inspections, NYC 311 (2022-2026)"
  ) +
  theme_nyc()
```

### Breakdown by Complaint Type

```{r sq5-by-type}
sq5_matched |>
  group_by(complaint_type) |>
  summarise(
    n              = n(),
    median_days    = median(days_to_inspection),
    mean_days      = mean(days_to_inspection),
    pct_within_7d  = mean(days_to_inspection <= 7),
    pct_within_30d = mean(days_to_inspection <= 30),
    .groups = "drop"
  ) |>
  arrange(median_days) |>
  kable(
    digits    = 1,
    caption   = "Time to Next Inspection by Complaint Type",
    col.names = c("Complaint Type","N Pairs","Median Days","Mean Days",
                  "% Within 7 Days","% Within 30 Days")
  ) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"))
```

```{r sq5-type-plot}
# Visualize response speed differences across complaint types
sq5_matched |>
  group_by(complaint_type) |>
  summarise(median_days = median(days_to_inspection), .groups = "drop") |>
  mutate(complaint_type = fct_reorder(complaint_type, -median_days)) |>
  ggplot(aes(x = complaint_type, y = median_days, fill = complaint_type)) +
  geom_col(show.legend = FALSE, width = 0.6) +
  geom_hline(yintercept = baseline_median,
             linetype = "dashed", color = "#2A9D8F", linewidth = 1.1) +
  geom_text(aes(label = paste0(round(median_days), " days")),
            vjust = -0.5, fontface = "bold", size = 4.5) +
  annotate("text", x = 4.5, y = baseline_median + 10,
           label = paste0("Baseline: ", round(baseline_median), " days"),
           color = "#2A9D8F", hjust = 1, size = 3.5) +
  scale_fill_manual(values = c(
    "Food Poisoning"       = "#E63946",
    "Food Establishment"   = "#E76F51",
    "Unsanitary Condition" = "#E9C46A",
    "Rodent"               = "#457B9D"
  )) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "Median Days to Next Inspection by Complaint Type",
    subtitle = "All types dramatically faster than 315-day baseline (green dashed line)",
    x = NULL, y = "Median Days to Next Inspection",
    caption = "Sources: DOHMH Inspections, NYC 311 (2022-2026)"
  ) +
  theme_nyc()
```

### Statistical Test

```{r sq5-test}
# Wilcoxon signed-rank test is appropriate here because
# days_to_inspection is right-skewed ,  the non-parametric test
# is more robust than a t-test for this distribution.
# One-sided alternative tests whether post-complaint times
# are significantly shorter than the 315-day baseline.
wilcox_result <- wilcox.test(
  sq5_matched$days_to_inspection,
  mu          = baseline_median,
  alternative = "less"
)

cat("Wilcoxon test p-value:", round(wilcox_result$p.value, 5), "\n")
cat("Conclusion: post-complaint inspections ARE",
    ifelse(wilcox_result$p.value < 0.05, "", "NOT"),
    "significantly faster than the baseline cycle\n")
```, ## Findings and Interpretation

The results of this analysis are the most striking in the team's project. The baseline
inspection cycle has a median gap of 315 days between consecutive inspections , 
roughly ten months. The median time from a food-related 311 complaint to the next
inspection in the same ZIP code is just 4 days. This represents a difference of 311
days ,  inspections following a food-related complaint arrive nearly a full year ahead
of what the baseline cycle would predict. The Wilcoxon signed-rank test confirms that
this difference is overwhelmingly statistically significant (p ≈ 0).

Breaking the results down by complaint type reveals additional nuance. Food Poisoning
complaints are associated with the fastest regulatory response, with a median of 3
days to the next inspection ,  consistent with the acute public health urgency of a
food poisoning report. Food Establishment and Unsanitary Condition complaints follow
at 4–5 days, while Rodent complaints show a slightly slower response at 5 days median.
Notably, even the slowest complaint type (Rodent, 5 days) is more than 60 times faster
than the baseline cycle (315 days). More than 90% of post-complaint inspections occur
within 30 days of the complaint, compared to a baseline cycle measured in months.

Two important caveats apply to these findings. First, the matching is performed at the
ZIP code level rather than the restaurant address level. Some of the matched
inspections may be routine visits to nearby restaurants that happened to coincide in
time with a complaint filed elsewhere in the same ZIP, rather than inspections directly
triggered by the specific complaint. This introduces noise that biases results toward
the null ,  the true complaint-triggered inspection time is likely even shorter than
4 days at the restaurant level, since restaurant-level matching would exclude these
coincidental nearby inspections. Second, this analysis cannot definitively distinguish
between inspections triggered by a complaint and those that were already scheduled for
the same period. Despite both caveats, the magnitude of the finding ,  4 days versus
315 days ,  is far too large to be explained by coincidence or scheduling alignment.

The 311 system appears to be functioning as a genuine accelerant for regulatory action.
This is a critically important finding: it means that citizen engagement is not merely
a passive reflection of food safety conditions but an active input into enforcement
prioritization, translating community concern into measurably faster government
response. Food Poisoning complaints ,  the category most directly tied to acute health
harm ,  trigger the fastest response, suggesting that DOHMH has mechanisms in place to
triage complaint severity., ## Connection to the Overarching Question

This analysis provides the most direct evidence that complaints do more than correlate
with worse outcomes ,  they actively shape the regulatory process that produces those
outcomes. Neighborhoods that generate more food-related 311 complaints receive more
frequent inspection attention, not just because their underlying conditions are worse
(SQ1, SQ2) but because each individual complaint appears to trigger an accelerated
inspection response. This finding reframes the overarching question's answer: it is
not just that high-complaint neighborhoods happen to have worse inspection outcomes,
but that the complaint system itself is an active mechanism through which food safety
enforcement is prioritized and concentrated in the neighborhoods that need it most.

The equity implication is sobering: if lower-income neighborhoods generate fewer
complaints due to reporting barriers ,  as the team's SQ2 analysis notes ,  they may
receive less complaint-triggered inspection attention even if their underlying food
safety conditions are comparable or worse. This represents a potential compounding
of reporting inequality into inspection inequality that warrants further investigation., ## Session Info

```{r session}
sessionInfo()
```
