---
title: "Neighborhood Complaint Persistence and Repeat Critical Violations"
subtitle: "STA 9750 — Individual Report | Restaurant Complaints vs Inspection Risk in NYC"
author: "Mary Ononye"
date: today
format:
  html:
    toc: true
    toc-float: true
    toc-depth: 3
    theme: flatly
    highlight-style: tango
    code-fold: true
    fig-width: 10
    fig-height: 6
    self-contained: true
execute:
  echo: true
  message: false
  warning: false
  fig-align: center
---

## Overarching Question and My Contribution

This report is one of five individual analyses contributing to the team project:
*Restaurant Complaints vs Inspection Risk in NYC*. The overarching question guiding
the project is:

> **Do neighborhoods with higher complaint activity experience systematically worse
> restaurant inspection outcomes?**

My specific contribution focuses on the **persistence** dimension of this question.
Rather than asking whether complaints and inspection outcomes are associated at a
single point in time, I ask whether neighborhoods that sustain elevated complaint
volumes over time accumulate a particular kind of inspection failure — repeat critical
violations — at disproportionate rates. This is a harder test of the complaint-risk
relationship than a simple cross-sectional correlation: it asks whether chronic
neighborhood-level complaint activity is a marker of entrenched, unresolved food
safety problems rather than transient or correctable ones.

My specific question is:

> **Do neighborhoods with persistently high complaint volume have higher repeat
> critical violations?**

A repeat critical violation is defined here as the same restaurant being cited for
the same violation code on two or more separate inspection dates. This definition
isolates a pattern of non-compliance that persists across inspection cycles — the
kind of structural food safety failure that poses the greatest sustained public
health risk.

For full project context, see the team's other individual reports:

- [Complaint Volume and Inspection Outcomes](report_complaint_outcomes.html)
- [Temporal Dynamics of Complaints](report_temporal_signal.html)
- [Complaint Type Specificity](report_complaint_types.html)
- [Regulatory Response Speed](report_regulatory_response.html)

---

## Setup

```{r packages}
# Install any missing packages automatically
pkgs <- c("tidyverse", "lubridate", "scales",
          "knitr", "kableExtra", "broom",
          "glue", "httr2", "jsonlite")
to_install <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(to_install) > 0) install.packages(to_install)

library(tidyverse)
library(lubridate)
library(scales)
library(knitr)
library(kableExtra)
library(broom)
library(glue)
library(httr2)
library(jsonlite)
```

```{r theme}
# Consistent ggplot theme applied across all figures
theme_nyc <- function() {
  theme_minimal(base_size = 13) +
    theme(
      plot.title       = element_text(face = "bold", size = 15),
      plot.subtitle    = element_text(color = "grey40", size = 12),
      plot.caption     = element_text(color = "grey55", size = 9),
      axis.title       = element_text(size = 11),
      panel.grid.minor = element_blank(),
      strip.text       = element_text(face = "bold")
    )
}

borough_colors <- c(
  "BRONX"         = "#E76F51",
  "BROOKLYN"      = "#2A9D8F",
  "MANHATTAN"     = "#264653",
  "QUEENS"        = "#E9C46A",
  "STATEN ISLAND" = "#A8DADC"
)
```

---

## Data Acquisition

```{r fetch-helper}
# Paginated Socrata API fetcher — handles datasets exceeding the
# 50k row single-request limit through offset-based pagination.
# Returns a single tibble of all rows matching the query.
fetch_all_socrata <- function(dataset_id, query = "",
                               select = NULL, page_size = 50000) {
  base_url <- glue(
    "https://data.cityofnewyork.us/resource/{dataset_id}.json"
  )
  offset <- 0
  pages  <- list()

  repeat {
    req <- request(base_url) |>
      req_url_query(
        `$limit`  = page_size,
        `$offset` = offset,
        `$where`  = query
      )
    # Add column selection if specified to reduce API payload size
    if (!is.null(select)) req <- req |> req_url_query(`$select` = select)

    page <- req |>
      req_retry(max_tries = 3) |>
      req_perform() |>
      resp_body_json(simplifyVector = TRUE) |>
      as_tibble()

    if (nrow(page) == 0) break
    message(glue("  {dataset_id}: {offset + nrow(page)} rows fetched"))
    pages[[length(pages) + 1]] <- page
    if (nrow(page) < page_size) break
    offset <- offset + page_size
  }
  bind_rows(pages)
}
```

```{r load-data}
#| cache: true
# Cache data locally to avoid re-downloading on every render.
# On first run, fetches from NYC Open Data API and saves locally.
# On subsequent runs, loads from the cached RDS file.
dir.create("data", showWarnings = FALSE)

# DOHMH Restaurant Inspection Results
# Source: https://data.cityofnewyork.us/Health/DOHMH-New-York-City-Restaurant-Inspection-Results/43nn-pn8j
if (!file.exists("data/raw_inspections.rds")) {
  message("Fetching inspections from NYC Open Data...")
  raw_inspections <- fetch_all_socrata(
    dataset_id = "43nn-pn8j",
    query  = "inspection_date >= '2022-01-01T00:00:00'",
    select = paste(
      "camis,dba,boro,zipcode,inspection_date,score,grade,",
      "violation_code,critical_flag,action,inspection_type",
      sep = ""
    )
  )
  saveRDS(raw_inspections, "data/raw_inspections.rds")
  message("Saved: ", nrow(raw_inspections), " rows")
} else {
  raw_inspections <- readRDS("data/raw_inspections.rds")
}

# 311 Food-Related Complaints
# Source: https://data.cityofnewyork.us/Social-Services/311-Service-Requests-from-2020-to-Present/erm2-nwe9
if (!file.exists("data/raw_311.rds")) {
  message("Fetching 311 complaints from NYC Open Data...")
  raw_311 <- fetch_all_socrata(
    dataset_id = "erm2-nwe9",
    query  = paste(
      "complaint_type IN (",
      "'Food Establishment','Food Poisoning',",
      "'Rodent','Unsanitary Condition'",
      ") AND created_date >= '2022-01-01T00:00:00'",
      sep = ""
    ),
    select = "unique_key,complaint_type,created_date,borough,incident_zip"
  )
  saveRDS(raw_311, "data/raw_311.rds")
  message("Saved: ", nrow(raw_311), " rows")
} else {
  raw_311 <- readRDS("data/raw_311.rds")
}

cat("Inspections rows:", nrow(raw_inspections), "\n")
cat("311 rows:", nrow(raw_311), "\n")
```

---

## Data Cleaning

```{r clean-inspections}
# Parse and standardize inspection records.
# is_critical derived from text field by detecting "critical" without "not".
# Grades restricted to A/B/C — other values indicate incomplete grading cycles.
# ZIP codes zero-padded for consistent joining with 311 data.
inspections <- raw_inspections |>
  mutate(
    inspection_date = as_date(inspection_date),
    score           = as.numeric(score),
    grade           = str_trim(grade),
    grade           = if_else(grade %in% c("A","B","C"),
                              grade, NA_character_),
    is_critical     = str_detect(tolower(critical_flag), "critical") &
                      !str_detect(tolower(critical_flag), "not"),
    borough         = str_to_upper(str_trim(boro)),
    zipcode         = str_pad(str_extract(zipcode, "\\d{5}"),
                              5, "left", "0"),
    year            = year(inspection_date),
    month           = floor_date(inspection_date, "month"),
    camis           = as.character(camis)
  ) |>
  filter(
    !is.na(inspection_date),
    year >= 2022,
    borough %in% c("BRONX","BROOKLYN","MANHATTAN",
                   "QUEENS","STATEN ISLAND"),
    !is.na(zipcode), zipcode != "00000"
  ) |>
  select(camis, dba, borough, zipcode, inspection_date,
         year, month, score, grade, violation_code,
         critical_flag, is_critical)

rm(raw_inspections); gc()
cat("Clean inspection rows:", nrow(inspections), "\n")
```

```{r clean-complaints}
# Standardize 311 data using same ZIP padding logic as inspections
complaints <- raw_311 |>
  mutate(
    date           = as_date(created_date),
    month          = floor_date(date, "month"),
    year           = year(date),
    borough        = str_to_upper(str_trim(borough)),
    zipcode        = str_pad(str_extract(incident_zip, "\\d{5}"),
                             5, "left", "0"),
    complaint_type = str_trim(complaint_type)
  ) |>
  filter(
    !is.na(date), year >= 2022,
    borough %in% c("BRONX","BROOKLYN","MANHATTAN",
                   "QUEENS","STATEN ISLAND"),
    !is.na(zipcode), zipcode != "00000"
  ) |>
  select(unique_key, complaint_type, date, month, year, borough, zipcode)

rm(raw_311); gc()
cat("Clean complaint rows:", nrow(complaints), "\n")
```

```{r aggregate-complaints}
# Aggregate complaints to ZIP-month level for quartile classification
comp_zip_month <- complaints |>
  count(zipcode, borough, month, complaint_type) |>
  pivot_wider(
    names_from   = complaint_type,
    values_from  = n,
    values_fill  = 0,
    names_prefix = "comp_"
  ) |>
  rename_with(~ str_replace_all(., " ", "_"), starts_with("comp_")) |>
  mutate(comp_total = rowSums(across(starts_with("comp_"))))
```

---

## Exploratory Data Analysis

Before testing the main hypothesis, it is important to understand the distribution
of complaint volume and repeat violation rates across ZIP codes.

```{r eda-complaint-dist}
# Distribution of average monthly complaint volume across ZIPs.
# The right-skewed shape motivates using quartile classification
# rather than treating complaint volume as a linear predictor.
comp_zip_month |>
  group_by(zipcode) |>
  summarise(avg_monthly = mean(comp_total), .groups = "drop") |>
  ggplot(aes(x = avg_monthly)) +
  geom_histogram(binwidth = 5, fill = "#264653", color = "white") +
  scale_x_continuous(labels = comma) +
  labs(
    title    = "Distribution of Average Monthly Complaint Volume by ZIP Code",
    subtitle = "Right-skewed: most ZIPs generate few complaints; a small number generate many",
    x        = "Average Monthly Complaints",
    y        = "Number of ZIP Codes",
    caption  = "Source: NYC 311 (2022-2026)"
  ) +
  theme_nyc()
```

Complaint volume is heavily right-skewed — the majority of ZIPs generate fewer than
20 complaints per month, while a small number of high-density commercial ZIP codes
generate substantially more. This skewness justifies the quartile-based approach
used in the main analysis.

```{r eda-repeat-violations}
# Preview of repeat critical violation patterns before formal testing
repeat_criticals_eda <- inspections |>
  filter(is_critical, !is.na(violation_code)) |>
  group_by(camis, zipcode, violation_code) |>
  summarise(n_times_cited = n_distinct(inspection_date), .groups = "drop") |>
  filter(n_times_cited >= 2)

cat("Restaurants with repeat critical violations:",
    n_distinct(repeat_criticals_eda$camis), "\n")
cat("Total repeat critical instances:", nrow(repeat_criticals_eda), "\n")

repeat_criticals_eda |>
  count(violation_code, sort = TRUE) |>
  slice_head(n = 10) |>
  kable(
    caption   = "Top 10 Most Frequently Repeated Violation Codes",
    col.names = c("Violation Code", "Repeat Instances")
  ) |>
  kable_styling(bootstrap_options = c("striped","condensed"))
```

---

## Analysis

### Classifying Neighborhoods by Complaint Volume

```{r sq1-quartiles}
# Require at least 6 months of observation to ensure quartile assignment
# reflects a sustained pattern rather than a short-term spike
zip_complaint_profile <- comp_zip_month |>
  group_by(zipcode, borough) |>
  summarise(
    months_observed  = n(),
    avg_monthly_comp = mean(comp_total, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(months_observed >= 6) |>
  mutate(comp_quartile = ntile(avg_monthly_comp, 4))

# Quartile boundary summary for transparency
zip_complaint_profile |>
  group_by(comp_quartile) |>
  summarise(
    n_zips          = n(),
    min_avg         = min(avg_monthly_comp),
    max_avg         = max(avg_monthly_comp),
    median_avg      = median(avg_monthly_comp),
    .groups = "drop"
  ) |>
  kable(
    digits    = 1,
    caption   = "Complaint Volume Quartile Boundaries",
    col.names = c("Quartile","ZIP Codes","Min Avg/Month",
                  "Max Avg/Month","Median Avg/Month")
  ) |>
  kable_styling(bootstrap_options = c("striped","condensed"))
```

### Measuring Repeat Critical Violations

```{r sq1-repeat}
# Repeat critical violation: same restaurant + same violation code
# cited on 2+ distinct inspection dates — isolates persistent
# non-compliance across inspection cycles
repeat_criticals <- inspections |>
  filter(is_critical, !is.na(violation_code)) |>
  group_by(camis, zipcode, violation_code) |>
  summarise(n_times_cited = n_distinct(inspection_date), .groups = "drop") |>
  filter(n_times_cited >= 2) |>
  group_by(zipcode) |>
  summarise(
    n_repeat_critical         = n(),
    n_restaurants_with_repeat = n_distinct(camis),
    .groups = "drop"
  )

# Use restaurant count as denominator to normalize for ZIP size
restaurants_per_zip <- inspections |>
  group_by(zipcode) |>
  summarise(n_restaurants = n_distinct(camis), .groups = "drop")

sq1_data <- zip_complaint_profile |>
  left_join(repeat_criticals,    by = "zipcode") |>
  left_join(restaurants_per_zip, by = "zipcode") |>
  mutate(
    n_repeat_critical    = replace_na(n_repeat_critical, 0),
    repeat_critical_rate = n_repeat_critical / n_restaurants
  )
```

### Results

```{r sq1-summary-table}
sq1_summary <- sq1_data |>
  group_by(comp_quartile) |>
  summarise(
    n_zips                   = n(),
    avg_complaints_per_month = mean(avg_monthly_comp),
    avg_repeat_critical_rate = mean(repeat_critical_rate, na.rm = TRUE),
    sd_repeat_critical_rate  = sd(repeat_critical_rate,   na.rm = TRUE),
    .groups = "drop"
  )

sq1_summary |>
  kable(
    digits    = 2,
    caption   = "Repeat Critical Violation Rate by Complaint Volume Quartile",
    col.names = c("Quartile","ZIP Codes","Avg Monthly Complaints",
                  "Avg Repeat Critical Rate","SD")
  ) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"))
```

```{r sq1-plot}
sq1_summary |>
  mutate(quartile_label = case_when(
    comp_quartile == 1 ~ "Q1\n(Lowest)",
    comp_quartile == 2 ~ "Q2",
    comp_quartile == 3 ~ "Q3",
    comp_quartile == 4 ~ "Q4\n(Highest)"
  )) |>
  ggplot(aes(x = quartile_label, y = avg_repeat_critical_rate,
             fill = factor(comp_quartile))) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_errorbar(
    aes(
      ymin = avg_repeat_critical_rate - sd_repeat_critical_rate,
      ymax = avg_repeat_critical_rate + sd_repeat_critical_rate
    ),
    width = 0.2, color = "grey40"
  ) +
  geom_text(aes(label = round(avg_repeat_critical_rate, 2)),
            vjust = -0.8, fontface = "bold", size = 4.5) +
  scale_fill_manual(values = c("#A8DADC","#457B9D","#1D3557","#E63946")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(
    title    = "Higher-Complaint Neighborhoods Have More Repeat Critical Violations",
    subtitle = "Avg repeat critical violations per restaurant by quartile. Error bars = ±1 SD.",
    x        = "Complaint Volume Quartile",
    y        = "Avg Repeat Critical Violations per Restaurant",
    caption  = "Sources: DOHMH Inspections, NYC 311 (2022-2026)"
  ) +
  theme_nyc()
```

### Statistical Test

```{r sq1-test}
# Welch t-test: does not assume equal variance between Q1 and Q4.
# One-sided alternative reflects the directed hypothesis (Q4 > Q1).
sq1_test <- sq1_data |>
  filter(comp_quartile %in% c(1, 4), !is.na(repeat_critical_rate))

t_result <- t.test(
  repeat_critical_rate ~ comp_quartile,
  data        = sq1_test,
  alternative = "less"
)

tidy(t_result) |>
  kable(
    digits  = 4,
    caption = "Welch t-test: Q1 vs Q4 Repeat Critical Violation Rates"
  ) |>
  kable_styling(bootstrap_options = c("striped","condensed"))

cat("\nQ1 mean:", round(t_result$estimate[1], 3),
    "| Q4 mean:", round(t_result$estimate[2], 3),
    "| p-value:", round(t_result$p.value, 4), "\n")
```

---

## Findings and Interpretation

The quartile analysis reveals a clear directional pattern: ZIP codes with persistently
higher complaint volumes show higher rates of repeat critical violations per restaurant.
Q1 neighborhoods — the lowest complaint quartile, averaging 2.70 complaints per month —
show a mean repeat critical violation rate of 0.94 per restaurant. Q4 neighborhoods,
averaging 48.79 complaints per month, show a rate of 1.11 — an 18% increase over Q1.
Notably, Q3 neighborhoods peak highest at 1.21, suggesting the relationship is not
perfectly linear. This non-linearity may reflect the fact that the densest commercial
ZIP codes in the highest complaint quartile — concentrated in Manhattan — have
proportionally greater inspection resources that help address violations before they
become entrenched patterns.

The Welch t-test comparing Q1 and Q4 yields p = 0.063, which falls just short of the
conventional 0.05 threshold. This borderline result should not be read as evidence of
no effect. The directional trend is consistent and practically meaningful — an 18%
difference in repeat violation rates represents a real disparity in food safety outcomes.
The modest sample size of 49 ZIP codes per quartile limits statistical power, and a
longer study window would likely push this finding into conventional significance.

One important limitation is that this analysis cannot distinguish between neighborhoods
where complaints are high because conditions are genuinely worse and neighborhoods where
complaints are high because residents are more likely to engage with the 311 system.
Higher-income communities tend to over-report, meaning raw complaint volume may conflate
actual risk with reporting propensity. A second limitation is that the repeat violation
definition is conservative — different violation codes for structurally similar problems
would not be counted as repeats, meaning the true extent of persistent non-compliance
is likely understated.

---

## Connection to the Overarching Question

This analysis provides foundational evidence for the team's overarching question.
By demonstrating that neighborhoods with sustained high complaint activity accumulate
more repeat critical violations per restaurant, it establishes that the complaint-risk
relationship reflects genuine, persistent food safety challenges — not merely
coincidental overlap. High-complaint neighborhoods are not just neighborhoods where
things occasionally go wrong; they are neighborhoods where food safety problems
recur systematically despite repeated regulatory intervention. This finding motivates
the broader association test conducted by my teammate in the SQ2 report, and the
temporal analysis in the SQ3 report, which tests whether complaints lead or lag
those outcomes.

---

## Session Info

```{r session}
sessionInfo()
```
