---
title: "Temporal Dynamics of Complaints and Inspection Outcomes"
subtitle: "STA 9750 | Individual Report | Restaurant Complaints vs Inspection Risk in NYC"
author: "Group Member 3"
date: "2026-05-12"
format:
  html:
    toc: true
    toc-float: true
    toc-depth: 3
    theme: flatly
    highlight-style: tango
    code-fold: true
    fig-width: 10
    fig-height: 6
    self-contained: true
execute:
  echo: true
  message: false
  warning: false
  fig-align: center
---

## Overarching Question and My Contribution

This report is one of five individual analyses contributing to the team project
*Restaurant Complaints vs Inspection Risk in NYC*. The project's overarching question
asks whether neighborhoods with higher complaint activity experience systematically
worse restaurant inspection outcomes. The first two analyses in our project establish
that the answer to this question is yes in a cross-sectional sense: ZIP codes with
more food-related 311 complaints also show more repeat critical violations and higher
average inspection scores. My contribution takes the analysis one step further by
examining the temporal structure of this relationship.

Knowing that complaints and bad inspection outcomes co-occur is useful, but it does
not tell us whether complaints are leading or lagging indicators of food safety
deterioration. This distinction has critical practical implications. If complaint
spikes in month T reliably predict worse inspection outcomes in months T+1, T+2, and
T+3, then 311 complaints could function as a genuine early warning system: regulators
who monitored complaint trends could identify deteriorating neighborhoods and deploy
inspection resources proactively before conditions worsen further. If instead
complaints and violations are purely contemporaneous signals of the same underlying
conditions, then 311 data is a concurrent diagnostic of current risk rather than a
forward-looking predictor of future deterioration.

My specific question is whether increases in neighborhood-level food-related complaints
precede worse inspection outcomes. I test this by constructing a ZIP-month panel
dataset and examining whether complaint volume in month T predicts inspection outcomes
in months T+1, T+2, and T+3, using both lagged Pearson correlations and a panel
regression with ZIP-level fixed effects that controls for all time-invariant
neighborhood characteristics.

The answer I find is a carefully qualified no: complaint spikes do not reliably
predict future inspection deterioration within a one-to-three month window. This null
finding is not a failure but rather an important clarification of the mechanism
underlying the cross-sectional association documented by my teammates. It tells us
that complaints and violations are simultaneous manifestations of the same underlying
neighborhood food safety environment rather than complaints serving as a leading
indicator of upcoming violations. This distinction shapes how policymakers should
think about the appropriate role of 311 data in regulatory targeting.

---

## Setup

```{r packages}
pkgs <- c("tidyverse", "lubridate", "scales", "knitr",
          "kableExtra", "broom", "glue", "httr2", "jsonlite")
to_install <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(to_install) > 0) install.packages(to_install)

library(tidyverse)
library(lubridate)
library(scales)
library(knitr)
library(kableExtra)
library(broom)
library(glue)
library(httr2)
library(jsonlite)
```

```{r theme}
theme_nyc <- function() {
  theme_minimal(base_size = 13) +
    theme(plot.title = element_text(face = "bold", size = 15),
          plot.subtitle = element_text(color = "grey40", size = 12),
          plot.caption = element_text(color = "grey55", size = 9),
          axis.title = element_text(size = 11),
          panel.grid.minor = element_blank())
}
```

---

## Data Acquisition

```{r fetch-helper}
# Standard paginated Socrata fetcher shared across team reports.
# The temporal analysis requires both datasets at monthly granularity,
# so no additional aggregation parameters are needed at acquisition time.
fetch_all_socrata <- function(dataset_id, query = "",
                               select = NULL, page_size = 50000) {
  base_url <- glue("https://data.cityofnewyork.us/resource/{dataset_id}.json")
  offset <- 0; pages <- list()
  repeat {
    req <- request(base_url) |>
      req_url_query(`$limit` = page_size, `$offset` = offset, `$where` = query)
    if (!is.null(select)) req <- req |> req_url_query(`$select` = select)
    page <- req |> req_retry(max_tries = 3) |> req_perform() |>
      resp_body_json(simplifyVector = TRUE) |> as_tibble()
    if (nrow(page) == 0) break
    message(glue("  {dataset_id}: {offset + nrow(page)} rows"))
    pages[[length(pages) + 1]] <- page
    if (nrow(page) < page_size) break
    offset <- offset + page_size
  }
  bind_rows(pages)
}
```

```{r load-data}
#| cache: true
dir.create("data", showWarnings = FALSE)

if (!file.exists("data/raw_inspections.rds")) {
  raw_inspections <- fetch_all_socrata(
    "43nn-pn8j", query = "inspection_date >= '2022-01-01T00:00:00'",
    select = "camis,boro,zipcode,inspection_date,score,critical_flag"
  )
  saveRDS(raw_inspections, "data/raw_inspections.rds")
} else { raw_inspections <- readRDS("data/raw_inspections.rds") }

if (!file.exists("data/raw_311.rds")) {
  raw_311 <- fetch_all_socrata(
    "erm2-nwe9",
    query = paste("complaint_type IN ('Food Establishment','Food Poisoning',",
                  "'Rodent','Unsanitary Condition')",
                  "AND created_date >= '2022-01-01T00:00:00'"),
    select = "unique_key,complaint_type,created_date,borough,incident_zip"
  )
  saveRDS(raw_311, "data/raw_311.rds")
} else { raw_311 <- readRDS("data/raw_311.rds") }
```

---

## Data Cleaning

```{r clean}
inspections <- raw_inspections |>
  mutate(
    inspection_date = as_date(inspection_date),
    score           = as.numeric(score),
    is_critical     = str_detect(tolower(critical_flag), "critical") &
                      !str_detect(tolower(critical_flag), "not"),
    borough         = str_to_upper(str_trim(boro)),
    zipcode         = str_pad(str_extract(zipcode, "\\d{5}"), 5, "left", "0"),
    year            = year(inspection_date),
    month           = floor_date(inspection_date, "month")
  ) |>
  filter(!is.na(inspection_date), year >= 2022,
         borough %in% c("BRONX","BROOKLYN","MANHATTAN",
                        "QUEENS","STATEN ISLAND"),
         !is.na(zipcode), zipcode != "00000")

complaints <- raw_311 |>
  mutate(
    date    = as_date(created_date),
    year    = year(date),
    month   = floor_date(date, "month"),
    borough = str_to_upper(str_trim(borough)),
    zipcode = str_pad(str_extract(incident_zip, "\\d{5}"), 5, "left", "0")
  ) |>
  filter(!is.na(date), year >= 2022,
         borough %in% c("BRONX","BROOKLYN","MANHATTAN",
                        "QUEENS","STATEN ISLAND"),
         !is.na(zipcode), zipcode != "00000")

rm(raw_inspections, raw_311); gc()

# Monthly time series for both datasets
zip_insp_monthly <- inspections |>
  group_by(zipcode, month) |>
  summarise(pct_critical = mean(is_critical, na.rm = TRUE),
            mean_score   = mean(score,       na.rm = TRUE),
            .groups = "drop")

zip_comp_monthly <- complaints |>
  count(zipcode, month, name = "comp_total")

cat("ZIP-month inspection rows:", nrow(zip_insp_monthly), "\n")
cat("ZIP-month complaint rows:",  nrow(zip_comp_monthly),  "\n")
```

---

## Data Quality Assessment

The temporal analysis faces a data quality challenge that does not arise in the
cross-sectional analyses: the alignment of complaint and inspection timing. Both
complaints and inspections are recorded with specific dates, but inspections do not
occur on a regular calendar schedule. The DOHMH inspects restaurants on a risk-based
cycle that can result in gaps of anywhere from a few weeks to well over a year between
consecutive inspection visits. When both datasets are aggregated to the calendar month,
a ZIP code's inspection outcome in a given month reflects only those restaurants that
happened to be inspected that month, which may be a non-representative subset if
inspections are not uniformly distributed across the calendar year.

This temporal misalignment is a genuine source of noise in the lagged analysis. A
ZIP code might show high complaint volume in month T but have very few inspections
in months T+1 through T+3, making it impossible to detect a complaint-to-outcome
relationship at those lags even if one existed at the restaurant level. Conversely,
a ZIP code with high inspection activity in a given month might show high critical
violation rates simply because more inspections were conducted, not because conditions
were particularly bad that month. These sources of noise would tend to attenuate
lagged correlations toward zero, meaning the null finding reported here should be
interpreted as an upper bound on the true lagged relationship rather than as
definitive evidence of no temporal connection.

A second quality consideration is the length of the study window. The dataset covers
January 2022 through April 2026, providing approximately 52 months of panel data.
With a three-month lag specification, the effective analysis window is reduced to
49 months of outcome observations for each ZIP code. While this is sufficient for
the lagged correlation analysis, the panel regression with ZIP fixed effects removes
all between-ZIP variation and relies entirely on within-ZIP changes over time, which
further reduces the effective variation available for identification. Both of these
structural limitations reduce statistical power and would tend to produce null results
even if a genuine lagged relationship existed.

---

## Exploratory Data Analysis

```{r eda-city-trends}
# City-wide monthly trends in both outcomes simultaneously.
# Visual inspection of whether complaint and violation trends move together
# or show lead-lag patterns is informative before formal testing.
city_monthly <- zip_insp_monthly |>
  group_by(month) |>
  summarise(avg_pct_critical = mean(pct_critical, na.rm = TRUE),
            .groups = "drop") |>
  left_join(
    zip_comp_monthly |>
      group_by(month) |>
      summarise(total_comp = sum(comp_total), .groups = "drop"),
    by = "month"
  )

ggplot(city_monthly, aes(x = month)) +
  geom_col(aes(y = total_comp / 5000), fill = "#A8DADC", alpha = 0.7) +
  geom_line(aes(y = avg_pct_critical), color = "#E63946", linewidth = 1.2) +
  geom_point(aes(y = avg_pct_critical), color = "#E63946", size = 2) +
  scale_y_continuous(
    name     = "Avg % Critical Violations",
    sec.axis = sec_axis(~ . * 5000, name = "Total Complaints",
                        labels = comma)
  ) +
  labs(title = "City-Wide Monthly Trends: Complaints vs Critical Violations",
       subtitle = "Bars = total complaints (right axis). Line = avg % critical (left axis).",
       x = NULL, caption = "Sources: DOHMH Inspections, NYC 311 (2022-2026)") +
  theme_nyc()
```

The city-wide trend chart reveals an important visual pattern: complaint volumes
show clear seasonal peaks driven by summer rodent activity, but these seasonal spikes
do not appear to be followed by corresponding increases in critical violation rates
in subsequent months. The critical violation rate line remains relatively stable
across the study period, without obvious upward movements following the complaint
peaks visible in the bars. This visual pattern anticipates the null finding from the
formal lagged analysis: complaints and violations are moving according to different
rhythms, with complaints driven partly by seasonal factors (rodent activity in summer)
that may not be directly related to the kind of food handling and preparation
violations that inspectors cite as critical.

```{r eda-within-zip}
# Examine within-ZIP temporal variation in both complaints and outcomes
# to assess whether there is sufficient variation for the panel regression
zip_variation <- zip_comp_monthly |>
  group_by(zipcode) |>
  summarise(cv_complaints = sd(comp_total) / mean(comp_total),
            n_months = n(), .groups = "drop") |>
  filter(n_months >= 12)

cat("Median coefficient of variation in monthly complaints:",
    round(median(zip_variation$cv_complaints, na.rm = TRUE), 3), "\n")
cat("ZIP codes with at least 12 months of data:",
    nrow(zip_variation), "\n")
```

The coefficient of variation in monthly complaint volume across ZIP codes is
relatively high, indicating that complaint volumes do fluctuate substantially from
month to month within ZIP codes. This within-ZIP variation is necessary for the
fixed-effects panel regression to have statistical power: if complaint volumes were
completely stable within each ZIP over time, there would be no within-ZIP variation
to identify a temporal relationship. The substantial within-ZIP variation in complaint
volume confirms that the panel regression has the necessary data variation to detect
a lagged relationship if one exists.

---

## Analysis

### Building the Lagged Panel Dataset

```{r sq3-lag}
# The make_lag helper shifts complaint month forward by lag_months,
# effectively asking: given the complaint volume we observed in month T,
# what were the inspection outcomes in month T + lag_months?
# This is constructed by shifting the complaint date forward rather than
# shifting inspection outcomes backward, which is analytically equivalent
# but computationally cleaner with the left_join approach.
make_lag <- function(df, lag_months) {
  df |>
    mutate(month = month + months(lag_months)) |>
    rename_with(~ paste0(.x, "_lag", lag_months), comp_total)
}

sq3_data <- zip_insp_monthly |>
  left_join(make_lag(zip_comp_monthly, 1), by = c("zipcode","month")) |>
  left_join(make_lag(zip_comp_monthly, 2), by = c("zipcode","month")) |>
  left_join(make_lag(zip_comp_monthly, 3), by = c("zipcode","month")) |>
  filter(!is.na(pct_critical))

cat("Panel rows for lagged analysis:", nrow(sq3_data), "\n")
cat("Distinct ZIPs:", n_distinct(sq3_data$zipcode), "\n")
cat("Distinct months:", n_distinct(sq3_data$month), "\n")
```

### Lagged Correlations

```{r sq3-cors}
# Pearson correlations between lagged complaint volume and
# future inspection outcomes across all ZIP-month observations.
# These pooled correlations do not control for ZIP-level confounders
# and should be interpreted as descriptive rather than causal.
lag_cors <- tibble(
  lag = c("Lag 1 month","Lag 2 months","Lag 3 months"),
  cor_pct_critical = c(
    cor(sq3_data$comp_total_lag1, sq3_data$pct_critical,
        use = "pairwise.complete.obs"),
    cor(sq3_data$comp_total_lag2, sq3_data$pct_critical,
        use = "pairwise.complete.obs"),
    cor(sq3_data$comp_total_lag3, sq3_data$pct_critical,
        use = "pairwise.complete.obs")
  ),
  cor_mean_score = c(
    cor(sq3_data$comp_total_lag1, sq3_data$mean_score,
        use = "pairwise.complete.obs"),
    cor(sq3_data$comp_total_lag2, sq3_data$mean_score,
        use = "pairwise.complete.obs"),
    cor(sq3_data$comp_total_lag3, sq3_data$mean_score,
        use = "pairwise.complete.obs")
  )
)

lag_cors |>
  kable(digits = 4,
        caption = "Lagged Correlations: Complaints in T vs Outcomes in T+N",
        col.names = c("Lag Window","Cor w/ % Critical","Cor w/ Mean Score")) |>
  kable_styling(bootstrap_options = c("striped","condensed"))
```

```{r sq3-plot}
lag_cors |>
  pivot_longer(starts_with("cor_"), names_to = "outcome",
               values_to = "correlation") |>
  mutate(outcome = recode(outcome,
    "cor_pct_critical" = "% Critical Violations",
    "cor_mean_score"   = "Mean Inspection Score")) |>
  ggplot(aes(x = lag, y = correlation, fill = outcome)) +
  geom_col(position = "dodge", width = 0.5) +
  geom_hline(yintercept = 0, linewidth = 0.8) +
  scale_fill_manual(values = c("#E76F51","#264653")) +
  labs(title = "Do Complaint Spikes Predict Future Inspection Outcomes?",
       subtitle = "Lagged Pearson correlations: complaints in month T vs outcomes in T+N",
       x = "Complaint Lag Window", y = "Pearson Correlation", fill = "Outcome",
       caption = "Sources: DOHMH Inspections, NYC 311 (2022-2026)") +
  theme_nyc()
```

### Panel Regression with ZIP Fixed Effects

```{r sq3-regression}
# ZIP fixed effects absorb all time-invariant neighborhood characteristics,
# including long-run differences in restaurant density, neighborhood income,
# and inspection staffing. The remaining identification comes entirely from
# within-ZIP changes in complaint volume over time. This is a much stricter
# test than the pooled correlations above: it asks whether, within a given
# ZIP code, months with unusually high complaint volumes are followed by
# months with unusually high critical violation rates.
sq3_model <- lm(
  pct_critical ~ comp_total_lag1 + comp_total_lag2 +
                 comp_total_lag3 + factor(zipcode),
  data = sq3_data
)

# Display only the lag coefficients of interest
tidy(sq3_model) |>
  filter(str_detect(term, "comp_total_lag")) |>
  kable(digits = 5,
        caption = "Panel Regression (ZIP FE): Lagged Complaints to % Critical") |>
  kable_styling(bootstrap_options = c("striped","condensed"))
```

---

## Findings and Interpretation

The lagged correlation analysis produces small but consistently positive correlations
between prior complaint volume and subsequent inspection outcomes across all three
lag windows and both outcome measures. The correlations for percent critical
violations range from 0.028 to 0.031 across the three lags, while the correlations
for mean inspection score range from 0.072 to 0.076. These magnitudes are substantially
smaller than the contemporaneous cross-sectional correlation of 0.41 reported by the
team's second analyst, which confirms the central hypothesis of this analysis: most
of the complaint-outcome association operates in real time rather than through a
temporal leading relationship.

The panel regression with ZIP fixed effects provides the most rigorous test available
with the current data. After removing all time-invariant neighborhood characteristics
through ZIP fixed effects, none of the three lagged complaint predictors reaches
statistical significance. The lag-1 coefficient is 0.00010 with a p-value of 0.455,
the lag-2 coefficient is 0.00011 with a p-value of 0.482, and the lag-3 coefficient
is negative at -0.00018 with a p-value of 0.206. The negative lag-3 coefficient,
while not significant, is directionally inconsistent with a simple causal story in
which complaints predict future deterioration.

These results constitute an important and honest null finding. The null should not
be read as evidence that complaints are uninformative: my teammates' cross-sectional
analyses establish clearly that complaint volume and inspection outcomes are associated.
Rather, the null temporal finding reveals the mechanism underlying that association.
Both complaints and violations are simultaneous responses to persistent underlying
neighborhood food safety conditions. A ZIP code with chronic food safety problems
generates both 311 complaints and health code violations in the same months because
the underlying conditions that produce both are relatively stable. When those
underlying conditions fluctuate from month to month, complaint volume and violation
rates move together rather than complaint volume leading violation rates by a fixed
lag. The complaint system, in this view, is a concurrent mirror of food safety
conditions rather than a predictive telescope pointed at future deterioration.

There are three reasons to treat this null finding cautiously rather than as
definitive. First, the monthly aggregation of inspection data introduces noise because
inspections do not occur on a calendar-month schedule. A ZIP code's inspection
outcomes in a given month reflect a potentially non-representative subset of its
restaurants, making it difficult to detect smooth temporal relationships. Second, the
three-month lag window may be too narrow. The team's fifth analyst demonstrates that
complaint-triggered inspections arrive within days of the complaint, not months, which
means any regulatory response operating through inspection acceleration would be
captured in the same or adjacent month rather than at a three-month lag. Third,
panel identification relies on within-ZIP temporal variation, which may be limited
relative to the between-ZIP variation that drives the cross-sectional findings.

---

## Connection to the Overarching Question

This analysis contributes to the team's overarching question by clarifying the
temporal structure of the complaint-outcome relationship. The answer to whether
high-complaint neighborhoods have worse inspection outcomes is yes, as established
by the cross-sectional analyses. But the mechanism is not that complaints predict
future deterioration; rather, complaints and violations co-occur as simultaneous
signals of the same persistent food safety environment. This finding is important
for policy because it shapes how 311 data should be used in regulatory planning.
Rather than monitoring complaint trends for early warning signals of impending
deterioration, regulators should treat current complaint volume as a real-time
indicator of which neighborhoods are currently struggling with food safety compliance.
The question of whether complaints also accelerate inspection timing, which would
represent a direct regulatory response mechanism, is addressed in the team's fifth
individual report.

---

## Session Info

```{r session}
sessionInfo()
```
